### 2025-04-07 inst/paper/main.tex
initial commit

\documentclass{article}

\usepackage{graphicx}
\usepackage{geometry}
\geometry{
 a4paper,
 total={170mm,257mm},
 left=20mm,
 top=20mm,
}

\title{Evaluating contact tracing and isolation to control pandemic potential influenza outbreaks}
\author{Joshua W. Lambert}
\date{}

\begin{document}

\maketitle

\section{Abstract}

\section{Introduction}

The likelihood of an outbreak being controlled, in other words going locally-extinct or -eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical. The primary epidemiological determinant for the spread of an infectious disease is the reproduction number, defined as the average number of secondary cases  per primary case. The higher the reproduction number the less likely a disease, without intervention, is to cease transmission in a population. If a pathogen is subcritical (i.e. R $<$ 1) it will eventually stop spreading but can still cause transient outbreaks (Farrington and Grant, 1999). Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process (Lloyd-Smith et al., 2005), with higher individual-level variation for a given reproduction leading to more outbreaks being controlled (however other outbreaks may become larger due to superspreading). In addition, the likelihood of containment is reduced (assuming R $>$ 1) if there are multiple introductions into a populations (Kucharski et al., 2020). \\

There are a variety of interventions that can be enacted to try and prevent transmission of an infectious disease, including both pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak NPIs are often used, as especially for outbreaks of a novel pathogen before pharmaceutical inventions are available. These interventions range in their specificity, from targeted contact tracing which aims to tackle specific transmission pathways, to closure of school and gathering, to regional or national lockdowns affecting everyone (ref). \\

Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic. It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined is needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source. Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts (ref; Kucharski et al., 2016). Isolation of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics (SARS ref; other refs). However, it's important to remember that it's not a panacea for all outbreak scenarios. In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely (Dhillon and Srikrishna, 2018; ref).

Outbreaks have been shown to be more easily controlled when a large proportion of onward transmission takes place after the onset of symptoms (Fraser et al., 2004).

- Pandemic flu has similar pre-symptomatic transmission is similar to COVID-19 (ref).
- SI shorter for pandemic flu (ref).

It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation become the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (ref). However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data (Ward et al., 2024) it may not be the case that contact tracing and isolation is unable to control influenza. Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications (Tanner et al., 2015). \\

The novelty of a pathogen introduced into a human population, in the sense of antigenically novel, is a leading risk factor for causing a pandemic (ref). Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 and H7N9 (Tanner et al., 2015) and was the case when H1N1 caused an outbreak in 2009 (ref). \\

In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model. This can be used to define the speed and effectiveness of isolation given an epidemic scenario to inform outbreak response complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals (ref).

\section{Methods}

\subsection{Branching process model with isolation}

To assess the efficaciousness of contact tracing and isolation to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{ringbp} (Hellewell et al., 2025). The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility). The mean of the negative binomial offspring distribution is the basic reproduction number, and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals (Lloyd-Smith et al., 2005; Kucharski et al., 2020). \\

The NPI implemented in the transmission model is an isolation of symptomatic cases. The timing of isolation is drawn from the onset-to-isolation distribution (see section \ref{epiparameters} for parameterisation). All infection times that are sampled to occur after the isolation of an infector are not infected. Therefore, the isolation of a case can reduce the number of secondary cases if isolation happens prior to any of the secondary infection times. \\

\subsection{Defining outbreak control}

We use the same definition for outbreak control as Hellewell et al. (2020): no new cases between 12 and 16 weeks after the start of an outbreak. The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct. \\

There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable with 12-16 weeks. We set this limit to 500 cases.

\subsection{Pathogen epidemiological parameters} \label{epiparameters}

In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previous caused sporadic human-to-human outbreaks with pandemic potential. We include A/H5N1, A/H1N1, and A/H7N9. These have caused outbreaks of various size and severity over the past 100 years. H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission (Yang et al., 2007; Aditama et al., 2012). H1N1 (Fraser et al. 2009). \\

We extracted incubation periods from the epidemiological literature that have been previously estimated for the chosen Influenza subtypes. All incubation periods were found to be well described by a Weibull distribution, Nishiura and Inaba (2011) report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:incub}). Cowling et al. (2013) report the incubation period for H5N1 and H7N9, both Weibull distributed. For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days (Cowling et al. 2013, Figure \ref{fig:incub}). \\

The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5 and 2.0. We chose these values and they represent either similar values to previously estimated $R$ values for an outbreak influenza H1N1 (Fraser et al., 2009), or they are inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g. recombination of subtypes (ref)) enhances the human-to-human transmissibility of these pathogens. Current estimates for H5N1 and H7N9 subtypes are below unity (i.e. $R < 1$) (Tanner et al., 2015; Ward et al., 2024, but see Yang et al. 2007 for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event). Therefore evaluating contact tracing and isolation requires simulating outbreaks that are likely to exponentially spread without control measures.

\begin{itemize}

\item H1N1-like
\item H5N1 (slightly increased reproduction number, R~1)
\item H7N9 (if parameters are available)

\end{itemize}

Figure for parameter space for flu-like pandemic

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../plots/incubation_period.png}
  \caption{Incubation period for the Influenza subtypes: H1N1, H5N1 and H7N9.}
  \label{fig:incub}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../plots/onset_to_isolation.png}
  \caption{Onset-to-isolation delay distribution for SARS (SARS-CoV-1) and
  COVID-19 (SARS-CoV-2).}
  \label{fig:onset-to-isolation}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{../plots/prop_presymptomatic_transmission.png}
  \caption{Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skewed-normal distribution.}
  \label{fig:prop-presym_trans}
\end{figure}

We extract the incubation period or serial interval for each influenza subtype. We define the proportion of pre-symptomatic transmission as ... We define the onset-to-isolation delay distribution as short or long following the approach from Hellewell et al. (2020).

\subsection{Simulation scenarios}

\section{Results}

The incubation period for H1N1 is noticable shorter than the other two subtypes, which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:incub}).

\section{Discussion}

\begin{itemize}
\item This study has looked at the feasibility and effectiveness of contact tracing and isolation for influenza subtypes that have cause epidemic outbreaks.
\item Discuss the existing flu vaccines and antivirals that can be used an interventions in addition to NPIs
\item In modelling the effectiveness of contact tracing and isolation we have assumed that there is no capacity limits on contact tracing, which does not hold in large outbreaks such as SARS-CoV-2 (ref). We also ignore potential socio-political factors that would prevent health worker from visiting or contacting infected individuals or contacts of infected individuals (Dhillon and Kelly, 2018). We also did not consider zoonotic spillover infections or formite transmission.
\item Digital tracing has capacity for large-scale contact tracing (pingdemic), also rapid tests are a complementary approach for limiting spread.
\item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic.
\item In this study we have parameterised the transmission model with parameter from avian influenza subtypes (H5N1 and H7N9) and a influenza subtype that caused an epidemic in the recent past (H1N1), and response delays from SARS and COVID-19 outbreaks. However, these results can be used to inform other influenza subtypes that exhibit similar epidemiological characteristics in transmissibility and severity from past outbreaks (H2N2 and H3N2) and still exist in animal reservoirs for potential spillovers back into human populations (ref).
\end{itemize}

\section{References}

\begin{itemize}
\item Farrington and Grant (1999) The distribution of time to extinction in subcritical branching processes: applications to outbreaks of infectious disease.
\item Fraser et al. (2004) Factors that make an infectious disease outbreak controllable
\item Hellewell et al. (2020) Feasibility of controlling COVID-19 outbreaks by isolation of cases and contacts
\item Kucharski et al. (2016) Effectiveness of Ring Vaccination as Control Strategy for Ebola Virus Disease
\item Kucharski et al. (2020) Early dynamics of transmission and control of COVID-19: a mathematical modelling study
\item Nishiura and Inaba (2011) Estimation of the incubation period of influenza A (H1N1-2009) among imported cases: Addressing censoring using outbreak data at the origin of importation
\item Yang et al. (2007) Detecting Human-to-Human Transmission of Avian Influenza A (H5N1)
\item Ward et al. (2024) Estimates of epidemiological parameters for H5N1 influenza in humans: a rapid review
\end{itemize}

\end{document}

### 2025-04-07 R/foo.R

@@ -1,0 +1,12 @@
+#' Placeholder function
+#'
+#' @param x \R object.
+#'
+#' @return Nothing, prints `x`.
+#' @export
+#'
+#' @examples
+#' foo(5)
+foo <- function(x) {
+  print(x)
+}

### 2025-04-07 inst/scripts/plot_incubation_period.R

@@ -1,0 +1,89 @@
+library(epiparameter)
+library(data.table)
+library(ggplot2)
+
+h5n1_weibull_params <- epiparameter::convert_summary_stats_to_params(
+  "weibull", mean = 3.3, sd = 1.5
+)
+
+h7n9_weibull_params <- epiparameter::convert_summary_stats_to_params(
+  "weibull", mean = 3.1, sd = 1.4
+)
+
+density_eval <- seq(0, 10, 0.1)
+
+incubation_period <- data.table(
+  x = density_eval,
+  H1N1 = dweibull(x = density_eval, shape = 1.77, scale = 1.86),
+  H5N1 = dweibull(
+    x = density_eval,
+    shape = h5n1_weibull_params$shape,
+    scale = h5n1_weibull_params$scale
+  ),
+  H7N9 = dweibull(
+    x = density_eval,
+    shape = h7n9_weibull_params$shape,
+    scale = h7n9_weibull_params$scale
+  )
+)
+
+incubation_period <- melt(
+  data = incubation_period,
+  id.vars = "x",
+  measure.vars = c("H1N1", "H5N1", "H7N9"),
+  variable.name = "pathogen",
+  value.name = "incubation_period"
+)
+
+h1n1_median <- epiparameter::convert_params_to_summary_stats(
+  "weibull", shape = 1.77, scale = 1.86
+)$median
+h5n1_median <- epiparameter::convert_params_to_summary_stats(
+  "weibull",
+  shape = h5n1_weibull_params$shape,
+  scale = h5n1_weibull_params$scale
+)$median
+h7n9_median <- epiparameter::convert_params_to_summary_stats(
+  "weibull",
+  shape = h7n9_weibull_params$shape,
+  scale = h7n9_weibull_params$scale
+)$median
+
+incubation_period_plot <- ggplot2::ggplot(data = incubation_period) +
+  ggplot2::geom_line(
+    mapping = ggplot2::aes(x = x, y = incubation_period, col = pathogen)
+  ) +
+  ggplot2::geom_ribbon(
+    mapping = ggplot2::aes(
+      x = x,
+      y = incubation_period,
+      ymax = incubation_period,
+      ymin = 0,
+      fill = pathogen),
+    alpha = 0.1
+  ) +
+  ggplot2::geom_vline(
+    mapping = ggplot2::aes(xintercept = h1n1_median), lty = 2, col = "#FFA500"
+  ) +
[…]

### 2025-04-07 inst/scripts/plot_onset_to_isolation.R

@@ -1,0 +1,67 @@
+library(data.table)
+library(ggplot2)
+
+density_eval <- seq(0, 25, 0.1)
+
+onset_to_isolation <- data.table(
+  x = density_eval,
+  SARS = dweibull(density_eval, shape = 1.651524, scale = 4.287786),
+  COVID = dweibull(density_eval, shape = 2.305172, scale = 9.483875)
+)
+
+onset_to_isolation <- melt(
+  data = onset_to_isolation,
+  id.vars = "x",
+  measure.vars = c("SARS", "COVID"),
+  variable.name = "disease",
+  value.name = "onset_to_isolation"
+)
+
+sars_median <- epiparameter::convert_params_to_summary_stats(
+  "weibull", shape = 1.651524, scale = 4.287786
+)$median
+covid_median <- epiparameter::convert_params_to_summary_stats(
+  "weibull", shape = 2.305172, scale = 9.483875
+)$median
+
+
+onset_to_isolation_plot <- ggplot2::ggplot(data = onset_to_isolation) +
+  ggplot2::geom_line(
+    mapping = ggplot2::aes(
+      x = x,
+      y = onset_to_isolation,
+      col = disease
+    )
+  ) +
+  ggplot2::geom_ribbon(
+    mapping = ggplot2::aes(
+      x = x,
+      y = onset_to_isolation,
+      ymin = 0,
+      ymax = onset_to_isolation,
+      fill = disease
+    ),
+    alpha = 0.1
+  ) +
+  ggplot2::geom_vline(
+    mapping = ggplot2::aes(xintercept = sars_median), lty = 2, col = "#1A85FF"
+  ) +
+  ggplot2::geom_vline(
+    mapping = ggplot2::aes(xintercept = covid_median), lty = 2, col = "#D41159"
+  ) +
+  ggplot2::scale_x_continuous(name = "Days from symptom onset to isolation") +
+  ggplot2::scale_y_continuous(name = "Density") +
+  ggplot2::scale_colour_manual(values = c("#1A85FF", "#D41159")) +
+  ggplot2::scale_fill_manual(values = c("#1A85FF", "#D41159")) +
+  ggplot2::labs(colour = "Disease", fill = "Disease") +
+  ggplot2::theme_bw()
+
+ggplot2::ggsave(
+  file.path("inst", "plots", "onset_to_isolation.png"),
+  plot = onset_to_isolation_plot,
+  device = "png",
+  width = 160,
+  height = 80,
+  units = "mm",
+  dpi = 300
+)

### 2025-04-07 inst/scripts/plot_prop_presymptomatic_transmission.R

@@ -1,0 +1,70 @@
+library(data.table)
+library(ggplot2)
+library(sn)
+
+density_eval <- seq(0, 15, 0.1)
+
+prop_presymptomatic_transmission <- data.table(
+  x = density_eval,
+  prop_presym_trans_30 = sn::dsn(
+    x = density_eval, xi = 5, omega = 2, alpha = 0.7
+  ),
+  prop_presym_trans_15 = sn::dsn(
+    x = density_eval, xi = 5, omega = 2, alpha = 1.95
+  ),
+  prop_presym_trans_lt1 = sn::dsn(
+    x = density_eval, xi = 5, omega = 2, alpha = 30
+  )
+)
+
+prop_presymptomatic_transmission <- melt(
+  data = prop_presymptomatic_transmission,
+  id.vars = "x",
+  measure.vars = c("prop_presym_trans_30", "prop_presym_trans_15", "prop_presym_trans_lt1"),
+  variable.name = "prop",
+  value.name = "prop_presym_trans"
+)
+
+incubation_period <- 5
+
+prop_presymptomatic_transmission_plot <- ggplot2::ggplot(
+  data = prop_presymptomatic_transmission
+) +
+  ggplot2::geom_line(
+    mapping = ggplot2::aes(
+      x = x,
+      y = prop_presym_trans,
+      col = prop
+    )
+  ) +
+  ggplot2::geom_ribbon(
+    mapping = ggplot2::aes(
+      x = x,
+      y = prop_presym_trans,
+      ymin = 0,
+      ymax = prop_presym_trans,
+      fill = prop
+    ),
+    alpha = 0.1
+  ) +
+  ggplot2::geom_vline(
+    mapping = ggplot2::aes(xintercept = incubation_period),
+    lty = 2
+  ) +
+  ggplot2::scale_x_continuous(name = "Days since infection") +
+  ggplot2::scale_y_continuous(name = "Density") +
+  ggplot2::labs(colour = "Proportion presymptomatic transmission", fill = "Proportion presymptomatic transmission") +
+  ggplot2::scale_colour_manual(values = c("#009e74", "#56b3e9", "#cc79a7"), labels = c("30%", "15%", "<1%")) +
+  ggplot2::scale_fill_manual(values = c("#009e74", "#56b3e9", "#cc79a7"), labels = c("30%", "15%", "<1%")) +
+  ggplot2::theme_bw() +
+  ggplot2::theme(legend.position = "bottom")
+
[…]

### 2025-04-07 inst/scripts/prop_outbreak_control_num_init_cases_plot.R

@@ -1,0 +1,37 @@
+library(ggplot2)
+library(data.table)
+
+# read simulation results saved by running inst/scripts/run_analysis.R
+
+sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
+
+res <- sweep_results %>%
+  group_by(scenario) %>%
+  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
+  ungroup(scenario)
+
+dt <- as.data.table(res)
+dt_data <- rbindlist(dt$data)
+dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
+
+prop_outbreak_control_num_init_cases <- dt_data %>%
+  dplyr::filter(delay == "SARS",
+                theta == "15%",
+                index_R0 == 2.5,
+                prop.asym == 0) %>%
+  ggplot2::ggplot(ggplot2::aes(x = control_effectiveness,
+                               y = pext,
+                               color = as.factor(num.initial.cases))) +
+  ggplot2::geom_line(size = 0.75) +
+  ggplot2::geom_point(shape = 21,
+                      col = "black",
+                      aes(fill = as.factor(num.initial.cases)),
+                      size = 3) +
+  ggplot2::scale_fill_manual(guide = "none",
+                             values = c("dodgerblue", "black", "dodgerblue3")) +
+  ggplot2::scale_color_manual(values = c("dodgerblue", "black", "dodgerblue3"),
+                              name = "Number of\ninitial cases")  +
+  cowplot::theme_cowplot()
+
+ggplot2::ggsave(file.path("inst", "plots", "prop_outbreak_control_num_init_cases.png"), plot = prop_outbreak_control_num_init_cases)
+

### 2025-04-07 inst/scripts/prop_outbreak_control_onset_to_isolation_plot.R

@@ -1,0 +1,42 @@
+library(ggplot2)
+library(data.table)
+
+# read simulation results saved by running inst/scripts/run_analysis.R
+
+sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
+
+res <- sweep_results %>%
+  group_by(scenario) %>%
+  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
+  ungroup(scenario)
+
+dt <- as.data.table(res)
+dt_data <- rbindlist(dt$data)
+dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
+
+prop_outbreak_control_onset_to_isolation <- dt_data %>%
+  dplyr::filter(num.initial.cases == 20,
+                theta == "15%",
+                index_R0 == 2.5,
+                prop.asym == 0) %>%
+  dplyr::mutate(delay = factor(delay,
+                               levels = c("SARS", "Wuhan"),
+                               labels = c("Short", "Long"))) %>%
+  ggplot2::ggplot(ggplot2::aes(x = control_effectiveness,
+                               y = pext,
+                               color = as.factor(delay))) +
+  ggplot2::geom_line(size = 0.75) +
+  ggplot2::geom_point(shape = 21,
+                      col = "black",
+                      aes(fill = as.factor(delay)),
+                      size = 3) +
+  ggplot2::scale_fill_manual(guide = "none",
+                             values = c("black", "forestgreen")) +
+  ggplot2::scale_color_manual(values = c("black", "forestgreen"),
+                              name = "Onset to\nisolation\ndelay") +
+  cowplot::theme_cowplot()
+
+ggplot2::ggsave(
+  file.path("inst", "plots", "prop_outbreak_control_onset_to_isolation.png"),
+  plot = prop_outbreak_control_onset_to_isolation
+)

### 2025-04-07 inst/scripts/prop_outbreak_control_plot.R

@@ -1,0 +1,45 @@
+# read simulation results saved by running inst/scripts/run_analysis.R
+
+sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
+
+res <- sweep_results %>%
+  group_by(scenario) %>%
+  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
+  ungroup(scenario)
+
+dt <- as.data.table(res)
+
+dt_data <- rbindlist(dt$data)
+
+dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
+
+prop_outbreaks_control <- dt_data %>%
+  dplyr::filter(num.initial.cases == 20,
+                theta == "15%",
+                delay == "SARS",
+                prop.asym == 0) %>%
+  dplyr::select(control_effectiveness, index_R0, pext) %>%
+  ggplot2::ggplot(ggplot2::aes(x = control_effectiveness,
+                               y = pext,
+                               color = as.factor(index_R0))) +
+  ggplot2::geom_line(size = 0.75) +
+  ggplot2::geom_point(shape = 21,
+                      col = "black",
+                      ggplot2::aes(fill = as.factor(index_R0)),
+                      size = 3) +
+  ggplot2::scale_fill_manual(guide = "none",
+                             values = c("red", "black", "firebrick4")) +
+  ggplot2::scale_color_manual(values = c("red", "black", "firebrick4"),
+                              name = "Reproduction\nnumber")  +
+  cowplot::theme_cowplot() +
+  ggplot2::scale_x_continuous(breaks = seq(0, 1, 0.2),
+                              labels = seq(0, 100, 20)) +
+  ggplot2::scale_y_continuous(breaks = seq(0, 1, 0.2),
+                              labels = seq(0, 100, 20)) +
+  ggplot2::theme(legend.position = "bottom") +
+  ggplot2::labs(tag = "A",
+                x = "Contacts traced (%)",
+                y = "Simulated outbreaks controlled (%)")
+
+ggplot2::ggsave(file.path("inst", "plots", "prop_outbreaks_control.png"), plot = prop_outbreaks_control)
+

### 2025-04-07 inst/scripts/run_analysis.R

@@ -1,0 +1,49 @@
+library(ringbp)
+library(dplyr)
+library(tidyr)
+library(purrr)
+library(future)
+library(tibble)
+library(data.table)
+
+scenarios <- tidyr::expand_grid(
+  ## Put parameters that are grouped by disease into this data.frame
+  delay_group = list(tibble::tibble(
+    delay = c("SARS", "Wuhan"),
+    delay_shape = c(1.651524, 2.305172),
+    delay_scale = c(4.287786, 9.483875)
+  )),
+  k_group = list(tibble::tibble(
+    theta = c("<1%", "15%"),
+    k = c(30, 1.95)
+  )),
+  index_R0 = c(1.5, 2.5),
+  quarantine = FALSE,
+  prop.asym = c(0, 0.1),
+  control_effectiveness = seq(0, 1, 0.2),
+  num.initial.cases = c(5, 20)) %>%
+  tidyr::unnest("k_group") %>%
+  tidyr::unnest("delay_group") %>%
+  dplyr::mutate(scenario = 1:dplyr::n())
+
+## Parameterise fixed paramters
+sim_with_params <- purrr::partial(ringbp::scenario_sim,
+                                  cap_max_days = 365,
+                                  cap_cases = 500,
+                                  r0isolated = 0,
+                                  disp.iso = 1,
+                                  disp.com = 0.16)
+
+
+## Set up multicore if using see ?future::plan for details
+## Use the workers argument to control the number of cores used.
+future::plan("multisession")
+
+scenarios <- as.data.table(scenarios)
+
+## Run paramter sweep
+sweep_results <- ringbp::parameter_sweep(scenarios,
+                                         sim_fn = sim_with_params,
+                                         samples = 2)
+
+saveRDS(sweep_results, file = file.path("inst", "extdata", "simulations.rds"))

### 2025-04-07 inst/scripts/plot_prop_outbreak_control_num_init_cases.R
update analysis and plotting scripts

@@ -1,0 +1,71 @@
+library(ggplot2)
+library(data.table)
+
+# read simulation results saved by running inst/scripts/run_analysis.R
+sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
+
+res <- sweep_results %>%
+  group_by(scenario) %>%
+  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
+  ungroup(scenario)
+
+dt <- as.data.table(res)
+dt_data <- rbindlist(dt$data)
+dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
+
+
+prop_outbreak_control_num_init_cases <- dt_data[
+  index_R0 == 1.5 & theta == "15%" & delay == "SARS" & prop.asym == 0,
+  .(control_effectiveness, num.initial.cases, pext)
+]
+
+# convert to percentages for plotting
+prop_outbreak_control_num_init_cases[, control_effectiveness := control_effectiveness * 100]
+prop_outbreak_control_num_init_cases[, pext := pext * 100]
+
+prop_outbreak_control_num_init_cases_plot <- ggplot2::ggplot(
+  data = prop_outbreak_control_num_init_cases
+) +
+  ggplot2::geom_line(
+    mapping = ggplot2::aes(
+      x = control_effectiveness,
+      y = pext,
+      colour = as.factor(num.initial.cases)
+    ),
+    size = 0.75
+  ) +
+  ggplot2::geom_point(
+    mapping = ggplot2::aes(
+      x = control_effectiveness,
+      y = pext,
+      fill = as.factor(num.initial.cases)
+    ),
+    shape = 21,
+    size = 3,
+    colour = "black"
+  ) +
+  ggplot2::scale_x_continuous(
+    name = "Contacts traced (%)",
+    limits = c(0, 100)
+  ) +
+  ggplot2::scale_y_continuous(
+    name = "Simulated outbreaks controlled (%)",
+    limits = c(0, 100)
+  ) +
+  ggplot2::scale_colour_manual(values = c("#ffbe6a", "#40b0a7")) +
+  ggplot2::scale_fill_manual(values = c("#ffbe6a", "#40b0a7")) +
+  ggplot2::labs(
+    colour = "Number of initial cases",
+    fill = "Number of initial cases"
+  ) +
+  ggplot2::theme_bw()
+
+ggplot2::ggsave(
[…]

### 2025-04-07 inst/scripts/plot_prop_outbreak_control_reproduction_number.R

@@ -1,0 +1,70 @@
+library(data.table)
+library(ggplot2)
+
+# read simulation results saved by running inst/scripts/run_analysis.R
+sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
+
+res <- sweep_results %>%
+  group_by(scenario) %>%
+  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
+  ungroup(scenario)
+
+dt <- as.data.table(res)
+dt_data <- rbindlist(dt$data)
+dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
+
+prop_outbreak_control <- dt_data[
+  num.initial.cases == 10 & theta == "15%" & delay == "SARS" & prop.asym == 0,
+  .(control_effectiveness, index_R0, pext)
+]
+
+# convert to percentages for plotting
+prop_outbreak_control[, control_effectiveness := control_effectiveness * 100]
+prop_outbreak_control[, pext := pext * 100]
+
+prop_outbreak_control_plot <- ggplot2::ggplot(data = prop_outbreak_control) +
+  ggplot2::geom_line(
+    mapping = ggplot2::aes(
+      x = control_effectiveness,
+      y = pext,
+      colour = as.factor(index_R0)
+    ),
+    size = 0.75
+  ) +
+  ggplot2::geom_point(
+    mapping = ggplot2::aes(
+      x = control_effectiveness,
+      y = pext,
+      fill = as.factor(index_R0)
+    ),
+    shape = 21,
+    size = 3,
+    colour = "black"
+  ) +
+  ggplot2::scale_x_continuous(
+    name = "Contacts traced (%)",
+    limits = c(0, 100)
+  ) +
+  ggplot2::scale_y_continuous(
+    name = "Simulated outbreaks controlled (%)",
+    limits = c(0, 100)
+  ) +
+  ggplot2::scale_colour_manual(values = c("#f4b301", "#db1048")) +
+  ggplot2::scale_fill_manual(values = c("#f4b301", "#db1048")) +
+  ggplot2::labs(
+    colour = "Reproduction Number (R)",
+    fill = "Reproduction Number (R)"
+  ) +
+  ggplot2::theme_bw() +
+  ggplot2::theme(legend.position = "bottom")
+
+ggplot2::ggsave(
+  file.path("inst", "plots", "prop_outbreak_control_reproduction_number.png"),
[…]

### 2025-04-07 inst/scripts/prop_outbreak_control_num_init_cases_plot.R

@@ -1,37 +1,0 @@
-library(ggplot2)
-library(data.table)
-
-# read simulation results saved by running inst/scripts/run_analysis.R
-
-sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
-
-res <- sweep_results %>%
-  group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
-  ungroup(scenario)
-
-dt <- as.data.table(res)
-dt_data <- rbindlist(dt$data)
-dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
-
-prop_outbreak_control_num_init_cases <- dt_data %>%
-  dplyr::filter(delay == "SARS",
-                theta == "15%",
-                index_R0 == 2.5,
-                prop.asym == 0) %>%
-  ggplot2::ggplot(ggplot2::aes(x = control_effectiveness,
-                               y = pext,
-                               color = as.factor(num.initial.cases))) +
-  ggplot2::geom_line(size = 0.75) +
-  ggplot2::geom_point(shape = 21,
-                      col = "black",
-                      aes(fill = as.factor(num.initial.cases)),
-                      size = 3) +
-  ggplot2::scale_fill_manual(guide = "none",
-                             values = c("dodgerblue", "black", "dodgerblue3")) +
-  ggplot2::scale_color_manual(values = c("dodgerblue", "black", "dodgerblue3"),
-                              name = "Number of\ninitial cases")  +
-  cowplot::theme_cowplot()
-
-ggplot2::ggsave(file.path("inst", "plots", "prop_outbreak_control_num_init_cases.png"), plot = prop_outbreak_control_num_init_cases)
-

### 2025-04-07 inst/scripts/prop_outbreak_control_plot.R

@@ -1,45 +1,0 @@
-# read simulation results saved by running inst/scripts/run_analysis.R
-
-sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
-
-res <- sweep_results %>%
-  group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
-  ungroup(scenario)
-
-dt <- as.data.table(res)
-
-dt_data <- rbindlist(dt$data)
-
-dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
-
-prop_outbreaks_control <- dt_data %>%
-  dplyr::filter(num.initial.cases == 20,
-                theta == "15%",
-                delay == "SARS",
-                prop.asym == 0) %>%
-  dplyr::select(control_effectiveness, index_R0, pext) %>%
-  ggplot2::ggplot(ggplot2::aes(x = control_effectiveness,
-                               y = pext,
-                               color = as.factor(index_R0))) +
-  ggplot2::geom_line(size = 0.75) +
-  ggplot2::geom_point(shape = 21,
-                      col = "black",
-                      ggplot2::aes(fill = as.factor(index_R0)),
-                      size = 3) +
-  ggplot2::scale_fill_manual(guide = "none",
-                             values = c("red", "black", "firebrick4")) +
-  ggplot2::scale_color_manual(values = c("red", "black", "firebrick4"),
-                              name = "Reproduction\nnumber")  +
-  cowplot::theme_cowplot() +
-  ggplot2::scale_x_continuous(breaks = seq(0, 1, 0.2),
-                              labels = seq(0, 100, 20)) +
-  ggplot2::scale_y_continuous(breaks = seq(0, 1, 0.2),
-                              labels = seq(0, 100, 20)) +
-  ggplot2::theme(legend.position = "bottom") +
-  ggplot2::labs(tag = "A",
-                x = "Contacts traced (%)",
-                y = "Simulated outbreaks controlled (%)")
-
-ggplot2::ggsave(file.path("inst", "plots", "prop_outbreaks_control.png"), plot = prop_outbreaks_control)
-

### 2025-04-07 inst/scripts/run_analysis.R

@@ -17,11 +17,11 @@ scenarios <- tidyr::expand_grid(
     theta = c("<1%", "15%"),
     k = c(30, 1.95)
   )),
-  index_R0 = c(1.5, 2.5),
+  index_R0 = c(1.1, 1.5),
   quarantine = FALSE,
   prop.asym = c(0, 0.1),
   control_effectiveness = seq(0, 1, 0.2),
-  num.initial.cases = c(5, 20)) %>%
+  num.initial.cases = c(5, 10)) %>%
   tidyr::unnest("k_group") %>%
   tidyr::unnest("delay_group") %>%
   dplyr::mutate(scenario = 1:dplyr::n())
@@ -42,8 +42,13 @@ future::plan("multisession")
 scenarios <- as.data.table(scenarios)
 
 ## Run paramter sweep
-sweep_results <- ringbp::parameter_sweep(scenarios,
-                                         sim_fn = sim_with_params,
-                                         samples = 2)
+run_time <- system.time(
+  sweep_results <- ringbp::parameter_sweep(
+    scenarios,
+    sim_fn = sim_with_params,
+    samples = 100
+  )
+)
+
 
 saveRDS(sweep_results, file = file.path("inst", "extdata", "simulations.rds"))

### 2025-04-07 inst/paper/main.tex
update paper

@@
 If a pathogen is subcritical (i.e.
-R $<$ 1) it will eventually stop spreading but can still cause transient outbreaks (Farrington and Grant, 1999).
-Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process (Lloyd-Smith et al., 2005), with higher individual-level variation for a given reproduction leading to more outbreaks being controlled (however other outbreaks may become larger due to superspreading).
-In addition, the likelihood of containment is reduced (assuming R $>$ 1) if there are multiple introductions into a populations (Kucharski et al., 2020).
+$R < 1$) it will eventually stop spreading but can still cause transient outbreaks (Farrington and Grant, 1999).
+Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process (Lloyd-Smith et al., 2005).
+Higher individual-level variation for a given reproduction leading to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading).
+In addition, the likelihood of containment is reduced (assuming $R > 1$) if there are multiple introductions into a populations (Kucharski et al., 2020).
 \\
@@
 In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely (Dhillon and Srikrishna, 2018; ref).
+\\
 
@@
 - Pandemic flu has similar pre-symptomatic transmission is similar to COVID-19 (ref).
-- SI shorter for pandemic flu (ref).
+- SI shorter for pandemic flu (ref). \\
 
@@
 
-The novelty of a pathogen introduced into a human population, in the sense of antigenically novel, is a leading risk factor for causing a pandemic (ref).
+The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic (ref).
 Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates.
@@
 H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission (Yang et al., 2007; Aditama et al., 2012).
-H1N1 (Fraser et al.
-2009).
+H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak (Fraser et al.
+2009; Lessler et al., 2009; ref on history of H1N1pdm).
 \\
@@
   \caption{Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skewed-normal distribution.}
-  \label{fig:prop-presym_trans}
+  \label{fig:prop-presym-trans}
 \end{figure}
@@
 
+We simulated scenarios with either 5 or 10 initial cases.
+
 \section{Results}
@@
 The incubation period for H1N1 is noticable shorter than the other two subtypes, which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:incub}).
+
+Predictably, a higher reproduction number for the outbreak results in a smaller percentage of outbreaks controlled at a given percentage of control effectiveness.
+(Figure \ref{fig:prop-outbreak-control-R}).
+The disparity of percentage outbreaks controlled becomes zero as the percentage of contacts traced exceeds 80\% (Figure \ref{fig:prop-outbreak-control-R}).
+When $R$ is only slightly above unity but below 2, the percentage of outbreaks controlled at 50\% of contacts traced and isolated is over 90\%, highlighting the effectiveness of this intervention for infectious diseases are not highly transmissible between humans.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/prop_outbreak_control_reproduction_number.png}
+  \caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an assumed reproduction number ($R$) of either 1.1 or 1.5.
+This is for a baselines scenario where the number of initial cases is 10, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
+  \label{fig:prop-outbreak-control-R}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\textwidth]{../plots/prop_outbreak_control_num_init_cases.png}
+  \caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an initial number of seeding infections assumed to be either 5 or 10.
+This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
+  \label{fig:prop-outbreak-control-num-init-cases}
+\end{figure}
 

### 2025-04-09 inst/paper/main.tex
update paper to use Bibtex for references and add .bib file

@@
 }
+\usepackage{crimson}
+\usepackage[T1]{fontenc}
+\usepackage[round]{natbib}
+\usepackage{nameref}
 
@@
 
-\section{Abstract}
+\section*{Abstract}
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 If a pathogen is subcritical (i.e.
-$R < 1$) it will eventually stop spreading but can still cause transient outbreaks (Farrington and Grant, 1999).
-Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process (Lloyd-Smith et al., 2005).
+$R < 1$) it will eventually stop spreading but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}.
+Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}.
 Higher individual-level variation for a given reproduction leading to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading).
-In addition, the likelihood of containment is reduced (assuming $R > 1$) if there are multiple introductions into a populations (Kucharski et al., 2020).
+In addition, the likelihood of containment is reduced (assuming $R > 1$) if there are multiple introductions into a populations \citep{kucharskiEarlyDynamicsTransmission2020}.
 \\
@@
 It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined is needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
-Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts (ref; Kucharski et al., 2016).
+Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016} (ref).
 Isolation of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics (SARS ref; other refs).
 However, it's important to remember that it's not a panacea for all outbreak scenarios.
-In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely (Dhillon and Srikrishna, 2018; ref).
+In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018} (ref on limited capacity of tracing).
 \\
 
-Outbreaks have been shown to be more easily controlled when a large proportion of onward transmission takes place after the onset of symptoms (Fraser et al., 2004).
+Outbreaks have been shown to be more easily controlled when a large proportion of onward transmission takes place after the onset of symptoms \citep{fraserFactorsThatMake2004}.
 
@@
 It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation become the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (ref).
-However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data (Ward et al., 2024) it may not be the case that contact tracing and isolation is unable to control influenza.
-Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications (Tanner et al., 2015).
+However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may not be the case that contact tracing and isolation is unable to control influenza.
+Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}.
 \\
@@
 Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates.
-This is the case for H5N1 and H7N9 (Tanner et al., 2015) and was the case when H1N1 caused an outbreak in 2009 (ref).
+This is the case for H5N1 and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 (ref).
 \\
@@
 
-\section{Methods}
+\section*{Methods}
 
-\subsection{Branching process model with isolation}
+\subsection*{Branching process model with isolation}
 
-To assess the efficaciousness of contact tracing and isolation to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{ringbp} (Hellewell et al., 2025).
+To assess the efficaciousness of contact tracing and isolation to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{{ringbp}} \citep{hellewellRingbpSimulateEvaluate2025}.
 The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility).
-The mean of the negative binomial offspring distribution is the basic reproduction number, and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals (Lloyd-Smith et al., 2005; Kucharski et al., 2020).
+The mean of the negative binomial offspring distribution is the basic reproduction number, and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}.
 \\
@@
 The NPI implemented in the transmission model is an isolation of symptomatic cases.
-The timing of isolation is drawn from the onset-to-isolation distribution (see section \ref{epiparameters} for parameterisation).
+The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation).
 All infection times that are sampled to occur after the isolation of an infector are not infected.
@@
 
-\subsection{Defining outbreak control}
+\subsection*{Defining outbreak control}
 
-We use the same definition for outbreak control as Hellewell et al.
-(2020): no new cases between 12 and 16 weeks after the start of an outbreak.
+We use the same definition for outbreak control as \cite{hellewellFeasibilityControllingCOVID192020}: no new cases between 12 and 16 weeks after the start of an outbreak.
 The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct.
@@
 
-\subsection{Pathogen epidemiological parameters} \label{epiparameters}
+\subsection*{Pathogen epidemiological parameters} \label{epiparameters}
 
@@
 These have caused outbreaks of various size and severity over the past 100 years.
-H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission (Yang et al., 2007; Aditama et al., 2012).
-H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak (Fraser et al.
-2009; Lessler et al., 2009; ref on history of H1N1pdm).
+H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}.
+H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009} (ref on history of H1N1pdm).
 \\
@@
 We extracted incubation periods from the epidemiological literature that have been previously estimated for the chosen Influenza subtypes.
-All incubation periods were found to be well described by a Weibull distribution, Nishiura and Inaba (2011) report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:incub}).
-Cowling et al.
-(2013) report the incubation period for H5N1 and H7N9, both Weibull distributed.
-For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days (Cowling et al.
-2013, Figure \ref{fig:incub}).
+All incubation periods were found to be well described by a Weibull distribution, \cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:incub}).
+\cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed.
+For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:incub}).
 \\
@@
 The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5 and 2.0.
-We chose these values and they represent either similar values to previously estimated $R$ values for an outbreak influenza H1N1 (Fraser et al., 2009), or they are inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g.
+We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g.
 recombination of subtypes (ref)) enhances the human-to-human transmissibility of these pathogens.
 Current estimates for H5N1 and H7N9 subtypes are below unity (i.e.
-$R < 1$) (Tanner et al., 2015; Ward et al., 2024, but see Yang et al.
-2007 for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event).
+$R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event).
 Therefore evaluating contact tracing and isolation requires simulating outbreaks that are likely to exponentially spread without control measures.
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
@@
 We define the proportion of pre-symptomatic transmission as ...
-We define the onset-to-isolation delay distribution as short or long following the approach from Hellewell et al.
-(2020).
+We define the onset-to-isolation delay distribution as short or long following the approach from \cite{hellewellFeasibilityControllingCOVID192020}.
 
-\subsection{Simulation scenarios}
+\subsection*{Simulation scenarios}
 
@@
 
-\section{Results}
+\section*{Results}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 
-\section{References}
-
-\begin{itemize}
-\item Farrington and Grant (1999) The distribution of time to extinction in subcritical branching processes: applications to outbreaks of infectious disease.
-\item Fraser et al. (2004) Factors that make an infectious disease outbreak controllable
-\item Hellewell et al. (2020) Feasibility of controlling COVID-19 outbreaks by isolation of cases and contacts
-\item Kucharski et al. (2016) Effectiveness of Ring Vaccination as Control Strategy for Ebola Virus Disease
-\item Kucharski et al. (2020) Early dynamics of transmission and control of COVID-19: a mathematical modelling study
-\item Nishiura and Inaba (2011) Estimation of the incubation period of influenza A (H1N1-2009) among imported cases: Addressing censoring using outbreak data at the origin of importation
-\item Yang et al. (2007) Detecting Human-to-Human Transmission of Avian Influenza A (H5N1)
-\item Ward et al. (2024) Estimates of epidemiological parameters for H5N1 influenza in humans: a rapid review
-\end{itemize}
+\bibliographystyle{plainnat}
+\bibliography{FluTracer.bib}
 

### 2025-04-14 inst/scripts/run_analysis.R
update run_analysis script to vary incubation period

@@ -1,54 +1,89 @@
 library(ringbp)
-library(dplyr)
-library(tidyr)
 library(purrr)
 library(future)
-library(tibble)
 library(data.table)
+library(epiparameter)
 
-scenarios <- tidyr::expand_grid(
-  ## Put parameters that are grouped by disease into this data.frame
-  delay_group = list(tibble::tibble(
-    delay = c("SARS", "Wuhan"),
-    delay_shape = c(1.651524, 2.305172),
-    delay_scale = c(4.287786, 9.483875)
-  )),
-  k_group = list(tibble::tibble(
-    theta = c("<1%", "15%"),
-    k = c(30, 1.95)
-  )),
-  index_R0 = c(1.1, 1.5),
-  quarantine = FALSE,
-  prop.asym = c(0, 0.1),
-  control_effectiveness = seq(0, 1, 0.2),
-  num.initial.cases = c(5, 10)) %>%
-  tidyr::unnest("k_group") %>%
-  tidyr::unnest("delay_group") %>%
-  dplyr::mutate(scenario = 1:dplyr::n())
+h5n1_weibull_params <- epiparameter::convert_summary_stats_to_params(
+  "weibull", mean = 3.3, sd = 1.5
+)
+h7n9_weibull_params <- epiparameter::convert_summary_stats_to_params(
+  "weibull", mean = 3.1, sd = 1.4
+)
 
-## Parameterise fixed paramters
-sim_with_params <- purrr::partial(ringbp::scenario_sim,
-                                  cap_max_days = 365,
-                                  cap_cases = 500,
-                                  r0isolated = 0,
-                                  disp.iso = 1,
-                                  disp.com = 0.16)
-
-
-## Set up multicore if using see ?future::plan for details
-## Use the workers argument to control the number of cores used.
-future::plan("multisession")
-
-scenarios <- as.data.table(scenarios)
-
-## Run paramter sweep
-run_time <- system.time(
-  sweep_results <- ringbp::parameter_sweep(
-    scenarios,
-    sim_fn = sim_with_params,
-    samples = 100
+# Put parameters that are grouped by disease into this data.table
+scenarios <- data.table(
+  expand.grid(
+    delay_group = list(data.table(
+      delay = c("SARS", "Wuhan"),
+      onset_to_isolation = c(
+        \(x) stats::rweibull(n = x, shape = 1.651524, scale = 4.287786),
[…]

### 2025-04-14 inst/scripts/plot_prop_outbreak_control_onset_to_isolation.R
update prop_outbreak_control plotting scripts

@@ -14,36 +14,35 @@ dt <- as.data.table(res)
 dt_data <- rbindlist(dt$data)
 dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
 
-prop_outbreak_control_num_init_cases <- dt_data[
-  index_R0 == 1.5 & theta == "15%" & delay == "SARS" & prop.asym == 0,
-  .(control_effectiveness, num.initial.cases, pext, subtype)
+prop_outbreak_control_onset_to_isolation <- dt_data[
+  num.initial.cases == 10 & theta == "15%" & index_R0 == 1.5 & prop.asym == 0,
+  .(control_effectiveness, onset_to_isolation, delay, pext, subtype)
 ]
 
 # convert to percentages for plotting
-prop_outbreak_control_num_init_cases[, control_effectiveness := control_effectiveness * 100]
-prop_outbreak_control_num_init_cases[, pext := pext * 100]
+prop_outbreak_control_onset_to_isolation[, control_effectiveness := control_effectiveness * 100]
+prop_outbreak_control_onset_to_isolation[, pext := pext * 100]
 
-prop_outbreak_control_num_init_cases_plot <- ggplot2::ggplot(
-  data = prop_outbreak_control_num_init_cases
+prop_outbreak_control_onset_to_isolation_plot <- ggplot2::ggplot(
+  data = prop_outbreak_control_onset_to_isolation
 ) +
   ggplot2::geom_point(
     mapping = ggplot2::aes(
       x = control_effectiveness,
       y = pext,
-      fill = as.factor(num.initial.cases),
+      fill = as.factor(delay),
       shape = as.factor(subtype)
     ),
-    size = 3,
-    stroke = 0.75
+    size = 3
   ) +
   ggplot2::geom_line(
     mapping = ggplot2::aes(
       x = control_effectiveness,
       y = pext,
-      colour = as.factor(num.initial.cases),
+      colour = as.factor(delay),
       linetype = as.factor(subtype)
     ),
-    size = 0.75
+    linewidth = 0.75
   ) +
   ggplot2::scale_x_continuous(
     name = "Contacts traced (%)",
@@ -53,12 +52,12 @@ prop_outbreak_control_num_init_cases_plot <- ggplot2::ggplot(
     name = "Simulated outbreaks controlled (%)",
     limits = c(0, 100)
   ) +
[…]

### 2025-04-14 inst/scripts/plot_prop_outbreak_control_reproduction_number.R

@@ -1,5 +1,6 @@
 library(data.table)
 library(ggplot2)
+library(dplyr)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
 sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
@@ -15,7 +16,7 @@ dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
 
 prop_outbreak_control <- dt_data[
   num.initial.cases == 10 & theta == "15%" & delay == "SARS" & prop.asym == 0,
-  .(control_effectiveness, index_R0, pext)
+  .(control_effectiveness, index_R0, pext, subtype)
 ]
 
 # convert to percentages for plotting
@@ -27,19 +28,20 @@ prop_outbreak_control_plot <- ggplot2::ggplot(data = prop_outbreak_control) +
     mapping = ggplot2::aes(
       x = control_effectiveness,
       y = pext,
-      colour = as.factor(index_R0)
+      colour = as.factor(index_R0),
+      linetype = as.factor(subtype)
     ),
-    size = 0.75
+    linewidth = 0.75
   ) +
   ggplot2::geom_point(
     mapping = ggplot2::aes(
       x = control_effectiveness,
       y = pext,
-      fill = as.factor(index_R0)
+      fill = as.factor(index_R0),
+      shape = as.factor(subtype)
     ),
-    shape = 21,
     size = 3,
-    colour = "black"
+    stroke = 0.75
   ) +
   ggplot2::scale_x_continuous(
     name = "Contacts traced (%)",
@@ -51,12 +53,23 @@ prop_outbreak_control_plot <- ggplot2::ggplot(data = prop_outbreak_control) +
   ) +
   ggplot2::scale_colour_manual(values = c("#f4b301", "#db1048")) +
   ggplot2::scale_fill_manual(values = c("#f4b301", "#db1048")) +
+  ggplot2::scale_shape_manual(values = c(21, 22, 24)) +
   ggplot2::labs(
     colour = "Reproduction Number (R)",
-    fill = "Reproduction Number (R)"
+    fill = "Reproduction Number (R)",
+    shape = "Pathogen Subtype",
+    linetype = "Pathogen Subtype"
+  ) +
+  ggplot2::guides(
+    fill = ggplot2::guide_legend(
+      override.aes = list(shape = 21)
+    )
   ) +
   ggplot2::theme_bw() +
[…]

### 2025-04-14 inst/scripts/prop_outbreak_control_onset_to_isolation_plot.R

@@ -1,42 +1,0 @@
-library(ggplot2)
-library(data.table)
-
-# read simulation results saved by running inst/scripts/run_analysis.R
-
-sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
-
-res <- sweep_results %>%
-  group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
-  ungroup(scenario)
-
-dt <- as.data.table(res)
-dt_data <- rbindlist(dt$data)
-dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
-
-prop_outbreak_control_onset_to_isolation <- dt_data %>%
-  dplyr::filter(num.initial.cases == 20,
-                theta == "15%",
-                index_R0 == 2.5,
-                prop.asym == 0) %>%
-  dplyr::mutate(delay = factor(delay,
-                               levels = c("SARS", "Wuhan"),
-                               labels = c("Short", "Long"))) %>%
-  ggplot2::ggplot(ggplot2::aes(x = control_effectiveness,
-                               y = pext,
-                               color = as.factor(delay))) +
-  ggplot2::geom_line(size = 0.75) +
-  ggplot2::geom_point(shape = 21,
-                      col = "black",
-                      aes(fill = as.factor(delay)),
-                      size = 3) +
-  ggplot2::scale_fill_manual(guide = "none",
-                             values = c("black", "forestgreen")) +
-  ggplot2::scale_color_manual(values = c("black", "forestgreen"),
-                              name = "Onset to\nisolation\ndelay") +
-  cowplot::theme_cowplot()
-
-ggplot2::ggsave(
-  file.path("inst", "plots", "prop_outbreak_control_onset_to_isolation.png"),
-  plot = prop_outbreak_control_onset_to_isolation
-)

### 2025-04-14 inst/paper/main.tex
add onset-to-isolation plot to paper and re-render pdf

@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_onset_to_isolation.png}
+\caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an onset-to-isolation delay distribution that is either SARS-like, or COVID-19-like (from Wuhan).
+This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the number of initial infections is 10.}
+\label{fig:prop-outbreak-control-onset-to-isolation}
+\end{figure}
+
 \section*{Discussion}

### 2025-04-14 inst/scripts/run_analysis.R
expand parameter space in run_analysis script

@@ -22,8 +22,8 @@ scenarios <- data.table(
       )
     )),
     k_group = list(data.table(
-      theta = c("<1%", "15%"),
-      k = c(30, 1.95)
+      theta = c("<1%", "15%", "30%"),
+      k = c(30, 1.95, 0.7)
     )),
     incubation_period_group = list(data.table(
       subtype = c("H1N1", "H5N1", "H7N9"),
@@ -41,10 +41,10 @@ scenarios <- data.table(
         )
       )
     )),
-    index_R0 = c(1.1, 1.5),
+    index_R0 = c(1.1, 1.5, 2.5, 3.5),
     prop.asym = c(0, 0.1),
     control_effectiveness = seq(0, 1, 0.2),
-    num.initial.cases = c(5, 10),
+    num.initial.cases = c(5, 20, 40),
     quarantine = FALSE
   )
 )
@@ -77,7 +77,7 @@ sim_with_params <- purrr::partial(
 
 # Set up multicore if using see ?future::plan for details
 # Use the workers argument to control the number of cores used.
-# future::plan("multisession")
+future::plan("multisession", workers = 8)
 
 # Run parameter sweep
 sweep_results <- ringbp::parameter_sweep(

### 2025-04-14 inst/scripts/plot_prop_outbreak_control_num_init_cases.R
update plotting scripts

@@ -7,7 +7,7 @@ sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
 
 res <- sweep_results %>%
   group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
+  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 500)) %>%
   ungroup(scenario)
 
 dt <- as.data.table(res)
@@ -15,8 +15,8 @@ dt_data <- rbindlist(dt$data)
 dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
 
 prop_outbreak_control_num_init_cases <- dt_data[
-  index_R0 == 1.5 & theta == "15%" & delay == "SARS" & prop.asym == 0,
-  .(control_effectiveness, num.initial.cases, pext, subtype)
+  theta == "15%" & delay == "SARS" & prop.asym == 0,
+  .(index_R0, control_effectiveness, num.initial.cases, pext, subtype)
 ]
 
 # convert to percentages for plotting
@@ -43,7 +43,7 @@ prop_outbreak_control_num_init_cases_plot <- ggplot2::ggplot(
       colour = as.factor(num.initial.cases),
       linetype = as.factor(subtype)
     ),
-    size = 0.75
+    linewidth = 0.75
   ) +
   ggplot2::scale_x_continuous(
     name = "Contacts traced (%)",
@@ -53,9 +53,19 @@ prop_outbreak_control_num_init_cases_plot <- ggplot2::ggplot(
     name = "Simulated outbreaks controlled (%)",
     limits = c(0, 100)
   ) +
-  ggplot2::scale_colour_manual(values = c("#ffbe6a", "#40b0a7")) +
-  ggplot2::scale_fill_manual(values = c("#ffbe6a", "#40b0a7")) +
+  ggplot2::scale_colour_manual(values = c("#92C5DE", "#F4A582", "#CA0020")) +
+  ggplot2::scale_fill_manual(values = c("#92C5DE", "#F4A582", "#CA0020")) +
   ggplot2::scale_shape_manual(values = c(21, 22, 24)) +
+  ggplot2::scale_linetype_manual(values = c(1, 1, 1)) +
+  ggplot2::facet_wrap(
+    facets = ggplot2::vars(index_R0),
+    labeller = ggplot2::as_labeller(c(
+        `1.1` = "R = 1.1",
[…]

### 2025-04-14 inst/scripts/plot_prop_outbreak_control_onset_to_isolation.R

@@ -7,7 +7,7 @@ sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
 
 res <- sweep_results %>%
   group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
+  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 500)) %>%
   ungroup(scenario)
 
 dt <- as.data.table(res)
@@ -15,7 +15,7 @@ dt_data <- rbindlist(dt$data)
 dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
 
 prop_outbreak_control_onset_to_isolation <- dt_data[
-  num.initial.cases == 10 & theta == "15%" & index_R0 == 1.5 & prop.asym == 0,
+  num.initial.cases == 20 & theta == "15%" & index_R0 == 1.5 & prop.asym == 0,
   .(control_effectiveness, onset_to_isolation, delay, pext, subtype)
 ]
 
@@ -52,9 +52,10 @@ prop_outbreak_control_onset_to_isolation_plot <- ggplot2::ggplot(
     name = "Simulated outbreaks controlled (%)",
     limits = c(0, 100)
   ) +
-  ggplot2::scale_colour_manual(values = c("#10559a", "#db4c77")) +
-  ggplot2::scale_fill_manual(values = c("#10559a", "#db4c77")) +
+  ggplot2::scale_colour_manual(values = c("#92C5DE", "#F4A582")) +
+  ggplot2::scale_fill_manual(values = c("#92C5DE", "#F4A582")) +
   ggplot2::scale_shape_manual(values = c(21, 22, 24)) +
+  ggplot2::scale_linetype_manual(values = c(1, 1, 1)) +
   ggplot2::labs(
     colour = "Onset-to-isolation",
     fill = "Onset-to-isolation",

### 2025-04-14 inst/scripts/plot_prop_outbreak_control_reproduction_number.R

@@ -7,7 +7,7 @@ sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
 
 res <- sweep_results %>%
   group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 5000)) %>%
+  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 500)) %>%
   ungroup(scenario)
 
 dt <- as.data.table(res)
@@ -15,7 +15,7 @@ dt_data <- rbindlist(dt$data)
 dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
 
 prop_outbreak_control <- dt_data[
-  num.initial.cases == 10 & theta == "15%" & delay == "SARS" & prop.asym == 0,
+  num.initial.cases == 20 & theta == "15%" & delay == "SARS" & prop.asym == 0,
   .(control_effectiveness, index_R0, pext, subtype)
 ]
 
@@ -51,9 +51,10 @@ prop_outbreak_control_plot <- ggplot2::ggplot(data = prop_outbreak_control) +
     name = "Simulated outbreaks controlled (%)",
     limits = c(0, 100)
   ) +
-  ggplot2::scale_colour_manual(values = c("#f4b301", "#db1048")) +
-  ggplot2::scale_fill_manual(values = c("#f4b301", "#db1048")) +
+  ggplot2::scale_color_brewer(palette = "RdBu", direction = -1) +
+  ggplot2::scale_fill_brewer(palette = "RdBu", direction = -1) +
   ggplot2::scale_shape_manual(values = c(21, 22, 24)) +
+  ggplot2::scale_linetype_manual(values = c(1, 1, 1)) +
   ggplot2::labs(
     colour = "Reproduction Number (R)",
     fill = "Reproduction Number (R)",

### 2025-04-14 R/plot_prop_outbreak_control_var_reproduction_number.R
add plotting script for subtype variance

@@ -1,0 +1,67 @@
+library(data.table)
+library(ggplot2)
+library(dplyr)
+
+# read simulation results saved by running inst/scripts/run_analysis.R
+sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
+
+res <- sweep_results %>%
+  group_by(scenario) %>%
+  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 500)) %>%
+  ungroup(scenario)
+
+dt <- as.data.table(res)
+dt_data <- rbindlist(dt$data)
+dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
+
+# convert to percentages for plotting
+dt_data[, control_effectiveness := control_effectiveness * 100]
+dt_data[, pext := pext * 100]
+
+# calculate sd by subtype, R and control
+subtype_sd <- dt_data[
+  , .(subtype_sd = sd(pext)),
+  by = .(index_R0, prop.asym, control_effectiveness, num.initial.cases, delay, theta)
+]
+
+prop_outbreak_control_var_reproduction_number_plot <- ggplot2::ggplot(
+  data = subtype_sd
+) +
+  ggplot2::geom_point(
+    mapping = ggplot2::aes(
+      x = as.factor(control_effectiveness),
+      y = subtype_sd,
+      fill = as.factor(index_R0)
+    ),
+    shape = 21,
+    size = 2,
+    stroke  = 0.75,
+    position = ggplot2::position_jitter(width = 0.2, height = 0)
+  ) +
+  ggplot2::scale_x_discrete(
+    name = "Contacts traced (%)"
+  ) +
+  ggplot2::scale_y_continuous(
+    name = "Standard deviation between subtype\nsimulated outbreaks controlled (%)",
+  ) +
+  ggplot2::scale_fill_brewer(palette = "RdBu", direction = -1) +
+  ggplot2::labs(
+    fill = "Reproduction Number (R)"
+  ) +
+  ggplot2::guides(
+    fill = ggplot2::guide_legend(
+      override.aes = list(shape = 21)
+    )
+  ) +
+  ggplot2::theme_bw() +
+  ggplot2::theme(legend.position = "bottom")
+
+ggplot2::ggsave(
+  file.path("inst", "plots", "prop_outbreak_control_var_reproduction_number.png"),
+  plot = prop_outbreak_control_var_reproduction_number_plot,
+  device = "png",
[…]

### 2025-04-14 inst/paper/main.tex
update paper and re-render pdf

@@
 
-To assess the efficaciousness of contact tracing and isolation to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{{ringbp}} \citep{hellewellRingbpSimulateEvaluate2025}.
+To assess the efficacy of contact tracing and isolation to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025}.
 The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility).
@@
 
-\begin{itemize}
-
-\item H1N1-like
-\item H5N1 (slightly increased reproduction number, R~1)
-\item H7N9 (if parameters are available)
-
-\end{itemize}
-
-Figure for parameter space for flu-like pandemic
+<Insert Figure for parameter space for flu-like pandemic>
 
@@
 \end{figure}
+
+\clearpage
+
+\subsection*{Simulation scenarios}
+
+To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of contact tracing and isolation we set up a simulation experiment.
+We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmissin, and number of initial cases.
+\\
+
+The onset-to-isolation delay distribution determines the time between a case developing symptoms and then becoming isolated.
+We assume that an isolated case cannot transmit, so isolation is assumed to be 100\% effective once enacted ($R_{isolated} = 0$).
+The onset-to-isolation delay is parameterised with two scenarios: a short, SARS-like, delay (Donnelly et al.) and a longer COVID-19-like delay (Li et al.) (Figure \ref{fig:onset-to-isolation}).
+Both onset-to-isolation delays are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:onset-to-isolation}).
+These two empirically informed delay distributions provide a guide to possible response times in the case of an influenza outbreak and will enable us to evaluate if such delays are too long to contain a disease that can transmit rapidly such as influenza.
+\\
 
@@
 
+\clearpage
+
+We define the proportion of pre-symptomatic transmission follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skewed-normal distribution to sample the infection times using the incubation period and a skew parameter ($\theta$).
+The skew parameter controls the proportion of pre-symptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}).
+We parameterised the skew-normal distribution with a skew of 0.7, 1.95 and 30, these correspond to a pre-symptomatic transmission percentage of 30\%, 15\% and $<1\%$, respectively.
+Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective contact tracing is hypothesied to be.
+\\
+
+The proportion of asymptomatic infections is assumes to be rare, so we test scenarios with zero asymptomatic tranmission and another with 10\% of cases being asymptomatic. \\
+
 \begin{figure}[ht]
@@
 
-We extract the incubation period or serial interval for each influenza subtype.
-We define the proportion of pre-symptomatic transmission as ...
-We define the onset-to-isolation delay distribution as short or long following the approach from \cite{hellewellFeasibilityControllingCOVID192020}.
+\clearpage
 
-\subsection*{Simulation scenarios}
+We simulated scenarios with either 5, 20 or 40 initial cases.
+This is to test whether contact tracing can contain outbreaks that may have already accumulated cases before contact tracing is started, or represents importation of multiple cases in a single event, such as a flight with multiple infectors (Khanh et al.
+2020).
+\\
 
-We simulated scenarios with either 5 or 10 initial cases.
+Some parameters we fixed across all simulations.
+The dispersion parameter for the negative binomial offspring distribution was fixed at 1 for isolated cases, 0.16 for subclinical and community cases.
+For computational efficiency we capped the maximum number of days the simulation could run for to 365.
+This is substantially longer than the 12-16 week window for evaluating outbreak control so does not influence the results.
+Additionally, we limit simulations to 500 cases.
+This is an arbitrary threshold we judge to be large epidemic that is not controllable.
+\\
 
@@
 
-The incubation period for H1N1 is noticable shorter than the other two subtypes, which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:incub}).
+The incubation period for H1N1 is noticable shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:incub}). \\
 
-Predictably, a higher reproduction number for the outbreak results in a smaller percentage of outbreaks controlled at a given percentage of control effectiveness.
+Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number.
+A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of control effectiveness.
 (Figure \ref{fig:prop-outbreak-control-R}).
-The disparity of percentage outbreaks controlled becomes zero as the percentage of contacts traced exceeds 80\% (Figure \ref{fig:prop-outbreak-control-R}).
-When $R$ is only slightly above unity but below 2, the percentage of outbreaks controlled at 50\% of contacts traced and isolated is over 90\%, highlighting the effectiveness of this intervention for infectious diseases are not highly transmissible between humans.
+There is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of contact tracing and isolation for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}).
+As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e.
+0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once control effectiveness exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}.
+Once the reproduction number is at 2.5 or 3.5, an outbreak is only controlled in at least 50\% of scenarios with contact tracing and isolation at greater than 60\% (Figure \ref{fig:prop-outbreak-control-R}).
+\\
+
+The incubation period influences the efficacy of tracing and isolation.
+This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e.
+not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}).
+The shortest incubation period, H1N1, predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection limit our ability to effectively tracing and isolate their potential infectees in time to prevent disease transmission.
+The difficulty to control an pandemic outbreak with a short incubation period is shown by the $\sim 40\%$ of simulated H1N1 outbreaks with 100\% control effectiveness that were not contained when $R = 3.5$, while longer incubation period outbreaks have $>80\%$ controlled when 100\% of contacts are traced and isolated (Figure \ref{fig:prop-outbreak-control-R}).
+Overall, both the transmissibility and incubation period impact the feasibility to control an outbreak.
+\\
+
+The variance between the influenza subtypes at each percentage of contacts traced and isolated varies (i.e.
+difference between circle, triangle and square points at each contacts traced \% in Figure \ref{fig:prop-outbreak-control-R}).
+This variance is greatest at intermediate (20\% - 80\%) contact tracing effectiveness and intermediate values of $R$ (1.5 and 2.5) (Figure \ref{fig:prop-outbreak-control-R}).
+This could indicate a substantive role for incubation period durartion in the control of disease transmission.
+However, this variance in proportion of outbreaks controlled between pathogen subtypes does not show a strong pattern across all scenarios simulated (Figure \ref{fig:prop-outbreak-control-var-R}).
+There is a weak pattern of increased variance in controllability between subtypes at intermediate values of $R$, overall it seems that incubation period causes differences in control-by-isolation effectivness when not at the boundary of epidemic controllability.
+In other words, differences between influenza subtypes arise when not in a situation of high transmissibility and low contact tracing effectivness that cannot contain any outbreaks, nor low transmissibility and high contact tracing effectivenss easily contain all outbreaks (Figure \ref{fig:prop-outbreak-control-var-R}).
+\\
 
@@
 
+\clearpage
+
+The ability to control an epidemic that can readily transmit human-to-human is moderated by the initial number of infectious individuals that seed the outbreak.
+The higher the number of cases the less chance that stochastic extinction of a transmission chain will end the whole epidemic.
+For an influenza pandemic, the number of initial infected individuals influences efficacy of control-by-isolation at all transmissibility scenarios evaluated ($R \in \{1.1, 1.5, 2.5, 3.5\}$, Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+When $R = 1.1$ most outbreaks can be controlled without contact tracing and isolation, even with 40 initial infectors Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+When $R = 1.5$ the increase in the number of initial infections from 5 to 20 has a large decrease in outbreaks controlled at low levels of contact tracing, the same pattern when the number of initial infectors is twofold larger at 40, but these differences in outbreak control diminish as contact tracing increases towards 100\% (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+When the number of cases seeding the outbreak is $>20$ at $R>2.5$ a large proportion of outbreaks become uncontrollable until contact tracing reaches $>60\%$.
+In the worst case transmissibility scenario ($R = 3.5$) if contact tracing is not activated until at least 20 individuals are infected then close to 100\% of contacts with need to be traced in order to control the majority of outbreaks (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+\\
+
 \begin{figure}[ht]
@@
 \end{figure}
+
+\clearpage
 
@@
 \item This study has looked at the feasibility and effectiveness of contact tracing and isolation for influenza subtypes that have cause epidemic outbreaks.
+\item Outbreaks controlled at 50\% of contacts traced and isolated is over ??\%, highlighting the effectiveness of this intervention for infectious diseases are not highly transmissible between humans.
 \item Discuss the existing flu vaccines and antivirals that can be used an interventions in addition to NPIs
@@
 
+\clearpage
+
+\section*{Supplementary Material}
+
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_var_reproduction_number.png}
+\caption{The standard deviation of the percentage of outbreaks controlled between influenza subtypes (H1N1, H5N1, H7N9), across percentages of contacts traced.
+Points are jittered horizontally reduce overlap.}
+\label{fig:prop-outbreak-control-var-R}
+\end{figure}
+
 \end{document}

### 2025-04-15 inst/paper/main.tex
add points on contact tracing speed and network structure to paper

@@
 In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model.
-This can be used to define the speed and effectiveness of isolation given an epidemic scenario to inform outbreak response complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals (ref).
+In effect the ability to control an epidemic with contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021).
+The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval (Ward et al., 2024).
+This study looks to define the speed and effectiveness of isolation given an epidemic scenario to inform outbreak response complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals (ref).
 
@@
 However, these results can be used to inform other influenza subtypes that exhibit similar epidemiological characteristics in transmissibility and severity from past outbreaks (H2N2 and H3N2) and still exist in animal reservoirs for potential spillovers back into human populations (ref).
+\item This study used a branching process model to simulate individual-level transmission and intervention.
+The model assumes an infinite population size with no interdependence of contact tracing and isolation between cases whereby a case may be in contact with multiple infectors and only needs to be traced in time by one contact to prevent transmission, as could happened in a fixed random network transmission model.
+It is unlikely that this model structure difference would have a considerable different on our results for the control of disease outbreaks with contact tracing.
+We suppose that because our model requires isolation sampled for a single individual, our results are more conservative on the ability to control outbreaks, and that using a random network model would indicate an increased effectivness of contact tracing (Juul and Strogatz, 2023).
+The infinite population assumption also means that although isolated individuals are assumed to not transmit, there are always other individuals that non-isolated infectors can transmit to, whereas a fixed random network model that assumes optimal indefinite isolation will block of parts of the network and overestimate contact tracing effectiveness (Juul and Strogatz, 2023).
 \end{itemize}

### 2025-04-15 inst/paper/main.tex
update reproduction number and asymptomatic cases info in paper

@@
 It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation become the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (ref).
-However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may not be the case that contact tracing and isolation is unable to control influenza.
+However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may not be the case that contact tracing and isolation is unable to control influenza.
 Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}.
@@
 
-The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5 and 2.0.
-We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g.
+The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5.
+We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g.
 recombination of subtypes (ref)) enhances the human-to-human transmissibility of these pathogens.
@@
 
-The proportion of asymptomatic infections is assumes to be rare, so we test scenarios with zero asymptomatic tranmission and another with 10\% of cases being asymptomatic. \\
+The proportion of pandemic-potential influenza infections that are asymptomatic is assumes to be rare.
+There is evidence from past outbreaks of H1N1 and ???
+that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009}.
+We ran scenarios with zero asymptomatic tranmission and another with 10\% of cases being asymptomatic.
+\\
 
@@
 The infinite population assumption also means that although isolated individuals are assumed to not transmit, there are always other individuals that non-isolated infectors can transmit to, whereas a fixed random network model that assumes optimal indefinite isolation will block of parts of the network and overestimate contact tracing effectiveness (Juul and Strogatz, 2023).
+\item The reproduction number for H5N1 (including clade 2.3.4.4b) and H7N9 influenza subtypes have been estimated to be far below 1 (Ward et al., Garg et al., 2025).
+Therefore, the transmissibility scenarios explored in this study are speculative and assume that pathogen evolution increases human-to-human transmission potential.
+If transmissibility of these avian influenza subtypes remains constant over time then outbreaks are expected to be subcritical and will not produce epidemics and thus not require interventions.
+If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence contact tracing and isolation while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
 \end{itemize}

### 2025-04-15 inst/paper/main.tex
add refs for reassortment and mutations to paper
polish interventions intro paragraph and re-render pdf

@@
 
-There are a variety of interventions that can be enacted to try and prevent transmission of an infectious disease, including both pharmaceutical and non-pharmaceutical interventions (NPIs).
-Early in an outbreak NPIs are often used, as especially for outbreaks of a novel pathogen before pharmaceutical inventions are available.
-These interventions range in their specificity, from targeted contact tracing which aims to tackle specific transmission pathways, to closure of school and gathering, to regional or national lockdowns affecting everyone (ref).
+A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs).
+Early in an outbreak -- especially in the case of a novel pathogen -- NPIs are often the first line of defence, as pharmaceutical options may not yet be available.
+These interventions vary in scope and specificity, from targeted measures such as contact tracing and isolation, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (ref).
+The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005).
+Even for highly transmissible pathogens (e.g.
+$R>3$) with short generation times (e.g.
+$<3$ days), swift and effective interventions can achieve containment.
 \\
@@
 It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined is needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
-Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016} (ref).
+Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016} (Whittaker et al., 2024) (ref).
 Isolation of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics (SARS ref; other refs).
@@
 We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g.
-recombination of subtypes (ref)) enhances the human-to-human transmissibility of these pathogens.
+genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens.
 Current estimates for H5N1 and H7N9 subtypes are below unity (i.e.

### 2025-04-30 inst/scripts/plot_patho_param_space.R
add plot_patho_param_space script

@@ -1,0 +1,208 @@
+# number of random samples from distribution
+n <- 1000
+
+flu_param_space <- data.frame(
+  pathogen = "influenza",
+  R = rnorm(n = 2000, mean = 5, sd = 2),
+  prop_asym = rnorm(n = n, mean = 40, sd = 2)
+)
+sars_param_space <- data.frame(
+  pathogen = "SARS-CoV-1",
+  R = rnorm(n = n, mean = 3, sd = 1),
+  prop_asym = rnorm(n = n, mean = 5, sd = 2)
+)
+smallpox_param_space <- data.frame(
+  pathogen = "Smallpox",
+  R = rnorm(n = n, mean = 8, sd = 1),
+  prop_asym = rnorm(n = n, mean = 10, sd = 2)
+)
+hiv_param_space <- data.frame(
+  pathogen = "HIV",
+  R = rnorm(n = n, mean = 4, sd = 1),
+  prop_asym = rnorm(n = n, mean = 90, sd = 2)
+)
+covid_param_space <- data.frame(
+  pathogen = "SARS-CoV-2",
+  R = rnorm(n = n, mean = 2, sd = 1),
+  prop_asym = rnorm(n = n, mean = 40, sd = 2)
+)
+
+flu_coords <- c(x = mean(flu_param_space$prop_asym), y = mean(flu_param_space$R))
+sars_coords <- c(x = mean(sars_param_space$prop_asym), y = mean(sars_param_space$R))
+hiv_coords <- c(x = mean(hiv_param_space$prop_asym), y = mean(hiv_param_space$R))
+smallpox_coords <- c(x = mean(smallpox_param_space$prop_asym), y = mean(smallpox_param_space$R))
+covid_coords <- c(x = mean(covid_param_space$prop_asym), y = mean(covid_param_space$R))
+
+param_space <- rbind(
+  flu_param_space,
+  sars_param_space,
+  smallpox_param_space,
+  hiv_param_space,
+  covid_param_space
+)
+
+# density plot
+ggplot2::ggplot(data = param_space) +
+  ggplot2::stat_density_2d(
+    mapping = ggplot2::aes(
+      x = prop_asym,
+      y = R,
+      fill = ggplot2::after_stat(density)
+    ),
+    geom = "raster",
+    contour = FALSE
+  ) +
+  ggplot2::annotate(
+    geom = "rect",
+    xmin = 0,
+    xmax = 35,
[…]

### 2025-04-30 inst/paper/main.tex
update paper tex and pdf with patho_param_space plot and minor edits to intro and discussion points

@@
 The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005).
-Even for highly transmissible pathogens (e.g.
+For highly transmissible pathogens (e.g.
 $R>3$) with short generation times (e.g.
-$<3$ days), swift and effective interventions can achieve containment.
+$<3$ days), swift and effective interventions are required to achieve containment.
 \\
@@
 
-<Insert Figure for parameter space for flu-like pandemic>
+<Insert Figure for ringbp schematic>
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/patho_param_space.png}
+\caption{The pathogen characteristics, in terms of tranmissibility ($R$) and proportion of disease transmission that occurs before the onset of symptoms (including cases that are entirely asymptomatic).
+The figure includes: Influenza, SARS-CoV-1, SARS-CoV-2, Smallpox, Ebola, Marburg, and HIV shown by green shaded areas.
+The size of the shaded area reflects the differences in pathogen characteristics, both between variants and outbreaks, the blurred edges reflects uncertainty in parameter estimates.
+The parameter space of transmissibility and pre- and asymptomatic transmission explored in this paper is shown in the orange box.}
+\label{fig:patho-param-space}
+\end{figure}
 
@@
 \item Digital tracing has capacity for large-scale contact tracing (pingdemic), also rapid tests are a complementary approach for limiting spread.
+\item In this study we model contact tracing and isolation as the sole intervention in containing an influenza outbreak.
+This is to provide an assessment of its effectiveness for rapidly transmission and potentially highly transmissible subtypes.
+However, in reality, pandemic preparedness to a pandemic potential influenza would not need to deploy interventions, such as contact tracing individually or sequentially, instead they can be used in combination, such as antivirals (e.g.
+oseltamivir), rapid tests, contact tracing, and vaccination.
 \item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic.
@@
 If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence contact tracing and isolation while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+\item This study has evaluated the ability to control the outbreak of an infectious disease independent on its severity.
+Infection severity and fatality risks are of course critical when formulating a pandemic response plan.
+By understanding the efficacy of contact trace and isolation for a given epidemiological characterstic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.
 \end{itemize}

### 2025-04-30 inst/paper/main.tex
update onset-to-isolation text and add first draft of model schematic figure and update pdf

@@
 \usepackage{nameref}
+\usepackage{tikz}
 
@@
 
-<Insert Figure for ringbp schematic>
+\begin{figure}[ht]
+\centering
+{\sffamily
+  \begin{tikzpicture}[every node/.style={font=\tiny}]
+  % Labels above the A row
+  \node at (0,0.5) {Infected};
+  \node at (2,0.5) {Symptomatic};
+  \node at (5,0.5) {Isolated};
+
+  % Top row: A circles with red background
+  \fill[red!30] (0,0) circle (5pt);
+  \draw (0,0) circle (5pt);
+  \node at (0,0) {A};
+
+  \fill[red!30] (2,0) circle (5pt);
+  \draw (2,0) circle (5pt);
+  \node at (2,0) {A};
+
+  \fill[red!30] (5,0) circle (5pt);
+  \draw (5,0) circle (5pt);
+  \node at (5,0) {A};
+
+  % Horizontal line between A circles
+  \draw (0,0) -- (2,0);
+  \draw (2,0) -- (5,0);
+
+  % Individual A delays
+  \draw[<->] (0.5,-0.3) -- (1.5,-0.3);
+  \node at (1,-0.5) {Incubation period};
+  \draw[<->] (2,-0.3) -- (5,-0.3);
+  \node at (3.5,-0.5) {Onset-to-isolation};
+
+  % Labels above the B row
+  \node at (4,-1.5) {Infected};
+  \node at (6,-1.5) {Symptomatic};
+  \node at (8,-1.5) {Isolated};
+
+  % Middle row: B circles with blue background
+  \fill[blue!30] (4,-2) circle (5pt);
+  \draw (4,-2) circle (5pt);
+  \node at (4,-2) {B};
+
+  \fill[blue!30] (6,-2) circle (5pt);
+  \draw (6,-2) circle (5pt);
+  \node at (6,-2) {B};
+
+  \fill[blue!30] (8,-2) circle (5pt);
+  \draw (8,-2) circle (5pt);
+  \node at (8,-2) {B};
+
+  % Horizontal line between B circles
+  \draw (4,-2) -- (6,-2);
+  \draw (6,-2) -- (8,-2);
+
+  % Labels above the C row
+  \node at (4.5,-3.5) {Infected};
+  \node at (7.5,-3.5) {Symptomatic};
+  \node at (9,-3.5) {Isolated};
+
+  % Bottom row: C circles with green background
+  \fill[green!30] (4.5,-4) circle (5pt);
+  \draw (4.5,-4) circle (5pt);
+  \node at (4.5,-4) {C};
+
+  \fill[green!30] (7.5,-4) circle (5pt);
+  \draw (7.5,-4) circle (5pt);
+  \node at (7.5,-4) {C};
+
+  \fill[green!30] (9,-4) circle (5pt);
+  \draw (9,-4) circle (5pt);
+  \node at (9,-4) {C};
+
+  % Horizontal line between C circles
+  \draw (4.5,-4) -- (7.5,-4);
+  \draw (7.5,-4) -- (9,-4);
+
+  % Labels above the D row
+  \node at (8,-5.5) {Infected};
+
+  % Individual C circles with green background
+  \fill[teal!30] (8,-6) circle (5pt);
+  \draw (8,-6) circle (5pt);
+  \node at (8,-6) {D};
+
+  % arrow for asymptomatic D individual
+  \draw[->] (8,-6) -- (11,-6); % arrow for asymptomatic
+
+  % Labels above the C row
+  \node at (8.75,-7.5) {Infected};
+  \node at (10,-7.5) {Symptomatic};
+  \node at (11,-7.5) {Isolated};
+
+  % E individual
+  \fill[violet!30] (8.75,-8) circle (5pt);
+  \draw (8.75,-8) circle (5pt);
+  \node at (8.75,-8) {E};
+
+  \fill[violet!30] (10,-8) circle (5pt);
+  \draw (10,-8) circle (5pt);
+  \node at (10,-8) {E};
+
+  \fill[violet!30] (11,-8) circle (5pt);
+  \draw (11,-8) circle (5pt);
+  \node at (11,-8) {E};
+
+  % Horizontal line between E circles
+  \draw (8.75,-8) -- (10,-8);
+  \draw (10,-8) -- (11,-8);
+
+  % Vertical dashed lines
+  \draw[dashed] (4,0) -- (4,-2);   % A to B
+  \draw[dashed] (4.5,-2) -- (4.5,-4);  % B to C
+  \draw[dashed] (8,-4) -- (8,-6);  % C to D
+  \draw[dashed] (8.75,-4) -- (8.75,-8);  % C to E
+
+  \end{tikzpicture}
+}
+\caption{A schematic of the branching process model with isolation.}
+\label{fig:ringbp-model}
+\end{figure}
 
@@
 
-The onset-to-isolation delay distribution determines the time between a case developing symptoms and then becoming isolated.
+The ability to contain an outbreak by isolating infected individuals is highly contingent on the response time, or delay, between individuals becoming symptomatic, getting tested, the test results being returned and going into isolation.
+The probability that this series of steps will take an given amount of time is encapsulated in the onset-to-isolation delay distribution, i.e.
+onset-to-isolation distribution determines the time between a case developing symptoms and then becoming isolated.
+Here we define the onset-to-isolation as a response time as it is determined by the time delay between an individual responding to symptoms by getting tested and the turn around time for the tests to be analysed and the result returned.
+We view the onset-to-isolation as a feature of the public health outbreak response rather than an epidemiological characteristic of a disease.
+As such it is possible for pandemic response to adapt their response time to meet the needs of an outbreak.
+\\
+
 We assume that an isolated case cannot transmit, so isolation is assumed to be 100\% effective once enacted ($R_{isolated} = 0$).
-The onset-to-isolation delay is parameterised with two scenarios: a short, SARS-like, delay (Donnelly et al.) and a longer COVID-19-like delay (Li et al.) (Figure \ref{fig:onset-to-isolation}).
-Both onset-to-isolation delays are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:onset-to-isolation}).
+The onset-to-isolation delay is parameterised with three scenarios: a fast response, based on the SARS in Hong Kong (Donnelly et al.
+2003) and a slower response based on COVID-19 in Wuhan, China (Li et al.
+2020) (Figure \ref{fig:onset-to-isolation}).
+Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:onset-to-isolation}).
 These two empirically informed delay distributions provide a guide to possible response times in the case of an influenza outbreak and will enable us to evaluate if such delays are too long to contain a disease that can transmit rapidly such as influenza.
+We also include a third onset-to-isolation scenario based on the case where lateral flow rapid tests (LFTs) are provided to all individuals in the susceptible population and these people are freely able to self-test if they suspect they are symptomatic.
+We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test).
 \\
@@
 \includegraphics[width=\textwidth]{../plots/onset_to_isolation.png}
-\caption{Onset-to-isolation delay distribution for SARS (SARS-CoV-1) and
-  COVID-19 (SARS-CoV-2).}
+\caption{Three assumed onset-to-isolation delays used in the simuluation experiment.
+The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong (Donnelly et al.
+2003), the median onset-to-isolation time is 3.4 days (blue dashed line).
+The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China (Li et al., 2020), the median onset-to-isolatiojn is 8.1 days (red dashed line).
+The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days.}
 \label{fig:onset-to-isolation}

### 2025-04-30 inst/scripts/plot_onset_to_isolation.R
add LFT to onset-to-isolation plot

@@ -5,25 +5,26 @@ density_eval <- seq(0, 25, 0.1)
 
 onset_to_isolation <- data.table(
   x = density_eval,
-  SARS = dweibull(density_eval, shape = 1.651524, scale = 4.287786),
-  COVID = dweibull(density_eval, shape = 2.305172, scale = 9.483875)
+  Fast = dweibull(density_eval, shape = 1.651524, scale = 4.287786),
+  Slow = dweibull(density_eval, shape = 2.305172, scale = 9.483875),
+  LFT = dexp(x = density_eval, rate = 0.5)
 )
 
 onset_to_isolation <- melt(
   data = onset_to_isolation,
   id.vars = "x",
-  measure.vars = c("SARS", "COVID"),
+  measure.vars = c("Fast", "Slow", "LFT"),
   variable.name = "disease",
   value.name = "onset_to_isolation"
 )
 
-sars_median <- epiparameter::convert_params_to_summary_stats(
+fast_median <- epiparameter::convert_params_to_summary_stats(
   "weibull", shape = 1.651524, scale = 4.287786
 )$median
-covid_median <- epiparameter::convert_params_to_summary_stats(
+slow_median <- epiparameter::convert_params_to_summary_stats(
   "weibull", shape = 2.305172, scale = 9.483875
 )$median
-
+lft_median <- log(2) / 0.5
 
 onset_to_isolation_plot <- ggplot2::ggplot(data = onset_to_isolation) +
   ggplot2::geom_line(
@@ -44,16 +45,19 @@ onset_to_isolation_plot <- ggplot2::ggplot(data = onset_to_isolation) +
     alpha = 0.1
   ) +
   ggplot2::geom_vline(
-    mapping = ggplot2::aes(xintercept = sars_median), lty = 2, col = "#1A85FF"
+    mapping = ggplot2::aes(xintercept = fast_median), lty = 2, col = "#1A85FF"
   ) +
   ggplot2::geom_vline(
-    mapping = ggplot2::aes(xintercept = covid_median), lty = 2, col = "#D41159"
+    mapping = ggplot2::aes(xintercept = slow_median), lty = 2, col = "#D41159"
+  ) +
+  ggplot2::geom_vline(
+    mapping = ggplot2::aes(xintercept = lft_median), lty = 2, col = "#FFB000"
   ) +
[…]

### 2025-04-30 inst/scripts/run_analysis.R
add LFT onset-to-isolation to run_analysis script

@@ -18,7 +18,8 @@ scenarios <- data.table(
       delay = c("SARS", "Wuhan"),
       onset_to_isolation = c(
         \(x) stats::rweibull(n = x, shape = 1.651524, scale = 4.287786),
-        \(x) stats::rweibull(n = x, shape = 2.305172, scale = 9.483875)
+        \(x) stats::rweibull(n = x, shape = 2.305172, scale = 9.483875),
+        \(x) stats::rexp(n = x, rate = 0.5)
       )
     )),
     k_group = list(data.table(
@@ -64,7 +65,7 @@ scenarios <- merge(
 )
 scenarios[, scenario :=  1:.N]
 
-# Parameterise fixed paramters
+# Parameterise fixed parameters
 sim_with_params <- purrr::partial(
   ringbp::scenario_sim,
   cap_max_days = 365,

### 2025-05-01 inst/paper/main.tex
update intro and add references

@@
 
+Infectious disease outbreaks pose an everpresent public health threat.
+The COVID-19 pandemic caused over 7 million deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}.
+Less transmissible pathogens, such as Ebola, can cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}.
+Thus, timely and accurate response is critical early in an outbreak when the pandemic or epidemic potential is still uncertain \citep{kucharskiControllingMinorOutbreaks2024}.
+An effective response requires understanding the efficacy and feasibility of intervention strategies across pathogens with varying epidemiological characteristics \citep{fraserFactorsThatMake2004}.
+\\
+
 The likelihood of an outbreak being controlled, in other words going locally-extinct or -eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical.
-The primary epidemiological determinant for the spread of an infectious disease is the reproduction number, defined as the average number of secondary cases  per primary case.
+The primary epidemiological determinant for the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases  per primary case.
 The higher the reproduction number the less likely a disease, without intervention, is to cease transmission in a population.
@@
 Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}.
-Higher individual-level variation for a given reproduction leading to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading).
-In addition, the likelihood of containment is reduced (assuming $R > 1$) if there are multiple introductions into a populations \citep{kucharskiEarlyDynamicsTransmission2020}.
+Higher individual-level variation for a given reproduction leads to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading).
+In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{kucharskiEarlyDynamicsTransmission2020}.
 \\
@@
 A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs).
-Early in an outbreak -- especially in the case of a novel pathogen -- NPIs are often the first line of defence, as pharmaceutical options may not yet be available.
-These interventions vary in scope and specificity, from targeted measures such as contact tracing and isolation, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (ref).
+Early in an outbreak --- especially in the case of a novel pathogen --- NPIs are often the first line of defence, as pharmaceutical options may not yet be available.
+These interventions vary in scope and specificity, from targeted measures such as contact tracing and isolation, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020).
 The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005).
@@
 $<3$ days), swift and effective interventions are required to achieve containment.
+Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}.
 \\
@@
 Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic.
-It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined is needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
-Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016} (Whittaker et al., 2024) (ref).
-Isolation of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics (SARS ref; other refs).
+It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
+Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, whittakerQuantifyingImpactBroadly2024} (Keeling et al).
+Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}.
 However, it's important to remember that it's not a panacea for all outbreak scenarios.
-In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018} (ref on limited capacity of tracing).
+In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}.
 \\
 
-Outbreaks have been shown to be more easily controlled when a large proportion of onward transmission takes place after the onset of symptoms \citep{fraserFactorsThatMake2004}.
-
-- Pandemic flu has similar pre-symptomatic transmission is similar to COVID-19 (ref).
-- SI shorter for pandemic flu (ref). \\
-
-It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation become the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (ref).
-However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may not be the case that contact tracing and isolation is unable to control influenza.
+It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation because the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (Fraser et al., 2004; Klinkenberg et al., 2006).
+However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that contact tracing and isolation is able to control influenza.
+The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic.
+Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates.
+This is the case for H5N1 and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}.
 Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}.
@@
 
-The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic (ref).
-Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates.
-This is the case for H5N1 and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 (ref).
-\\
-
 In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model.
-In effect the ability to control an epidemic with contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021).
+In effect the ability to control an epidemic with contact tracing and isolation is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021).
 The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval (Ward et al., 2024).
-This study looks to define the speed and effectiveness of isolation given an epidemic scenario to inform outbreak response complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals (ref).
+This study looks to define the speed and effectiveness of isolation under different response delays across epidemic scenarios to inform outbreak response; quantifying the impact of reducing the delay to isolation on outbreak containment and the utility of rapid testing.
 
@@
 Therefore, the isolation of a case can reduce the number of secondary cases if isolation happens prior to any of the secondary infection times.
+Another way to iterpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).
 \\
@@
 H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}.
-H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009} (ref on history of H1N1pdm).
+H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}.
 \\
+
+- Pandemic flu has similar pre-symptomatic transmission is similar to COVID-19 (ref).
+- SI shorter for pandemic flu (ref). \\
 
@@
 By understanding the efficacy of contact trace and isolation for a given epidemiological characterstic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.
+\item Our branching process model used a simple forward contact-tracing mechanism that reduces the number of secondary contacts based on the timing of isolation of symptomatic cases.
+This contact tracing stragety is minimal cost and effort.
+In scenarios where outbreaks are uncontrolled a more comprehensive and costly contact tracing system where infectors and contacts of infectors (i.e.
+backward and forward contact tracing) are followed up, which has been shown to be more effective at outbreak containment (Klinkenberg et al., 2006).
+\item Contact tracing and isolation complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals \citep{haydenPerspectivesAntiviralUse2001}.
 \end{itemize}

### 2025-05-06 inst/paper/main.tex
add info on subcritical cluster size for cases limit
switch order of contact tracing and isolation

@@
 
-\title{Evaluating contact tracing and isolation to control pandemic potential influenza outbreaks}
+\title{Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks}
 \author{Joshua W. Lambert}
@@
 Early in an outbreak --- especially in the case of a novel pathogen --- NPIs are often the first line of defence, as pharmaceutical options may not yet be available.
-These interventions vary in scope and specificity, from targeted measures such as contact tracing and isolation, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020).
+These interventions vary in scope and specificity, from targeted measures such as isolation and contact tracing, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020).
 The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005).
@@
 It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
-Contact tracing and isolation is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, whittakerQuantifyingImpactBroadly2024} (Keeling et al).
+Isolation and contact tracing is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, whittakerQuantifyingImpactBroadly2024} (Keeling et al).
 Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}.
@@
 
-It has become widely accepted that influenza viruses cannot be controlled with contact tracing and isolation because the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (Fraser et al., 2004; Klinkenberg et al., 2006).
-However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that contact tracing and isolation is able to control influenza.
+It has become widely accepted that influenza viruses cannot be controlled with isolation and contact tracing because the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (Fraser et al., 2004; Klinkenberg et al., 2006).
+However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza.
 The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic.
@@
 In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model.
-In effect the ability to control an epidemic with contact tracing and isolation is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021).
+In effect the ability to control an epidemic with isolation and contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021).
 The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval (Ward et al., 2024).
@@
 
-To assess the efficacy of contact tracing and isolation to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025}.
+To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025}.
 The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility).
@@
 
-There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable with 12-16 weeks. We set this limit to 500 cases.
+There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable with 12-16 weeks.
+We set this limit to 500 cases.
+This number needs to be larger than the expected cluster size from an subcritical outbreak produce by chance (i.e.
+larger than $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number).
 
@@
 The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5.
-We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of contact tracing and isolation in the event that pathogen evolution (e.g.
+We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution (e.g.
 genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens.
@@
 $R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event).
-Therefore evaluating contact tracing and isolation requires simulating outbreaks that are likely to exponentially spread without control measures.
+Therefore evaluating isolation and contact tracing requires simulating outbreaks that are likely to exponentially spread without control measures.
 
@@
 
-To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of contact tracing and isolation we set up a simulation experiment.
+To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of isolation and contact tracing we set up a simulation experiment.
 We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmissin, and number of initial cases.
@@
 (Figure \ref{fig:prop-outbreak-control-R}).
-There is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of contact tracing and isolation for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}).
+There is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}).
 As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e.
 0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once control effectiveness exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}.
-Once the reproduction number is at 2.5 or 3.5, an outbreak is only controlled in at least 50\% of scenarios with contact tracing and isolation at greater than 60\% (Figure \ref{fig:prop-outbreak-control-R}).
+Once the reproduction number is at 2.5 or 3.5, an outbreak is only controlled in at least 50\% of scenarios with isolation and contact tracing at greater than 60\% (Figure \ref{fig:prop-outbreak-control-R}).
 \\
 
-The incubation period influences the efficacy of tracing and isolation.
+The incubation period influences the efficacy of isolating and tracing.
 This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e.
@@
 \includegraphics[width=\textwidth]{../plots/prop_outbreak_control_reproduction_number.png}
-\caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an assumed reproduction number ($R$) of either 1.1 or 1.5.
+\caption{The percentage of outbreaks controlled across values of isolation and contact tracing effectiveness, for outbreaks with an assumed reproduction number ($R$) of either 1.1 or 1.5.
 This is for a baselines scenario where the number of initial cases is 10, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
@@
 For an influenza pandemic, the number of initial infected individuals influences efficacy of control-by-isolation at all transmissibility scenarios evaluated ($R \in \{1.1, 1.5, 2.5, 3.5\}$, Figure \ref{fig:prop-outbreak-control-num-init-cases}).
-When $R = 1.1$ most outbreaks can be controlled without contact tracing and isolation, even with 40 initial infectors Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+When $R = 1.1$ most outbreaks can be controlled without isolation and contact tracing, even with 40 initial infectors Figure \ref{fig:prop-outbreak-control-num-init-cases}).
 When $R = 1.5$ the increase in the number of initial infections from 5 to 20 has a large decrease in outbreaks controlled at low levels of contact tracing, the same pattern when the number of initial infectors is twofold larger at 40, but these differences in outbreak control diminish as contact tracing increases towards 100\% (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
@@
 \includegraphics[width=\textwidth]{../plots/prop_outbreak_control_num_init_cases.png}
-\caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an initial number of seeding infections assumed to be either 5 or 10.
+\caption{The percentage of outbreaks controlled across values of isolation and contact tracing effectiveness, for outbreaks with an initial number of seeding infections assumed to be either 5 or 10.
 This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the onset-to-isolation delay is assumed to be SARS-like.}
@@
 \includegraphics[width=\textwidth]{../plots/prop_outbreak_control_onset_to_isolation.png}
-\caption{The percentage of outbreaks controlled across values of contact and isolation effectiveness, for outbreaks with an onset-to-isolation delay distribution that is either SARS-like, or COVID-19-like (from Wuhan).
+\caption{The percentage of outbreaks controlled across values of isolation and contact tracing effectiveness, for outbreaks with an onset-to-isolation delay distribution that is either SARS-like, or COVID-19-like (from Wuhan).
 This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the number of initial infections is 10.}
@@
 \begin{itemize}
-\item This study has looked at the feasibility and effectiveness of contact tracing and isolation for influenza subtypes that have cause epidemic outbreaks.
+\item This study has looked at the feasibility and effectiveness of isolation and contact tracing for influenza subtypes that have cause epidemic outbreaks.
 \item Outbreaks controlled at 50\% of contacts traced and isolated is over ??\%, highlighting the effectiveness of this intervention for infectious diseases are not highly transmissible between humans.
 \item Discuss the existing flu vaccines and antivirals that can be used an interventions in addition to NPIs
-\item In modelling the effectiveness of contact tracing and isolation we have assumed that there is no capacity limits on contact tracing, which does not hold in large outbreaks such as SARS-CoV-2 (ref).
+\item In modelling the effectiveness of isolation and contact tracing we have assumed that there is no capacity limits on contact tracing, which does not hold in large outbreaks such as SARS-CoV-2 (ref).
 We also ignore potential socio-political factors that would prevent health worker from visiting or contacting infected individuals or contacts of infected individuals (Dhillon and Kelly, 2018).
@@
 \item Digital tracing has capacity for large-scale contact tracing (pingdemic), also rapid tests are a complementary approach for limiting spread.
-\item In this study we model contact tracing and isolation as the sole intervention in containing an influenza outbreak.
+\item In this study we model isolation and contact tracing as the sole intervention in containing an influenza outbreak.
 This is to provide an assessment of its effectiveness for rapidly transmission and potentially highly transmissible subtypes.
@@
 \item This study used a branching process model to simulate individual-level transmission and intervention.
-The model assumes an infinite population size with no interdependence of contact tracing and isolation between cases whereby a case may be in contact with multiple infectors and only needs to be traced in time by one contact to prevent transmission, as could happened in a fixed random network transmission model.
+The model assumes an infinite population size with no interdependence of isolation and contact tracing between cases whereby a case may be in contact with multiple infectors and only needs to be traced in time by one contact to prevent transmission, as could happened in a fixed random network transmission model.
 It is unlikely that this model structure difference would have a considerable different on our results for the control of disease outbreaks with contact tracing.
@@
 If transmissibility of these avian influenza subtypes remains constant over time then outbreaks are expected to be subcritical and will not produce epidemics and thus not require interventions.
-If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence contact tracing and isolation while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence isolation and contact tracing while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
 \item This study has evaluated the ability to control the outbreak of an infectious disease independent on its severity.
 Infection severity and fatality risks are of course critical when formulating a pandemic response plan.
-By understanding the efficacy of contact trace and isolation for a given epidemiological characterstic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.
+By understanding the efficacy of isolation and contact trace for a given epidemiological characterstic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.
 \item Our branching process model used a simple forward contact-tracing mechanism that reduces the number of secondary contacts based on the timing of isolation of symptomatic cases.
@@
 backward and forward contact tracing) are followed up, which has been shown to be more effective at outbreak containment (Klinkenberg et al., 2006).
-\item Contact tracing and isolation complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals \citep{haydenPerspectivesAntiviralUse2001}.
+\item Isolation and contact tracing complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals \citep{haydenPerspectivesAntiviralUse2001}.
 \end{itemize}

### 2025-05-06 inst/paper/main.tex
add info on serial interval from skew-normal

@@
 The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility).
-The mean of the negative binomial offspring distribution is the basic reproduction number, and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}.
+The mean of the negative binomial offspring distribution is the basic reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}.
+\\
+
+There are three categories for individuals in the simulation, with each having their own parameterisation for the negative binomial offspring distribution: 1) \textit{community} (i.e.
+general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}.
+Each group have a distinct $R$ and dispersion ($k$) parameters due to the assumed differences in transmission dynamics between groups.
+\\
+
+The time between an infector being infected and then infecting an infectee (i.e.
+serial interval) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
+This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector.
+The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed.
+The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution.
+We select this value as to provides a realistic amount of variance when transforming onset time to exposure time (variance range 1.5 - 3.2).
+The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect.
+Overall, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we compute the time an infectee is exposed/infected.
 \\
@@
 Therefore, the isolation of a case can reduce the number of secondary cases if isolation happens prior to any of the secondary infection times.
-Another way to iterpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).
+Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).
 \\
@@
 
-We define the proportion of pre-symptomatic transmission follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skewed-normal distribution to sample the infection times using the incubation period and a skew parameter ($\theta$).
-The skew parameter controls the proportion of pre-symptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}).
-We parameterised the skew-normal distribution with a skew of 0.7, 1.95 and 30, these correspond to a pre-symptomatic transmission percentage of 30\%, 15\% and $<1\%$, respectively.
+We define the proportion of pre-symptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$).
+The shape parameter controls the proportion of pre-symptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}).
+We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:prop-presym-trans}).
 Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective contact tracing is hypothesied to be.
@@
 \includegraphics[width=\textwidth]{../plots/prop_presymptomatic_transmission.png}
-\caption{Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skewed-normal distribution.}
+\caption{Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skew-normal distribution.}
 \label{fig:prop-presym-trans}

### 2025-05-06 inst/paper/main.tex
add info to model description in methods and move model schematic figure

@@
 general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}.
-Each group have a distinct $R$ and dispersion ($k$) parameters due to the assumed differences in transmission dynamics between groups.
+Each group have a distinct $R$ and dispersion ($k$) parameter due to the assumed differences in transmission dynamics between groups.
 \\
@@
 The time between an infector being infected and then infecting an infectee (i.e.
-serial interval) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
+serial interval) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
 This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector.
@@
 The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect.
-Overall, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we compute the time an infectee is exposed/infected.
+Overall, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected.
 \\
 
-The NPI implemented in the transmission model is an isolation of symptomatic cases.
+The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing.
+When a case is symptomatic it is isolated [CHECK THIS].
 The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation).
-All infection times that are sampled to occur after the isolation of an infector are not infected.
-Therefore, the isolation of a case can reduce the number of secondary cases if isolation happens prior to any of the secondary infection times.
+Any infection times that are sampled to occur after the isolated of an infector are remove, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times.
+In other words, once a case is isolated at time $t_{isolated}$, the probability of that case having contact with anyone at $t > t_{isolated}$ is zero.
 Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).
+Symptomatic individuals that are traced are isolated either 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated.
+The isolation of the infectee is whatever occurs first of these two events.
 \\
 
-\subsection*{Defining outbreak control}
-
-We use the same definition for outbreak control as \cite{hellewellFeasibilityControllingCOVID192020}: no new cases between 12 and 16 weeks after the start of an outbreak.
-The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct.
+There are several mechanisms where isolation and contact tracing are imperfect at reducing onward disease transmission.
+The proportion of contacts missed by contact tracing is defined by the proportion of ascertainment and can range between zero, that is nobody is traced, to one, every case is traced.
+Symptomatic cases that have been missed when contact tracing their infectors are isolated a short delay after their symptom onset time.
+This delay is defined by the onset-to-isolation delay distribution.
+Additionally, contact tracing misses infections when infectees are infected by an asymptomatic infector.
+In this event the infectees are missed by contact tracing owing to the infector not being known to the response system.
 \\
-
-There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable with 12-16 weeks.
-We set this limit to 500 cases.
-This number needs to be larger than the expected cluster size from an subcritical outbreak produce by chance (i.e.
-larger than $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number).
-
-\subsection*{Pathogen epidemiological parameters} \label{epiparameters}
-
-In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previous caused sporadic human-to-human outbreaks with pandemic potential.
-We include A/H5N1, A/H1N1, and A/H7N9.
-These have caused outbreaks of various size and severity over the past 100 years.
-H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}.
-H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}.
-\\
-
-- Pandemic flu has similar pre-symptomatic transmission is similar to COVID-19 (ref).
-- SI shorter for pandemic flu (ref). \\
-
-We extracted incubation periods from the epidemiological literature that have been previously estimated for the chosen Influenza subtypes.
-All incubation periods were found to be well described by a Weibull distribution, \cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:incub}).
-\cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed.
-For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:incub}).
-\\
-
-The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5.
-We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution (e.g.
-genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens.
-Current estimates for H5N1 and H7N9 subtypes are below unity (i.e.
-$R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event).
-Therefore evaluating isolation and contact tracing requires simulating outbreaks that are likely to exponentially spread without control measures.
 
@@
 
+\subsection*{Defining outbreak control}
+
+We use the same definition for outbreak control as \cite{hellewellFeasibilityControllingCOVID192020}: no new cases between 12 and 16 weeks after the start of an outbreak.
+The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct.
+\\
+
+There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable with 12-16 weeks.
+We set this limit to 500 cases.
+This number needs to be larger than the expected cluster size from an subcritical outbreak produce by chance (i.e.
+larger than $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number).
+
+\subsection*{Pathogen epidemiological parameters} \label{epiparameters}
+
+In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previous caused sporadic human-to-human outbreaks with pandemic potential.
+We include A/H5N1, A/H1N1, and A/H7N9.
+These have caused outbreaks of various size and severity over the past 100 years.
+H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}.
+H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}.
+\\
+
+- Pandemic flu has similar pre-symptomatic transmission is similar to COVID-19 (ref).
+- SI shorter for pandemic flu (ref). \\
+
+We extracted incubation periods from the epidemiological literature that have been previously estimated for the chosen Influenza subtypes.
+All incubation periods were found to be well described by a Weibull distribution, \cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:incub}).
+\cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed.
+For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:incub}).
+\\
+
+The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5.
+We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution (e.g.
+genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens.
+Current estimates for H5N1 and H7N9 subtypes are below unity (i.e.
+$R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event).
+Therefore evaluating isolation and contact tracing requires simulating outbreaks that are likely to exponentially spread without control measures.
+
 \begin{figure}[ht]
@@
 
-The proportion of pandemic-potential influenza infections that are asymptomatic is assumes to be rare.
+The proportion of pandemic-potential influenza infections that are asymptomatic varies for past outbreaks.
 There is evidence from past outbreaks of H1N1 and ???
 that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009}.
-We ran scenarios with zero asymptomatic tranmission and another with 10\% of cases being asymptomatic.
+Here we assume that asymptomatic cases occur and can infect others, but are uncommon and that the majority of cases will develop symptoms during their infectious period.
+We ran three scenarios, one with zero asymptomatic tranmission, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.
 \\
@@
 \\
+
+The proportion of ascertainment is varied between 0 and 1, at intervals of 0.2, to test at what proportion of missed cases would lead to outbreaks becoming uncontrolled epidemics. \\
 

### 2025-05-12 inst/paper/main.tex
minor edits to intro and methods

@@
 Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}.
-Higher individual-level variation for a given reproduction leads to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading).
+Higher individual-level variation for a given reproduction number leads to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading).
 In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{kucharskiEarlyDynamicsTransmission2020}.
@@
 Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic.
-It can involve forward contact tracing, whereby infected individuals are asked who they have in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
-Isolation and contact tracing is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, whittakerQuantifyingImpactBroadly2024} (Keeling et al).
+It can involve forward contact tracing, whereby infected individuals are asked who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
+Isolation and contact tracing is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020, whittakerQuantifyingImpactBroadly2024}.
 Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}.
@@
 
-To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025}.
-The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (without depletion of susceptibility).
+To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}).
+The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (infinite population size, without depletion of susceptibility).
 The mean of the negative binomial offspring distribution is the basic reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}.
@@
 The time between an infector being infected and then infecting an infectee (i.e.
-serial interval) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
+generation time) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
 This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector.
@@
 The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution.
-We select this value as to provides a realistic amount of variance when transforming onset time to exposure time (variance range 1.5 - 3.2).
-The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect.
-Overall, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected.
+We select this value as to provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 - 3.2).
+The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution).
+In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector.
 \\
 
-The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing.
-When a case is symptomatic it is isolated [CHECK THIS].
+The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing (Figure \ref{fig:ringbp-model}).
+When a case is symptomatic they are isolated after a delay.
 The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation).
 Any infection times that are sampled to occur after the isolated of an infector are remove, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times.
-In other words, once a case is isolated at time $t_{isolated}$, the probability of that case having contact with anyone at $t > t_{isolated}$ is zero.
+In other words, once a case is isolated at time $t_{isolated}$, the probability of that case infecting anyone at $t > t_{isolated}$ is zero.
 Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).
-Symptomatic individuals that are traced are isolated either 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated.
+Symptomatic individuals that are traced are isolated either when 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated.
 The isolation of the infectee is whatever occurs first of these two events.
@@
 The proportion of contacts missed by contact tracing is defined by the proportion of ascertainment and can range between zero, that is nobody is traced, to one, every case is traced.
-Symptomatic cases that have been missed when contact tracing their infectors are isolated a short delay after their symptom onset time.
-This delay is defined by the onset-to-isolation delay distribution.
-Additionally, contact tracing misses infections when infectees are infected by an asymptomatic infector.
+Symptomatic cases that have been missed when contact tracing their infectors are isolated at the onset-to-isolation delay after their symptom onset time.
+Additionally, contact tracing misses infections by an asymptomatic infector.
 In this event the infectees are missed by contact tracing owing to the infector not being known to the response system.
@@
 
-- Pandemic flu has similar pre-symptomatic transmission is similar to COVID-19 (ref).
-- SI shorter for pandemic flu (ref). \\
-
 We extracted incubation periods from the epidemiological literature that have been previously estimated for the chosen Influenza subtypes.
@@
 Current estimates for H5N1 and H7N9 subtypes are below unity (i.e.
-$R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention (except in a extremely rare case of superspreading event).
-Therefore evaluating isolation and contact tracing requires simulating outbreaks that are likely to exponentially spread without control measures.
+$R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention; with most outbreaks being small except in rare instances of superspreading events.
+Therefore evaluating isolation and contact tracing requires simulating outbreaks that have pandemic potential with exponential spread in the absence of control measures.
 

### 2025-05-12 inst/paper/main.tex
update results on onset-to-isolation scenarios

@@
 We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test).
+We assume that LFTs are 100\% accurate so all symptomatic individuals will receive a positive test result.
 \\
@@
 
-The incubation period for H1N1 is noticable shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:incub}). \\
+The incubation period for H1N1 is noticably shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:incub}). \\
 
-Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number.
-A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of control effectiveness.
-(Figure \ref{fig:prop-outbreak-control-R}).
-There is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}).
+Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number and the speed of isolation after symptom onset (Figure \ref{fig:prop-outbreak-control-R}).
+A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of contacts traced, except in cases where they are equal because zero or all outbreaks are contained, across all onset-to-isolation scenarios (Figure \ref{fig:prop-outbreak-control-R}).
+When the reponse to isolate cases is slow then outbreaks with $R = 1.5$ require at least 50\% of contacts traced to contain at least half of simulated outbreaks.
+When $R$ is 2.5 or 3.5 then contact tracing needs to be extremely comprehensive ($>90\%$ to have a chance at containment, and even then $R = 3.5$ influenza outbreaks on average are uncontrolled (Figure \ref{fig:prop-outbreak-control-R}A).
+Especially for H1N1, given it's faster tempo of transmission, having a much lower probability of containment at 100\% of contacts traced with a slow, COVID-19-like, isolation response time (Figure \ref{fig:prop-outbreak-control-R}A).
+In a scenario with a fast, Hong Kong SARS-like, onset-to-isolation response, outbreaks with $R = 1.1$ are almost always contained at any proportion of contacts traced (Figure \ref{fig:prop-outbreak-control-R}B).
+For slow and fast isolation response delays there is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}A,B).
 As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e.
-0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once control effectiveness exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}.
-Once the reproduction number is at 2.5 or 3.5, an outbreak is only controlled in at least 50\% of scenarios with isolation and contact tracing at greater than 60\% (Figure \ref{fig:prop-outbreak-control-R}).
+0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once percentage of contacts ascertained in tracing exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}B).
+For more transmissible influenza scenarios, even a fast isolation response with an average of 3.4 days between symptom onset and isolation requires $>70\%$ of contacts traced for most outbreaks to be controlled (Figure \ref{fig:prop-outbreak-control-R}B).
 \\
 
-The incubation period influences the efficacy of isolating and tracing.
+The most rapid scenario we simulated to emulate individuals self-testing with LFTs, has the same trend of increased chance of outbreak containment by reducing the time between symptom onset and isolation, as the slow and fast scenarios (Figure \ref{fig:prop-outbreak-control-R}).
+The contol of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C).
+Assuming ubiquitous access to home testing and that 63\% of symptomatic cases testing and isolating within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$).
+The LFT isolation strategy is also the only onset-to-isolation delay which can achieve 100\% of outbreaks controlled when all contacts are traced (Figure \ref{fig:prop-outbreak-control-R}C).
+\\
+
+The difference in influenza subtype incubation period influences the efficacy of isolating and tracing.
 This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e.
 not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}).
-The shortest incubation period, H1N1, predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection limit our ability to effectively tracing and isolate their potential infectees in time to prevent disease transmission.
-The difficulty to control an pandemic outbreak with a short incubation period is shown by the $\sim 40\%$ of simulated H1N1 outbreaks with 100\% control effectiveness that were not contained when $R = 3.5$, while longer incubation period outbreaks have $>80\%$ controlled when 100\% of contacts are traced and isolated (Figure \ref{fig:prop-outbreak-control-R}).
-Overall, both the transmissibility and incubation period impact the feasibility to control an outbreak.
-\\
-
+The shortest incubation period, H1N1, predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection and in some instances transmission while subclinical limit our ability to effectively tracing and isolate their potential infectees in time to prevent disease transmission.
 The variance between the influenza subtypes at each percentage of contacts traced and isolated varies (i.e.

### 2025-05-12 inst/scripts/plot_patho_param_space.R
update scenarios and influenza in patho_param_space plot

@@ -125,22 +125,22 @@ ggplot2::ggplot(param_space, ggplot2::aes(x = prop_asym, y = R, color = pathogen
 # deterministic plot
 patho_param_space_plot <- ggplot2::ggplot() +
     ggplot2::annotate(
-      geom = "rect", xmin = 0, xmax = 35, ymin = 1.1, ymax = 3.5,
+      geom = "rect", xmin = 0, xmax = 50, ymin = 1.1, ymax = 3.5,
       fill = "coral", alpha = 0.5
   ) +
   ggplot2::annotate(
-    geom = "text", x = 17.5, y = 2.3, label = "H1N1/H5N1/H7N9\n Scenarios",
+    geom = "text", x = 25, y = 2.3, label = "H1N1/H5N1/H7N9\n Scenarios",
     size = 3
   ) +
   ggfx::with_blur(
     ggplot2::annotate(
-      geom = "rect", xmin = 35, xmax = 45, ymin = 1, ymax = 10,
+      geom = "rect", xmin = 15, xmax = 45, ymin = 1, ymax = 5,
       fill = "darkgreen", alpha = 0.5
     ),
     sigma = 7.5
   ) +
   ggplot2::annotate(
-    geom = "text", x = 40, y = 4.5, label = "Influenza", size = 3
+    geom = "text", x = 40, y = 3, label = "Influenza", size = 3
   ) +
   ggfx::with_blur(
     ggplot2::annotate(

### 2025-05-12 inst/scripts/plot_prop_outbreak_control_reproduction_number.R
update prop_outbreak_control_reproduction_number plot

@@ -3,25 +3,48 @@ library(ggplot2)
 library(dplyr)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
-sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
+h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
 
-res <- sweep_results %>%
-  group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 500)) %>%
-  ungroup(scenario)
+h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
 
-dt <- as.data.table(res)
-dt_data <- rbindlist(dt$data)
-dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
+h5n1_data <- rbindlist(h5n1_results$data)
+h5n1_data[, `:=`(scenario = h5n1_results$scenario, pext = h5n1_results$pext)]
 
-prop_outbreak_control <- dt_data[
-  num.initial.cases == 20 & theta == "15%" & delay == "SARS" & prop.asym == 0,
-  .(control_effectiveness, index_R0, pext, subtype)
+h1n1_data <- rbindlist(h1n1_results$data)
+h1n1_data[, `:=`(scenario = h1n1_results$scenario, pext = h1n1_results$pext)]
+
+h7n9_data <- rbindlist(h7n9_results$data)
+h7n9_data[, `:=`(scenario = h7n9_results$scenario, pext = h7n9_results$pext)]
+
+rm(h5n1_results)
+rm(h1n1_results)
+rm(h7n9_results)
+
+flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
+
+prop_outbreak_control <- flu_data[
+  num.initial.cases == 20 & theta == "15%" & prop.asym == 0,
+  .(control_effectiveness, index_R0, pext, subtype, delay)
 ]
 
 # convert to percentages for plotting
[…]

### 2025-05-12 inst/scripts/run_analysis.R
update run_analysis script to run lft and high asymptomatic scenarios, and save pathogen output separately

@@ -15,7 +15,7 @@ h7n9_weibull_params <- epiparameter::convert_summary_stats_to_params(
 scenarios <- data.table(
   expand.grid(
     delay_group = list(data.table(
-      delay = c("SARS", "Wuhan"),
+      delay = c("fast", "slow", "lft"),
       onset_to_isolation = c(
         \(x) stats::rweibull(n = x, shape = 1.651524, scale = 4.287786),
         \(x) stats::rweibull(n = x, shape = 2.305172, scale = 9.483875),
@@ -43,7 +43,7 @@ scenarios <- data.table(
       )
     )),
     index_R0 = c(1.1, 1.5, 2.5, 3.5),
-    prop.asym = c(0, 0.1),
+    prop.asym = c(0, 0.1, 0.3),
     control_effectiveness = seq(0, 1, 0.2),
     num.initial.cases = c(5, 20, 40),
     quarantine = FALSE
@@ -81,10 +81,26 @@ sim_with_params <- purrr::partial(
 future::plan("multisession", workers = 8)
 
 # Run parameter sweep
-sweep_results <- ringbp::parameter_sweep(
-  scenarios,
+h5n1_results <- ringbp::parameter_sweep(
+  scenarios[subtype == "H5N1"],
   sim_fn = sim_with_params,
   samples = 100
 )
 
-saveRDS(sweep_results, file = file.path("inst", "extdata", "simulations.rds"))
+saveRDS(h5n1_results, file = file.path("inst", "extdata", "h5n1_simulations.rds"))
+
+h1n1_results <- ringbp::parameter_sweep(
+  scenarios[subtype == "H1N1"],
+  sim_fn = sim_with_params,
+  samples = 100
+)
+
+saveRDS(h1n1_results, file = file.path("inst", "extdata", "h1n1_simulations.rds"))
+
+h7n9_results <- ringbp::parameter_sweep(
+  scenarios[subtype == "H7N9"],
+  sim_fn = sim_with_params,
+  samples = 100
+)
+
+saveRDS(h7n9_results, file = file.path("inst", "extdata", "h7n9_simulations.rds"))

### 2025-05-13 inst/scripts/plot_prop_outbreak_control_prop_asym_iso.R
add prop_outbreak_control_prop_asym_iso script and plot

@@ -1,0 +1,116 @@
+library(data.table)
+library(ggplot2)
+library(dplyr)
+
+# read simulation results saved by running inst/scripts/run_analysis.R
+h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+
+h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+
+h5n1_data <- rbindlist(h5n1_results$data)
+h5n1_data[, `:=`(scenario = h5n1_results$scenario, pext = h5n1_results$pext)]
+
+h1n1_data <- rbindlist(h1n1_results$data)
+h1n1_data[, `:=`(scenario = h1n1_results$scenario, pext = h1n1_results$pext)]
+
+h7n9_data <- rbindlist(h7n9_results$data)
+h7n9_data[, `:=`(scenario = h7n9_results$scenario, pext = h7n9_results$pext)]
+
+rm(h5n1_results)
+rm(h1n1_results)
+rm(h7n9_results)
+
+flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
+
+prop_outbreak_control <- flu_data[
+  num.initial.cases == 20 & theta == "15%",
+  .(control_effectiveness, index_R0, pext, subtype, delay, prop.asym)
+]
+
+# convert to percentages for plotting
+prop_outbreak_control[, control_effectiveness := control_effectiveness * 100]
+prop_outbreak_control[, pext := pext * 100]
+
+# factorise and order delay type for plotting order
+prop_outbreak_control[, delay := as.factor(delay)]
+
+prop_outbreak_control[, delay := factor(delay, levels = c("slow", "fast", "lft"))]
+
+delay_labels <- c(
+  "slow" = "Slow",
+  "fast" = "Fast",
+  "lft" = "LFT",
+  "0" = "0% asymptomatic",
+  "0.1" = "10% asymptomatic",
+  "0.3" = "30% asymptomatic"
+)
+
[…]

### 2025-05-13 inst/paper/main.tex
update asymptomatic methods and results

@@
 It has become widely accepted that influenza viruses cannot be controlled with isolation and contact tracing because the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (Fraser et al., 2004; Klinkenberg et al., 2006).
-However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers (Garg et al., 2025) and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza.
+However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza.
 The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic.
@@
 In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model.
-In effect the ability to control an epidemic with isolation and contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent (Reyna-Lara et al., 2021).
+In effect the ability to control an epidemic with isolation and contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent \citep{reyna-laraVirusSpreadContact2021}.
 The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval (Ward et al., 2024).
@@
 The proportion of pandemic-potential influenza infections that are asymptomatic varies for past outbreaks.
-There is evidence from past outbreaks of H1N1 and ???
-that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009}.
+There is evidence from past human outbreaks of H1N1pdm and H5N1 (2.3.4.4b) that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009, gargHighlyPathogenicAvian2025}; but whether human-to-human transmission can occur while subclinical is unknown.
+There is no evidence of asymptomatic transmission for the H7N9 subtype \citep{xuSerologicalInvestigationSubclinical2013}.
 Here we assume that asymptomatic cases occur and can infect others, but are uncommon and that the majority of cases will develop symptoms during their infectious period.
@@
 \clearpage
+
+The pattern of outbreak control is broadly similar between 10\% asymptomatic cases and the 0\% and 30\% asymptomatic scenarios (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
+Qualitative difference are in the 30\% asymptomatic scenario for LFT isolation, where the ability to control highly transmissible influenzas reduces to less than a quarter of outbreaks controlled at maxmimum contact tracing coverage (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
+When every case is symptomatic the likelihood of containing highly transmissible outbreaks drastically increases for all onset-to-isolation response delays (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
+\\
 
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_prop_asym_iso.png}
+\caption{Stub.}
+\label{fig:prop-outbreak-control-prop-asym-iso}
+\end{figure}
+
 \end{document}

### 2025-05-13 inst/scripts/plot_prop_outbreak_control_prop_presym_iso.R
added prop_outbreak_control_prop_presym_iso script and plot

@@ -1,0 +1,116 @@
+library(data.table)
+library(ggplot2)
+library(dplyr)
+
+# read simulation results saved by running inst/scripts/run_analysis.R
+h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+
+h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+
+h5n1_data <- rbindlist(h5n1_results$data)
+h5n1_data[, `:=`(scenario = h5n1_results$scenario, pext = h5n1_results$pext)]
+
+h1n1_data <- rbindlist(h1n1_results$data)
+h1n1_data[, `:=`(scenario = h1n1_results$scenario, pext = h1n1_results$pext)]
+
+h7n9_data <- rbindlist(h7n9_results$data)
+h7n9_data[, `:=`(scenario = h7n9_results$scenario, pext = h7n9_results$pext)]
+
+rm(h5n1_results)
+rm(h1n1_results)
+rm(h7n9_results)
+
+flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
+
+prop_outbreak_control <- flu_data[
+  num.initial.cases == 20 & prop.asym == 0.1,
+  .(control_effectiveness, index_R0, pext, subtype, delay, theta)
+]
+
+# convert to percentages for plotting
+prop_outbreak_control[, control_effectiveness := control_effectiveness * 100]
+prop_outbreak_control[, pext := pext * 100]
+
+# factorise and order delay type for plotting order
+prop_outbreak_control[, delay := as.factor(delay)]
+
+prop_outbreak_control[, delay := factor(delay, levels = c("slow", "fast", "lft"))]
+
+delay_labels <- c(
+  "slow" = "Slow",
+  "fast" = "Fast",
+  "lft" = "LFT",
+  "<1%" = "<1% presymptomatic",
+  "15%" = "15% presymptomatic",
+  "30%" = "30% presymptomatic"
+)
+
[…]

### 2025-05-13 inst/paper/main.tex
add presymptomatic results

@@
 
+The proportion of transmission that occurs before on the onset of symptoms has a moderate affect on epidemic containment for across isolation response rates (Figure \ref{fig:prop-outbreak-control-prop-presym-iso}).
+In all scenarios decreasing the proportion of presympatomatic transmission result in an greater or equal probability of outbreak control.
+When presymptomatic transmission is minimal, most $R = 1.1$ outbreaks go extinct without any contact tracing, at any rate of isolation after symptom onset, with extinction ensured if isolated after LFT testing (Figure \ref{fig:prop-outbreak-control-prop-presym-iso} leftmost column).
+Furthermore, when  $<1\%$ of transmission is presymptomatic, rapidly isolating after symptoms (e.g.
+after self-testing) can contain more than half of outbreaks for all transmissibility scenarios ($R \leq 3.5$) (Figure \ref{fig:prop-outbreak-control-prop-presym-iso}).
+Conversely, when 30\% of transmission is presymptomatic, less than half of highly transmissible epidemics are controlled, even with the most optimistic isolation response time and contact tracing coverage (Figure \ref{fig:prop-outbreak-control-prop-presym-iso} rightmost column).
+\\
+
 The ability to control an epidemic that can readily transmit human-to-human is moderated by the initial number of infectious individuals that seed the outbreak.
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_prop_presym_iso.png}
+\caption{Stub.}
+\label{fig:prop-outbreak-control-prop-presym-iso}
+\end{figure}
+
 \end{document}

### 2025-05-14 inst/paper/main.tex
minor edits to intro to include isolation and cite references correctly

@@
 For highly transmissible pathogens (e.g.
-$R>3$) with short generation times (e.g.
-$<3$ days), swift and effective interventions are required to achieve containment.
+$R \geq 2.5$) with short generation times (e.g.
+$\leq 3$ days), swift and effective interventions are required to achieve containment.
 Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}.
@@
 
-Contact tracing, the process of finding the contacts of infected individuals is an often used NPI to try and curtail the growth of an epidemic.
-It can involve forward contact tracing, whereby infected individuals are asked who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
+Isolation of infected individuals aims to break the chain of transmission by preventing further infection of susceptible individuals.
+The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to infectiousness optimal.
+This is challenging for most pathogens because infectiousness preceeds prodromal symptoms.
+While contact tracing is the process of finding the contacts of infected individuals.
+Both isolation and contact tracing are often used NPIs to try and curtail the growth of an epidemic.
+Contact tracing can be forward, whereby infected individuals are asked who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
 Isolation and contact tracing is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020, whittakerQuantifyingImpactBroadly2024}.
@@
 
-It has become widely accepted that influenza viruses cannot be controlled with isolation and contact tracing because the rate at which infected individuals become infectious and show symptoms is faster than the latency in the process of contacting contacts (Fraser et al., 2004; Klinkenberg et al., 2006).
-However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval (SI) from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza.
+It has become widely accepted that influenza viruses cannot be controlled with isolation and contact tracing because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
+However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza.
 The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic.
 Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates.
-This is the case for H5N1 and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}.
-Historically, avian influenza (H5N1 and H7N9) have had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}.
+This is the case for H5N1 \citep{Ward2024.12.11.24318702} and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}.
+Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}.
 \\

### 2025-05-14 inst/scripts/plot_onset_to_isolation.R
update legend order in onset_to_isolation plot

@@ -14,7 +14,7 @@ onset_to_isolation <- melt(
   data = onset_to_isolation,
   id.vars = "x",
   measure.vars = c("Fast", "Slow", "LFT"),
-  variable.name = "disease",
+  variable.name = "response",
   value.name = "onset_to_isolation"
 )
 
@@ -26,12 +26,15 @@ slow_median <- epiparameter::convert_params_to_summary_stats(
 )$median
 lft_median <- log(2) / 0.5
 
+# order response factors for plotting order
+onset_to_isolation[, response := factor(response, levels = c("Slow", "Fast", "LFT"))]
+
 onset_to_isolation_plot <- ggplot2::ggplot(data = onset_to_isolation) +
   ggplot2::geom_line(
     mapping = ggplot2::aes(
       x = x,
       y = onset_to_isolation,
-      col = disease
+      col = response
     )
   ) +
   ggplot2::geom_ribbon(
@@ -40,15 +43,15 @@ onset_to_isolation_plot <- ggplot2::ggplot(data = onset_to_isolation) +
       y = onset_to_isolation,
       ymin = 0,
       ymax = onset_to_isolation,
-      fill = disease
+      fill = response
     ),
     alpha = 0.1
   ) +
   ggplot2::geom_vline(
-    mapping = ggplot2::aes(xintercept = fast_median), lty = 2, col = "#1A85FF"
+    mapping = ggplot2::aes(xintercept = slow_median), lty = 2, col = "#1A85FF"
   ) +
   ggplot2::geom_vline(
-    mapping = ggplot2::aes(xintercept = slow_median), lty = 2, col = "#D41159"
+    mapping = ggplot2::aes(xintercept = fast_median), lty = 2, col = "#D41159"
   ) +
   ggplot2::geom_vline(
     mapping = ggplot2::aes(xintercept = lft_median), lty = 2, col = "#FFB000"

### 2025-05-14 inst/scripts/plot_prop_outbreak_control_num_init_cases.R
update prop_outbreak_control_num_init_cases plot and script

@@ -1,30 +1,41 @@
 library(ggplot2)
 library(data.table)
-library(dplyr)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
-sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
+h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
 
-res <- sweep_results %>%
-  group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 500)) %>%
-  ungroup(scenario)
+h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
 
-dt <- as.data.table(res)
-dt_data <- rbindlist(dt$data)
-dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
+h5n1_data <- rbindlist(h5n1_results$data)
+h5n1_data[, `:=`(scenario = h5n1_results$scenario, pext = h5n1_results$pext)]
 
-prop_outbreak_control_num_init_cases <- dt_data[
-  theta == "15%" & delay == "SARS" & prop.asym == 0,
-  .(index_R0, control_effectiveness, num.initial.cases, pext, subtype)
+h1n1_data <- rbindlist(h1n1_results$data)
+h1n1_data[, `:=`(scenario = h1n1_results$scenario, pext = h1n1_results$pext)]
+
+h7n9_data <- rbindlist(h7n9_results$data)
+h7n9_data[, `:=`(scenario = h7n9_results$scenario, pext = h7n9_results$pext)]
+
+rm(h5n1_results)
+rm(h1n1_results)
+rm(h7n9_results)
+
+flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
+
+prop_outbreak_control <- flu_data[
+  theta == "15%" & prop.asym == 0 & delay == "fast",
+  .(control_effectiveness, index_R0, num.initial.cases, pext, subtype)
 ]
 
[…]

### 2025-05-14 inst/scripts/plot_prop_outbreak_control_reproduction_number.R
remove library(dplyr) from plot_prop_outbreak_control_reproduction_number.R

@@ -1,6 +1,5 @@
 library(data.table)
 library(ggplot2)
-library(dplyr)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
 h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))

### 2025-05-14 inst/scripts/plot_prop_outbreak_control_onset_to_isolation.R
remove prop_outbreak_control_onset_to_isolation plot and script

@@ -1,80 +1,0 @@
-library(ggplot2)
-library(data.table)
-library(dplyr)
-
-# read simulation results saved by running inst/scripts/run_analysis.R
-sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
-
-res <- sweep_results %>%
-  group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 500)) %>%
-  ungroup(scenario)
-
-dt <- as.data.table(res)
-dt_data <- rbindlist(dt$data)
-dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
-
-prop_outbreak_control_onset_to_isolation <- dt_data[
-  num.initial.cases == 20 & theta == "15%" & index_R0 == 1.5 & prop.asym == 0,
-  .(control_effectiveness, onset_to_isolation, delay, pext, subtype)
-]
-
-# convert to percentages for plotting
-prop_outbreak_control_onset_to_isolation[, control_effectiveness := control_effectiveness * 100]
-prop_outbreak_control_onset_to_isolation[, pext := pext * 100]
-
-prop_outbreak_control_onset_to_isolation_plot <- ggplot2::ggplot(
-  data = prop_outbreak_control_onset_to_isolation
-) +
-  ggplot2::geom_point(
-    mapping = ggplot2::aes(
-      x = control_effectiveness,
-      y = pext,
-      fill = as.factor(delay),
-      shape = as.factor(subtype)
-    ),
-    size = 3
-  ) +
-  ggplot2::geom_line(
-    mapping = ggplot2::aes(
-      x = control_effectiveness,
-      y = pext,
-      colour = as.factor(delay),
-      linetype = as.factor(subtype)
-    ),
-    linewidth = 0.75
-  ) +
-  ggplot2::scale_x_continuous(
-    name = "Contacts traced (%)",
-    limits = c(0, 100)
-  ) +
-  ggplot2::scale_y_continuous(
-    name = "Simulated outbreaks controlled (%)",
-    limits = c(0, 100)
-  ) +
-  ggplot2::scale_colour_manual(values = c("#92C5DE", "#F4A582")) +
-  ggplot2::scale_fill_manual(values = c("#92C5DE", "#F4A582")) +
-  ggplot2::scale_shape_manual(values = c(21, 22, 24)) +
[…]

### 2025-05-14 inst/paper/main.tex
update methods with H7N9 info, tidy end of methods, correctly cite all references in methods, fix wording in results and reorder presymptomatic and asymptomatic results

@@
 The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution.
-We select this value as to provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 - 3.2).
+We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2).
 The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution).
@@
 
-There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable with 12-16 weeks.
+There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable within 12-16 weeks.
 We set this limit to 500 cases.
@@
 
-In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previous caused sporadic human-to-human outbreaks with pandemic potential.
+In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previously caused sporadic human-to-human outbreaks with pandemic potential.
 We include A/H5N1, A/H1N1, and A/H7N9.
@@
 H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}.
+The first human case with the H7N9 subtype was in 2013, subsequently causing five outbreak waves, but no human cases since 2019 \citep{liH7N9InfluenzaVirus2021}.
+Most cases are of zoonotic origin with recorded animal exposure but there have been some small clusters (2 --- 3 cases) of possible human-to-human transmission \citep{liEpidemiologyHumanInfections2014}.
+For the purposes of this study we assume that there is no preexisting population immunity for these influenza subtypes.
+This is generally true for H5N1 and H7N9 at present, and for H1N1pdm09 prior to 2009.
 \\
@@
 To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of isolation and contact tracing we set up a simulation experiment.
-We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmissin, and number of initial cases.
+We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmission, and number of initial cases.
 \\
@@
 \caption{Three assumed onset-to-isolation delays used in the simuluation experiment.
-The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong (Donnelly et al.
-2003), the median onset-to-isolation time is 3.4 days (blue dashed line).
-The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China (Li et al., 2020), the median onset-to-isolatiojn is 8.1 days (red dashed line).
-The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days.}
+The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China \citep{liEarlyTransmissionDynamics2020}, the median onset-to-isolatiojn is 8.1 days (blue dashed line).
+The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a}, the median onset-to-isolation time is 3.4 days (red dashed line).
+The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days (orange dashed line).}
 \label{fig:onset-to-isolation}
@@
 
-We define the proportion of pre-symptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$).
-The shape parameter controls the proportion of pre-symptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}).
+We define the proportion of presymptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$).
+The shape parameter controls the proportion of presymptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}).
 We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:prop-presym-trans}).
-Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective contact tracing is hypothesied to be.
+Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective isolation and contact tracing is hypothesied to be.
 \\
@@
 Here we assume that asymptomatic cases occur and can infect others, but are uncommon and that the majority of cases will develop symptoms during their infectious period.
-We ran three scenarios, one with zero asymptomatic tranmission, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.
+We ran three scenarios, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.
 \\
@@
 
-We simulated scenarios with either 5, 20 or 40 initial cases.
-This is to test whether contact tracing can contain outbreaks that may have already accumulated cases before contact tracing is started, or represents importation of multiple cases in a single event, such as a flight with multiple infectors (Khanh et al.
-2020).
+The feasibility of epidemic containment by contact tracing is expected to depend on the proportion of close contacts of cases that are traced and isolated if symptomatic.
+We varied the proportion of contacts ascertained in the model between 0 and 1, at intervals of 0.2.
+In cases when the proportion of contacts traced is zero, isolation after the delay from symptom onset is the only NPI reducing disease transmission.
+To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial indepedent cases.
+This could represent importation of multiple cases in a single event, such as a flight with multiple positive cases.
 \\
 
-The proportion of ascertainment is varied between 0 and 1, at intervals of 0.2, to test at what proportion of missed cases would lead to outbreaks becoming uncontrolled epidemics. \\
-
-Some parameters we fixed across all simulations.
-The dispersion parameter for the negative binomial offspring distribution was fixed at 1 for isolated cases, 0.16 for subclinical and community cases.
+We fixed some parameters across all simulations.
+The dispersion parameter for the negative binomial offspring distribution was fixed at 1 for isolated cases, and 0.16 for subclinical and community cases.
 For computational efficiency we capped the maximum number of days the simulation could run for to 365.
@@
 This is an arbitrary threshold we judge to be large epidemic that is not controllable.
-\\
 
@@
 When the reponse to isolate cases is slow then outbreaks with $R = 1.5$ require at least 50\% of contacts traced to contain at least half of simulated outbreaks.
-When $R$ is 2.5 or 3.5 then contact tracing needs to be extremely comprehensive ($>90\%$ to have a chance at containment, and even then $R = 3.5$ influenza outbreaks on average are uncontrolled (Figure \ref{fig:prop-outbreak-control-R}A).
-Especially for H1N1, given it's faster tempo of transmission, having a much lower probability of containment at 100\% of contacts traced with a slow, COVID-19-like, isolation response time (Figure \ref{fig:prop-outbreak-control-R}A).
+When $R$ is 2.5 or 3.5 then contact tracing needs to be extremely comprehensive ($>90\%$) to have a chance at containment, and even then $R = 3.5$ influenza outbreaks on average are uncontrolled (Figure \ref{fig:prop-outbreak-control-R}A).
+Especially for H1N1, given its faster tempo of transmission, has a much lower probability of containment at 100\% of contacts traced with a slow, COVID-19-like, isolation response time (Figure \ref{fig:prop-outbreak-control-R}A).
 In a scenario with a fast, Hong Kong SARS-like, onset-to-isolation response, outbreaks with $R = 1.1$ are almost always contained at any proportion of contacts traced (Figure \ref{fig:prop-outbreak-control-R}B).
@@
 As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e.
-0\% control effectiveness), however, outbreaks with $R = 1.5$ are mostly contained once percentage of contacts ascertained in tracing exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}B).
-For more transmissible influenza scenarios, even a fast isolation response with an average of 3.4 days between symptom onset and isolation requires $>70\%$ of contacts traced for most outbreaks to be controlled (Figure \ref{fig:prop-outbreak-control-R}B).
+0\% contact ascertainment), however, outbreaks with $R = 1.5$ are mostly contained once the percentage of contacts ascertained in tracing exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}B).
+For more transmissible influenza scenarios, even a fast isolation response, with an average of 3.4 days between symptom onset and isolation, requires $>70\%$ of contacts traced for most outbreaks to be controlled (Figure \ref{fig:prop-outbreak-control-R}B).
 \\
@@
 The contol of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C).
-Assuming ubiquitous access to home testing and that 63\% of symptomatic cases testing and isolating within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$).
+Assuming ubiquitous access to home testing and that 63\% of symptomatic cases test and isolate within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$).
 The LFT isolation strategy is also the only onset-to-isolation delay which can achieve 100\% of outbreaks controlled when all contacts are traced (Figure \ref{fig:prop-outbreak-control-R}C).
@@
 not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}).
-The shortest incubation period, H1N1, predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection and in some instances transmission while subclinical limit our ability to effectively tracing and isolate their potential infectees in time to prevent disease transmission.
+The shortest incubation period, H1N1, is predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection and in some instances transmission while subclinical limit our ability to effectively trace and isolate their potential infectees in time to prevent disease transmission.
 The variance between the influenza subtypes at each percentage of contacts traced and isolated varies (i.e.
@@
 
-The pattern of outbreak control is broadly similar between 10\% asymptomatic cases and the 0\% and 30\% asymptomatic scenarios (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
-Qualitative difference are in the 30\% asymptomatic scenario for LFT isolation, where the ability to control highly transmissible influenzas reduces to less than a quarter of outbreaks controlled at maxmimum contact tracing coverage (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
-When every case is symptomatic the likelihood of containing highly transmissible outbreaks drastically increases for all onset-to-isolation response delays (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
-\\
-
 The proportion of transmission that occurs before on the onset of symptoms has a moderate affect on epidemic containment for across isolation response rates (Figure \ref{fig:prop-outbreak-control-prop-presym-iso}).
@@
 
+The pattern of outbreak control is broadly similar between 10\% asymptomatic cases and the 0\% and 30\% asymptomatic scenarios (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
+Qualitative difference are in the 30\% asymptomatic scenario for LFT isolation, where the ability to control highly transmissible influenzas reduces to less than a quarter of outbreaks controlled at maximum contact tracing coverage (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
+When every case is symptomatic the likelihood of containing highly transmissible outbreaks drastically increases for all onset-to-isolation response delays (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
+\\
+
 The ability to control an epidemic that can readily transmit human-to-human is moderated by the initial number of infectious individuals that seed the outbreak.
@@
 When $R = 1.5$ the increase in the number of initial infections from 5 to 20 has a large decrease in outbreaks controlled at low levels of contact tracing, the same pattern when the number of initial infectors is twofold larger at 40, but these differences in outbreak control diminish as contact tracing increases towards 100\% (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
-When the number of cases seeding the outbreak is $>20$ at $R>2.5$ a large proportion of outbreaks become uncontrollable until contact tracing reaches $>60\%$.
-In the worst case transmissibility scenario ($R = 3.5$) if contact tracing is not activated until at least 20 individuals are infected then close to 100\% of contacts with need to be traced in order to control the majority of outbreaks (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+When the number of cases seeding the outbreak is $\geq 20$ at $R = 2.5$ a large proportion of outbreaks become uncontrollable until contact tracing reaches $\geq 80\%$.
+In the worst case transmissibility scenario ($R = 3.5$) if contact tracing is not activated until at least 20 individuals are infected then close to 100\% of contacts with need to be traced in order to control the majority of outbreaks; except for H1N1 where only 25\% of outbreaks are controlled if seeded with 40 cases (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
 \\
@@
 \clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_onset_to_isolation.png}
-\caption{The percentage of outbreaks controlled across values of isolation and contact tracing effectiveness, for outbreaks with an onset-to-isolation delay distribution that is either SARS-like, or COVID-19-like (from Wuhan).
-This is for a baselines scenario where the reproduction number is 1.5, the proportion of presymptomatic transmission is 15\%, all cases are assumed symptomatic, and the number of initial infections is 10.}
-\label{fig:prop-outbreak-control-onset-to-isolation}
-\end{figure}
 
@@
 \centering
+\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_prop_presym_iso.png}
+\caption{Stub.}
+\label{fig:prop-outbreak-control-prop-presym-iso}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
 \includegraphics[width=\textwidth]{../plots/prop_outbreak_control_prop_asym_iso.png}
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/prop_outbreak_control_prop_presym_iso.png}
-\caption{Stub.}
-\label{fig:prop-outbreak-control-prop-presym-iso}
-\end{figure}
-
 \end{document}

### 2025-05-15 inst/scripts/plot_controlled_outbreak.R
add plot_controlled_outbreak.R script and median_controlled_outbreak_size.png and prop_controlled_outbreak.png plots

@@ -1,0 +1,168 @@
+library(data.table)
+library(ggplot2)
+
+# read simulation results saved by running inst/scripts/run_analysis.R
+h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+
+h5n1_total_cases <- lapply(h5n1_results$sims, function(x) {
+  total_cases <- x[, .SD[which.max(week)], by = sim][, cumulative]
+  total_cases[total_cases > 500] <- 500
+  return(total_cases)
+})
+
+h1n1_total_cases <- lapply(h1n1_results$sims, function(x) {
+  total_cases <- x[, .SD[which.max(week)], by = sim][, cumulative]
+  total_cases[total_cases > 500] <- 500
+  return(total_cases)
+})
+
+h7n9_total_cases <- lapply(h7n9_results$sims, function(x) {
+  total_cases <- x[, .SD[which.max(week)], by = sim][, cumulative]
+  total_cases[total_cases > 500] <- 500
+  return(total_cases)
+})
+
+median_h5n1_controlled_outbreak_size <- vapply(
+  h5n1_total_cases,
+  function(x) median(x[x != 500]),
+  FUN.VALUE = numeric(1)
+)
+
+median_h1n1_controlled_outbreak_size <- vapply(
+  h1n1_total_cases,
+  function(x) median(x[x != 500]),
+  FUN.VALUE = numeric(1)
+)
+
+median_h7n9_controlled_outbreak_size <- vapply(
+  h7n9_total_cases,
+  function(x) median(x[x != 500]),
+  FUN.VALUE = numeric(1)
+)
+
+prop_h5n1_controlled_outbreak <- vapply(
+  h5n1_total_cases,
+  function(x) sum(x < 500),
+  FUN.VALUE = numeric(1)
+)
+
+prop_h1n1_controlled_outbreak <- vapply(
+  h1n1_total_cases,
+  function(x) sum(x < 500),
+  FUN.VALUE = numeric(1)
+)
+
+prop_h7n9_controlled_outbreak <- vapply(
+  h7n9_total_cases,
+  function(x) sum(x < 500),
+  FUN.VALUE = numeric(1)
+)
+
[…]

### 2025-05-15 inst/paper/main.tex
add results and plots for median_controlled_outbreak_size and prop_controlled_outbreak

@@
 
+The median size of contained influenza outbreaks is predominantly below 100 cases for all influenza subtypes and isolation response delays (Figure \ref{fig:median-controlled-outbreak-size}).
+There are a few parameter sets that have a median cumulative outbreak size between 200 --- 400, but very few have 400 --- 500.
+This elucidates that when an outbreak is eliminated by isolation alone or in combination with contact tracing across ascertainment levels, it is done so with a relatively small number of cases (Figure \ref{fig:median-controlled-outbreak-size}).
+The total size of controlled outbreaks is also relatively constant across influenza subtypes and isolation response (Figure \ref{fig:median-controlled-outbreak-size}), noting that that the percentage of controlled outbreaks between parameter sets varies drastically with some only producing uncontrolled outbreaks (Figure \ref{fig:prop-controlled-outbreak}, \ref{fig:prop-outbreak-control-R}).
+\\
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/median_controlled_outbreak_size.png}
+\caption{Median...}
+\label{fig:median-controlled-outbreak-size}
+\end{figure}
+
+\clearpage
+
 \section*{Discussion}
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/prop_controlled_outbreak.png}
+\caption{Median...}
+\label{fig:prop-controlled-outbreak}
+\end{figure}
+
 \end{document}

### 2025-05-15 inst/scripts/plot_max_weekly_cases.R
add max_weekly_cases script and plot

@@ -1,0 +1,104 @@
+library(data.table)
+library(ggplot2)
+
+# read simulation results saved by running inst/scripts/run_analysis.R
+h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+
+h5n1_median_max_weekly_cases <- vapply(
+  h5n1_results$sims,
+  function(x) {
+    median(x[, .SD[which.max(weekly_cases)], by = sim][, weekly_cases])
+  },
+  FUN.VALUE = numeric(1)
+)
+
+h1n1_median_max_weekly_cases <- vapply(
+  h1n1_results$sims,
+  function(x) {
+    median(x[, .SD[which.max(weekly_cases)], by = sim][, weekly_cases])
+  },
+  FUN.VALUE = numeric(1)
+)
+
+h7n9_median_max_weekly_cases <- vapply(
+  h7n9_results$sims,
+  function(x) {
+    median(x[, .SD[which.max(weekly_cases)], by = sim][, weekly_cases])
+  },
+  FUN.VALUE = numeric(1)
+)
+
+h5n1_max_weekly_cases <- rbindlist(h5n1_results$data)[, c("index_R0", "control_effectiveness", "delay")]
+h5n1_max_weekly_cases[, max_weekly_cases := h5n1_median_max_weekly_cases]
+h5n1_max_weekly_cases[, subtype := as.factor("H5N1")]
+
+h1n1_max_weekly_cases <- rbindlist(h1n1_results$data)[, c("index_R0", "control_effectiveness", "delay")]
+h1n1_max_weekly_cases[, max_weekly_cases := h1n1_median_max_weekly_cases]
+h1n1_max_weekly_cases[, subtype := as.factor("H1N1")]
+
+h7n9_max_weekly_cases <- rbindlist(h7n9_results$data)[, c("index_R0", "control_effectiveness", "delay")]
+h7n9_max_weekly_cases[, max_weekly_cases := h7n9_median_max_weekly_cases]
+h7n9_max_weekly_cases[, subtype := as.factor("H7N9")]
+
+rm(h5n1_results)
+rm(h1n1_results)
+rm(h7n9_results)
+
+max_weekly_cases <- rbindlist(
[…]

### 2025-05-15 inst/paper/main.tex
add results and figure for max weekly cases

@@
 
+Contact tracing systems have a capacity limit which if exceeded will lower ascertainment of case contacts.
+The infinite susceptible population assumption in our model means uncontrolled outbreaks and contacts requiring tracing can become infinitely large.
+However, we find that in the early acute phase of both contained and uncontained epidemics the maximum weekly cases exceeds 600 in some scenarios (Figure \ref{fig:max-weekly-cases}).
+These cases and their contacts require testing and potentially isolating, showing the rapid speed at which outbreaks can overburden surveillance systems.
+The fact that some outbreaks have more maximum weekly case than the outbreak limit we define shows that outbreaks can become almost impossible to control on the scale of weeks from the initial cases (Figure \ref{fig:max-weekly-cases}).
+When isolation and 100\% of contacts are traced in the most optimal isolation response scenario (LFT) the maximum number of cases weekly is around 450 ($R = 3.5$).
+\\
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/max_weekly_cases.png}
+\caption{Max...}
+\label{fig:max-weekly-cases}
+\end{figure}
+
 \section*{Discussion}

### 2025-05-30 inst/scripts/plot_delay_distributions.R
add delay_distributions plot and script

@@ -1,0 +1,20 @@
+library(patchwork)
+
+source(file.path("inst", "scripts", "plot_incubation_period.R"))
+source(file.path("inst", "scripts", "plot_onset_to_isolation.R"))
+source(file.path("inst", "scripts", "plot_prop_presymptomatic_transmission.R"))
+
+delay_distributions_plot <- incubation_period_plot /
+  onset_to_isolation_plot /
+  prop_presymptomatic_transmission_plot +
+  plot_annotation(tag_levels = 'A')
+
+ggplot2::ggsave(
+  file.path("inst", "plots", "delay_distributions.png"),
+  plot = delay_distributions_plot,
+  device = "png",
+  width = 150,
+  height = 215,
+  units = "mm",
+  dpi = 300
+)

### 2025-05-30 inst/paper/main.tex
update ms with @adamkucharski comments and added references to bib

@@
 The COVID-19 pandemic caused over 7 million deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}.
-Less transmissible pathogens, such as Ebola, can cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}.
+Other pathogens, including those less transmissible, with longer delays between infections, and with less presymptomatic transmission, can also cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}.
 Thus, timely and accurate response is critical early in an outbreak when the pandemic or epidemic potential is still uncertain \citep{kucharskiControllingMinorOutbreaks2024}.
@@
 
-The likelihood of an outbreak being controlled, in other words going locally-extinct or -eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical.
-The primary epidemiological determinant for the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases  per primary case.
-The higher the reproduction number the less likely a disease, without intervention, is to cease transmission in a population.
+The likelihood of an outbreak being controlled, in other words being locally eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical.
+A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases  per primary case.
+If the reproduction number is greater than 1, the incidence of infections is likely to increase, and intervention is requireed to cease transmission in a population.
 If a pathogen is subcritical (i.e.
-$R < 1$) it will eventually stop spreading but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}.
+$R < 1$), incidence will decline and eventually stop spreading, but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}.
 Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}.
 Higher individual-level variation for a given reproduction number leads to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading).
-In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{kucharskiEarlyDynamicsTransmission2020}.
+In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}.
 \\
@@
 These interventions vary in scope and specificity, from targeted measures such as isolation and contact tracing, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020).
-The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness (Longini et al., 2005).
-For highly transmissible pathogens (e.g.
-$R \geq 2.5$) with short generation times (e.g.
-$\leq 3$ days), swift and effective interventions are required to achieve containment.
+The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness \citep{longiniContainingPandemicInfluenza2005}.
+For highly transmissible pathogens with short generation times, swift and effective interventions are required to achieve containment.
 Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}.
@@
 Isolation of infected individuals aims to break the chain of transmission by preventing further infection of susceptible individuals.
-The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to infectiousness optimal.
+The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to becoming infectious optimal.
 This is challenging for most pathogens because infectiousness preceeds prodromal symptoms.
-While contact tracing is the process of finding the contacts of infected individuals.
+Contact tracing is the process of finding the contacts of infected individuals.
 Both isolation and contact tracing are often used NPIs to try and curtail the growth of an epidemic.
-Contact tracing can be forward, whereby infected individuals are asked who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed; or backward contact tracing whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source.
-Isolation and contact tracing is a targeted approach, like ring vaccination, that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{kucharskiEffectivenessRingVaccination2016, ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020, whittakerQuantifyingImpactBroadly2024}.
+Forward contact tracing asks infected individuals who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed.
+Isolation and contact tracing is a targeted approach that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020}.
 Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}.
-However, it's important to remember that it's not a panacea for all outbreak scenarios.
-In epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}.
+However, in epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}.
 \\
 
-It has become widely accepted that influenza viruses cannot be controlled with isolation and contact tracing because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
-However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza.
+It has become widely accepted that influenza viruses cannot be controlled by isolating individuals at symptom onset and contact tracing, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
+However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential.
 The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic.
-Thus, influenza subtypes that have negligible levels of existing immunity in the population are probable epidemic and pandemic candidates.
+Thus, influenza subtypes that have low levels of existing immunity in the population are probable epidemic and pandemic candidates.
 This is the case for H5N1 \citep{Ward2024.12.11.24318702} and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}.
@@
 In effect the ability to control an epidemic with isolation and contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent \citep{reyna-laraVirusSpreadContact2021}.
-The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval (Ward et al., 2024).
+The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval \citep{Ward2024.12.11.24318702}.
 This study looks to define the speed and effectiveness of isolation under different response delays across epidemic scenarios to inform outbreak response; quantifying the impact of reducing the delay to isolation on outbreak containment and the utility of rapid testing.
@@
 The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (infinite population size, without depletion of susceptibility).
+Influenza subtypes have shown moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}, but evidence is limited and the negative binomial allows flexibility between highly heterogeneous and homogeneous individual-level transmission.
 The mean of the negative binomial offspring distribution is the basic reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}.
@@
 generation time) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
+This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, infectiousness onset and onward transmission \citep{hellewellFeasibilityControllingCOVID192020}.
+This formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is accurately estimated but the generation time (or serial interval) is estimated with more uncertainty or unknown.
 This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector.
@@
 Symptomatic cases that have been missed when contact tracing their infectors are isolated at the onset-to-isolation delay after their symptom onset time.
-Additionally, contact tracing misses infections by an asymptomatic infector.
+Additionally, contact tracing misses infections by an asymptomatic, or paucisymptomatic (subclinical), infector.
 In this event the infectees are missed by contact tracing owing to the infector not being known to the response system.
@@
 We extracted incubation periods from the epidemiological literature that have been previously estimated for the chosen Influenza subtypes.
-All incubation periods were found to be well described by a Weibull distribution, \cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:incub}).
+All incubation periods were found to be well described by a Weibull distribution, \cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}A).
 \cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed.
-For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:incub}).
+For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}A).
+Comparatively, the incubation period for H1N1 is noticably shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}A).
 \\
@@
 \centering
-\includegraphics[width=\textwidth]{../plots/incubation_period.png}
-\caption{Incubation period for the Influenza subtypes: H1N1, H5N1 and H7N9.}
-\label{fig:incub}
+\includegraphics[width=\textwidth,height=0.85\textheight]{../plots/delay_distributions.png}
+\caption{(A) Incubation period for the Influenza subtypes: H1N1, H5N1 and H7N9.
+(B) Three assumed onset-to-isolation delays used in the simuluation experiment.
+The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China \citep{liEarlyTransmissionDynamics2020}, the median onset-to-isolatiojn is 8.1 days (blue dashed line).
+The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a}, the median onset-to-isolation time is 3.4 days (red dashed line).
+The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days (orange dashed line).
+(C) Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skew-normal distribution.}
+\label{fig:delay-distributions}
 \end{figure}
@@
 2003) and a slower response based on COVID-19 in Wuhan, China (Li et al.
-2020) (Figure \ref{fig:onset-to-isolation}).
-Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:onset-to-isolation}).
+2020) (Figure \ref{fig:delay-distributions}B).
+Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:delay-distributions}B).
 These two empirically informed delay distributions provide a guide to possible response times in the case of an influenza outbreak and will enable us to evaluate if such delays are too long to contain a disease that can transmit rapidly such as influenza.
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/onset_to_isolation.png}
-\caption{Three assumed onset-to-isolation delays used in the simuluation experiment.
-The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China \citep{liEarlyTransmissionDynamics2020}, the median onset-to-isolatiojn is 8.1 days (blue dashed line).
-The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a}, the median onset-to-isolation time is 3.4 days (red dashed line).
-The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days (orange dashed line).}
-\label{fig:onset-to-isolation}
-\end{figure}
-
-\clearpage
-
 We define the proportion of presymptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$).
-The shape parameter controls the proportion of presymptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:prop-presym-trans}).
-We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:prop-presym-trans}).
+The shape parameter controls the proportion of presymptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:delay-distributions}C).
+We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:delay-distributions}C).
 Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective isolation and contact tracing is hypothesied to be.
@@
 \\
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/prop_presymptomatic_transmission.png}
-\caption{Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skew-normal distribution.}
-\label{fig:prop-presym-trans}
-\end{figure}
-
-\clearpage
 
@@
 
-The incubation period for H1N1 is noticably shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:incub}). \\
-
-Predictably, the ability to control an influenza epidemic is highly dependent on the reproduction number and the speed of isolation after symptom onset (Figure \ref{fig:prop-outbreak-control-R}).
+The ability to control an influenza epidemic is highly dependent on the reproduction number and the speed of isolation after symptom onset (Figure \ref{fig:prop-outbreak-control-R}).
 A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of contacts traced, except in cases where they are equal because zero or all outbreaks are contained, across all onset-to-isolation scenarios (Figure \ref{fig:prop-outbreak-control-R}).
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/median_controlled_outbreak_size.png}
-\caption{Median...}
-\label{fig:median-controlled-outbreak-size}
-\end{figure}
-
-\clearpage
-
 Contact tracing systems have a capacity limit which if exceeded will lower ascertainment of case contacts.
@@
 When isolation and 100\% of contacts are traced in the most optimal isolation response scenario (LFT) the maximum number of cases weekly is around 450 ($R = 3.5$).
-\\
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/max_weekly_cases.png}
-\caption{Max...}
-\label{fig:max-weekly-cases}
-\end{figure}
 
@@
 \item Isolation and contact tracing complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals \citep{haydenPerspectivesAntiviralUse2001}.
+\item This paper has focused on forward contact tracing.
+Backward contact tracing, whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source, is applicable to infectious with long infectious periods (e.g.
+HIV), but is not effective for more ephemeral infections with faster transmission, as is the case for influenza.
+\item Contact tracing and isolation is a targeted approach analogous to ring vaccination.
+It looks to prevent onward transmission by, either isolating or quarantining contacts in the case of contact tracing, or vaccinating contacts in the case of ring vaccination \citep{kucharskiEffectivenessRingVaccination2016, whittakerQuantifyingImpactBroadly2024}.
+Ring vaccination for an influenza outbreak would not be effective due to the generation time, proportion of presymptomatic and asymptomatic, and likelihood of geographically widespread (i.e.
+non-localised) transmission.
 \end{itemize}
@@
 \centering
+\includegraphics[width=\textwidth]{../plots/median_controlled_outbreak_size.png}
+\caption{Median...}
+\label{fig:median-controlled-outbreak-size}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
 \includegraphics[width=\textwidth]{../plots/prop_controlled_outbreak.png}
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/max_weekly_cases.png}
+\caption{Max...}
+\label{fig:max-weekly-cases}
+\end{figure}
+
 \end{document}

### 2025-06-02 inst/paper/main.tex
update intro and methods from discussion with Adam and update .bib, add model schematic figure and remove tikz figure

@@
 \title{Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks}
-\author{Joshua W. Lambert}
+\author{Joshua W. Lambert, Adam J. Kucharski}
 \date{}
@@
 
-The likelihood of an outbreak being controlled, in other words being locally eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical.
+The likelihood of an outbreak causing sustained transmission in a population is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical.
 A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases  per primary case.
@@
 $R < 1$), incidence will decline and eventually stop spreading, but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}.
-Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}.
-Higher individual-level variation for a given reproduction number leads to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading).
+Individual-level heterogeneity in transmission also influences the possibility a pathogen will fail to maintain transmission due to stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}.
+Higher individual-level variation for a given reproduction number leads to, on average, more outbreaks being unsustained (however a small number of outbreaks may become large due to superspreading).
 In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}.
@@
 The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness \citep{longiniContainingPandemicInfluenza2005}.
-For highly transmissible pathogens with short generation times, swift and effective interventions are required to achieve containment.
+For highly transmissible pathogens with short generation times (e.g.
+influenza), swift and effective interventions are required to achieve containment.
 Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}.
@@
 
-Isolation of infected individuals aims to break the chain of transmission by preventing further infection of susceptible individuals.
+Isolation of symptomatic infections aims to break the chain of transmission by preventing further infection of susceptible individuals.
 The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to becoming infectious optimal.
-This is challenging for most pathogens because infectiousness preceeds prodromal symptoms.
+This can make it challenging to control outbreaks with symptom-based isolation, as for most pathogens infectiousness preceeds prodromal symptoms.
+Quarantining contacts exposed to an infectious individuals, independent of whether they later become a case, is a control strategy to reduce transmission before or shortly after symptom onset.
 Contact tracing is the process of finding the contacts of infected individuals.
@@
 Forward contact tracing asks infected individuals who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed.
-Isolation and contact tracing is a targeted approach that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020}.
+Isolation, quarantine, and contact tracing is a targeted approach that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020}.
 Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}.
@@
 
-It has become widely accepted that influenza viruses cannot be controlled by isolating individuals at symptom onset and contact tracing, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
+Several pre-COVID-19 studies have concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and contact tracing, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
 However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential.
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/model-schematic.png}
+\caption{Model schematic of the branching process with isolation, quarantine and contact tracing implemented in the \texttt{\{ringbp\}} R package.
+(A) The transmission chain from the index case (Person A), infection events are shown by black vertical dashed lines.
+The time individuals are infected is shown by the red dot, the time of symptom onset is shown by the orange dot, and the time time individuals are isolated is shown by the blue dot.
+Quarantine is optional in the model, assuming quarantine is active it is shown by the small green dot with a Q.
+The time at whcih a infectors isolation triggers the isolation of a symtomatic case is shown by the blue vertical dashed line (Person A to Person B).
+The isolation times that would have been triggered after a delay from symptom onset, but are after isolation was triggered by contact tracing shown by a blue dot with a cross and connected by a dotted line (Person B).
+Person C is an asymptomatic individual (asymptomatic cases can infect others).
+Person D is missed by contact tracing (contacts are traced with a probability $\rho$, and thus missed at $1 - \rho$).
+The time Person D would have isolated if ascertained is shown by the vertical blue mark.
+Assuming quarantine is active, Person F would be isolated before symptom onset at the time of the infector's (Person D) isolation time.
+If quarantine is not active, then Person F would be isolated after a delay from symptom onset, as they had yet to show symptoms when the infector was isolated.
+(B) A branching process diagram showing who-infected-whom.}
+\label{fig:ringbp-model}
+\end{figure}
+
 The time between an infector being infected and then infecting an infectee (i.e.
-generation time) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
-This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, infectiousness onset and onward transmission \citep{hellewellFeasibilityControllingCOVID192020}.
-This formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is accurately estimated but the generation time (or serial interval) is estimated with more uncertainty or unknown.
-This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector.
+generation time) is computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
+This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, onset of symptoms, and onward transmission \citep{hellewellFeasibilityControllingCOVID192020}.
+This allows the proportion presymptomatic transmission to be specified, rather than defining incubation period and generation time distributions, which only indirectly allows assumptions of presymptomatic infectiousness to be derived.
+This simple formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is estimated but the generation time is estimated with more uncertainty or unknown, and the serial interval cannot be substituted in without introducing bias \citep{brittonEstimationEmergingEpidemics2019a, lehtinenRelationshipSerialInterval2021}.
+\\
+
+Given the incubation period for each case, the proportion of the infections that occur prior to, or after, the symptom onset of an infector is controlled by the parameterisation of the skew normal distribution (Figure \ref{fig:delay-distributions}C).
 The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed.
-The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution.
+The scale parameter ($\omega$) is fixed at 2, this parameter (with the shape parameter) defines the variance of the distribution.
 We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2).
@@
 \\
-
-\begin{figure}[ht]
-\centering
-{\sffamily
-  \begin{tikzpicture}[every node/.style={font=\tiny}]
-  % Labels above the A row
-  \node at (0,0.5) {Infected};
-  \node at (2,0.5) {Symptomatic};
-  \node at (5,0.5) {Isolated};
-
-  % Top row: A circles with red background
-  \fill[red!30] (0,0) circle (5pt);
-  \draw (0,0) circle (5pt);
-  \node at (0,0) {A};
-
-  \fill[red!30] (2,0) circle (5pt);
-  \draw (2,0) circle (5pt);
-  \node at (2,0) {A};
-
-  \fill[red!30] (5,0) circle (5pt);
-  \draw (5,0) circle (5pt);
-  \node at (5,0) {A};
-
-  % Horizontal line between A circles
-  \draw (0,0) -- (2,0);
-  \draw (2,0) -- (5,0);
-
-  % Individual A delays
-  \draw[<->] (0.5,-0.3) -- (1.5,-0.3);
-  \node at (1,-0.5) {Incubation period};
-  \draw[<->] (2,-0.3) -- (5,-0.3);
-  \node at (3.5,-0.5) {Onset-to-isolation};
-
-  % Labels above the B row
-  \node at (4,-1.5) {Infected};
-  \node at (6,-1.5) {Symptomatic};
-  \node at (8,-1.5) {Isolated};
-
-  % Middle row: B circles with blue background
-  \fill[blue!30] (4,-2) circle (5pt);
-  \draw (4,-2) circle (5pt);
-  \node at (4,-2) {B};
-
-  \fill[blue!30] (6,-2) circle (5pt);
-  \draw (6,-2) circle (5pt);
-  \node at (6,-2) {B};
-
-  \fill[blue!30] (8,-2) circle (5pt);
-  \draw (8,-2) circle (5pt);
-  \node at (8,-2) {B};
-
-  % Horizontal line between B circles
-  \draw (4,-2) -- (6,-2);
-  \draw (6,-2) -- (8,-2);
-
-  % Labels above the C row
-  \node at (4.5,-3.5) {Infected};
-  \node at (7.5,-3.5) {Symptomatic};
-  \node at (9,-3.5) {Isolated};
-
-  % Bottom row: C circles with green background
-  \fill[green!30] (4.5,-4) circle (5pt);
-  \draw (4.5,-4) circle (5pt);
-  \node at (4.5,-4) {C};
-
-  \fill[green!30] (7.5,-4) circle (5pt);
-  \draw (7.5,-4) circle (5pt);
-  \node at (7.5,-4) {C};
-
-  \fill[green!30] (9,-4) circle (5pt);
-  \draw (9,-4) circle (5pt);
-  \node at (9,-4) {C};
-
-  % Horizontal line between C circles
-  \draw (4.5,-4) -- (7.5,-4);
-  \draw (7.5,-4) -- (9,-4);
-
-  % Labels above the D row
-  \node at (8,-5.5) {Infected};
-
-  % Individual C circles with green background
-  \fill[teal!30] (8,-6) circle (5pt);
-  \draw (8,-6) circle (5pt);
-  \node at (8,-6) {D};
-
-  % arrow for asymptomatic D individual
-  \draw[->] (8,-6) -- (11,-6); % arrow for asymptomatic
-
-  % Labels above the C row
-  \node at (8.75,-7.5) {Infected};
-  \node at (10,-7.5) {Symptomatic};
-  \node at (11,-7.5) {Isolated};
-
-  % E individual
-  \fill[violet!30] (8.75,-8) circle (5pt);
-  \draw (8.75,-8) circle (5pt);
-  \node at (8.75,-8) {E};
-
-  \fill[violet!30] (10,-8) circle (5pt);
-  \draw (10,-8) circle (5pt);
-  \node at (10,-8) {E};
-
-  \fill[violet!30] (11,-8) circle (5pt);
-  \draw (11,-8) circle (5pt);
-  \node at (11,-8) {E};
-
-  % Horizontal line between E circles
-  \draw (8.75,-8) -- (10,-8);
-  \draw (10,-8) -- (11,-8);
-
-  % Vertical dashed lines
-  \draw[dashed] (4,0) -- (4,-2);   % A to B
-  \draw[dashed] (4.5,-2) -- (4.5,-4);  % B to C
-  \draw[dashed] (8,-4) -- (8,-6);  % C to D
-  \draw[dashed] (8.75,-4) -- (8.75,-8);  % C to E
-
-  \end{tikzpicture}
-}
-\caption{A schematic of the branching process model with isolation.}
-\label{fig:ringbp-model}
-\end{figure}
 

### 2025-06-07 inst/paper/main.tex
Edits to introduction

@@
 
-\title{Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks}
-\author{Joshua W. Lambert, Adam J. Kucharski}
+\title{Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study}
+\author{Joshua W. Lambert, Authors TBC, Adam J. Kucharski}
 \date{}
@@
 
-Infectious disease outbreaks pose an everpresent public health threat.
-The COVID-19 pandemic caused over 7 million deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}.
-Other pathogens, including those less transmissible, with longer delays between infections, and with less presymptomatic transmission, can also cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}.
-Thus, timely and accurate response is critical early in an outbreak when the pandemic or epidemic potential is still uncertain \citep{kucharskiControllingMinorOutbreaks2024}.
+Outbreaks of severe novel infectious diseases pose a major public health threat.
+The COVID-19 pandemic resulted in over 7 million reported deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}.
+Recent human cases of avian influenza A/H5N1, alongside transmission, illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption [CITE: Antia et al, Nature, 2003 + key papers on H5N1 - see my PLOS Bio piece].
+It is therefore crucial to understand the effectiveness and feasibility of different intervention strategies in the face of such infections.
+
+%Other pathogens, including those less transmissible, with longer delays between infections, and with less presymptomatic transmission, can also cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}.
+Timely and accurate response is therefore critical early in an outbreak when the pandemic or epidemic potential is still uncertain \citep{kucharskiControllingMinorOutbreaks2024}.
 An effective response requires understanding the efficacy and feasibility of intervention strategies across pathogens with varying epidemiological characteristics \citep{fraserFactorsThatMake2004}.
+\\ AK: Removed to improve flow, as this is mentioned in more detail later in intro.
+
+The likelihood of a novel outbreak causing sustained transmission depends both the characteristics of the pathogen and its host population, as well control measures  implemented in response to early cases.
+A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases by a typical infectious host.
+If the reproduction number is greater than 1, we would expect the incidence of infections to increase over time.
+If $R$ is below 1, we would expect incidence to decline on average, but several cases could still occur before transmission ceases \citep{farringtonDistributionTimeExtinction1999}.
+Two additional factors can influence the probability a pathogen with $R$ above 1 will successfully establish within a population~\citep{lloyd-smithSuperspreadingEffectIndividual2005}.
+If there is a large degree of individual-level variation in transmission, the majority of infectious individuals will infect nobody, even though a minority will infect many; as a result, most newly introduced outbreaks will fail to take off.
+In addition, the likelihood of establishment increases if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}.
+\\ % AK: Note 'primary case' means first in outbreak
+
+Early in a novel outbreak, pharmaceutical options such as treatments and vaccines may not yet be available, leading to reliance on non-pharmaceutical interventions (NPIs) to reduce infection and disease.
+These interventions include individual-level targeted measures such as isolation and contact tracing, to population-level policies such as venue closures or restrictions on social mixing (Huag et al., 2020) [AK: probably a better reference out there, e.g.
+Sharma et al, Nature Comms].
+The success of such interventions in curtailing transmission before it becomes sustained depends on several factors, including: the type of intervention, timing of introduction, and the intervention's effectiveness \citep{longiniContainingPandemicInfluenza2005}.
+For highly transmissible pathogens with a short generation times (e.g.
+pandemic influenza), incidence of new infections can grow quickly, outpacing a response.
+Moreover, pathogens such as influenza, which have a large proportion of onward transmission prior to symptom onset and have asymptomatic (subclinical) cases, are less easily controlled via isolation of symptomatic cases and tracing and quarantining of their contacts~\citep{fraserFactorsThatMake2004}.
 \\
 
-The likelihood of an outbreak causing sustained transmission in a population is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical.
-A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases  per primary case.
-If the reproduction number is greater than 1, the incidence of infections is likely to increase, and intervention is requireed to cease transmission in a population.
-If a pathogen is subcritical (i.e.
-$R < 1$), incidence will decline and eventually stop spreading, but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}.
-Individual-level heterogeneity in transmission also influences the possibility a pathogen will fail to maintain transmission due to stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}.
-Higher individual-level variation for a given reproduction number leads to, on average, more outbreaks being unsustained (however a small number of outbreaks may become large due to superspreading).
-In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}.
-\\
-
-A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs).
-Early in an outbreak --- especially in the case of a novel pathogen --- NPIs are often the first line of defence, as pharmaceutical options may not yet be available.
-These interventions vary in scope and specificity, from targeted measures such as isolation and contact tracing, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020).
-The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness \citep{longiniContainingPandemicInfluenza2005}.
-For highly transmissible pathogens with short generation times (e.g.
-influenza), swift and effective interventions are required to achieve containment.
-Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}.
-\\
-
-Isolation of symptomatic infections aims to break the chain of transmission by preventing further infection of susceptible individuals.
+%Isolation of symptomatic infections aims to break the chain of transmission by preventing further infection of susceptible individuals.
 The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to becoming infectious optimal.
-This can make it challenging to control outbreaks with symptom-based isolation, as for most pathogens infectiousness preceeds prodromal symptoms.
+This can make it challenging to control outbreaks with symptom-based isolation, as for most pathogens infectiousness preceeds prodromal symptoms [AK: reference?
+This is a strong claim – best to focus on flu I think].
 Quarantining contacts exposed to an infectious individuals, independent of whether they later become a case, is a control strategy to reduce transmission before or shortly after symptom onset.
@@
 However, in epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}.
-\\
+\\ AK: This is probably too much fundamental background for an applied journal – readers will know what isolation and contact tracing is.
 
-Several pre-COVID-19 studies have concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and contact tracing, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
+Several pre-COVID-19 studies  concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and tracing their contacts, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
 However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential.
@@
 Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}.
-\\
+\\ [AK: These headline fatality risks are likely to be biased: https://kucharski.substack.com/p/how-fatal-is-h5n1-influenza]
 

### 2025-06-19 inst/paper/main.tex
Changes from 19 June meeting

@@
 
-Outbreaks of severe novel infectious diseases pose a major public health threat.
+Outbreaks of severe respiratory infections pose a major public health threat.
 The COVID-19 pandemic resulted in over 7 million reported deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}.
-Recent human cases of avian influenza A/H5N1, alongside transmission, illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption [CITE: Antia et al, Nature, 2003 + key papers on H5N1 - see my PLOS Bio piece].
+Influenza subtypes to which populations have low levels of pre-existing immunity are a particular concern as potential pandemic candidates.
+This includes influenza A/H5N1 \citep{Ward2024.12.11.24318702}, A/H7N9 \citep{tannerPandemicPotentialAvian2015}, and A/H1N1p when it caused caused a pandemic in 2009 \citep{fraserPandemicPotentialStrain2009}.
+Recent human cases of avian influenza A/H5N1, alongside transmission, further illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption [CITE: Antia et al, Nature, 2003 + key papers on H5N1 - see my PLOS Bio piece].
 It is therefore crucial to understand the effectiveness and feasibility of different intervention strategies in the face of such infections.
@@
 
-The likelihood of a novel outbreak causing sustained transmission depends both the characteristics of the pathogen and its host population, as well control measures  implemented in response to early cases.
+%The likelihood of a novel outbreak causing sustained transmission depends both the characteristics of the pathogen and its host population, as well control measures  implemented in response to early cases.
 A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases by a typical infectious host.
@@
 Moreover, pathogens such as influenza, which have a large proportion of onward transmission prior to symptom onset and have asymptomatic (subclinical) cases, are less easily controlled via isolation of symptomatic cases and tracing and quarantining of their contacts~\citep{fraserFactorsThatMake2004}.
+
+Several pre-COVID-19 studies concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and tracing their contacts, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
+However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential.
 \\
@@
 
-Several pre-COVID-19 studies  concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and tracing their contacts, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
-However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential.
-The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic.
-Thus, influenza subtypes that have low levels of existing immunity in the population are probable epidemic and pandemic candidates.
-This is the case for H5N1 \citep{Ward2024.12.11.24318702} and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}.
-Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}.
+%The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic.
+Thus, Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}.
 \\ [AK: These headline fatality risks are likely to be biased: https://kucharski.substack.com/p/how-fatal-is-h5n1-influenza]
 
-In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model.
-In effect the ability to control an epidemic with isolation and contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent \citep{reyna-laraVirusSpreadContact2021}.
-The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval \citep{Ward2024.12.11.24318702}.
-This study looks to define the speed and effectiveness of isolation under different response delays across epidemic scenarios to inform outbreak response; quantifying the impact of reducing the delay to isolation on outbreak containment and the utility of rapid testing.
+In this study we evaluate the effectiveness of controlling outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model of isolation and contact tracing.
+In particular, we investigate the assumption that influenza will inevitably outpace ability to trace and isolate sufficiently effectively for control, and whether this assumption holds for all pathogen subtypes given emerging estimates of incubation period and serial interval \citep{Ward2024.12.11.24318702}.
+Finally, we quantify the speed and effectiveness of isolation for an effective response under different response delays across epidemic scenarios, and assess the impact of reducing the delay to isolation on outbreak containment alongside use of rapid testing.
 
@@
 
-\subsection*{Branching process model with isolation}
+\subsection*{Transmission model}
+We used a branching process transmission model to assess the potential effectiveness of isolation and contact tracing to control an outbreak of pandemic-potential influenza \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}).
+In the model, each infectious individual generates secondary cases from a negative binomial distribution with a given reproduction number $R$ and dispersion parameter ($k$).
+Onwards transmission from infectious individuals could be reduced in the following ways: 1) \textit{symptomatic individuals isolate effectively}, 2) \textit{contacts of symptomatic individuals are traced and isolate if they become symptomatic}, 3) \textit{contacts of symptomatic individuals are traced and effectively quarantined immediately}.
+Symptomatic individuals that have been traced (i.e.
+infectees) are isolated either 1) after the onset-to-isolation delay after symptom onset, or 2) when their infector is isolated.
+The isolation time of the infectee is whichever occurs first of these two events.
 
-To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}).
-The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (infinite population size, without depletion of susceptibility).
-Influenza subtypes have shown moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}, but evidence is limited and the negative binomial allows flexibility between highly heterogeneous and homogeneous individual-level transmission.
-The mean of the negative binomial offspring distribution is the basic reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}.
-\\
+We include several mechanisms by which isolation and contact tracing can be imperfect at reducing onward disease transmission.
+The proportion of contacts identified by contact tracing can range between 0\% (i.e.
+no contacts are traced) to 100\% (i.e.
+every contact of every symptomatic case is traced).
+Because the model focuses on onwards transmission prevented, identifying 50\% of cases and preventing 100\% of their onwards transmission via contact tracing is equivalent to identifying 100\% of cases and preventing 50\% of their onwards transmission.
+Subsequent symptomatic cases that have not been contact traced are only isolated after the onset-to-isolation delay post symptom onset.
+Additionally, we assume infections have an independent probability of being asymptomatic or paucisymptomatic; these infections were not identified and therefore their contacts were not traced.
+The advantage of this approach is that it enables estimation of the overall population transmission effect of multiple interventions without requiring each step to be individually specified as a parameter.
 
-There are three categories for individuals in the simulation, with each having their own parameterisation for the negative binomial offspring distribution: 1) \textit{community} (i.e.
-general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}.
-Each group have a distinct $R$ and dispersion ($k$) parameter due to the assumed differences in transmission dynamics between groups.
-\\
+\subsection*{Pathogen epidemiological parameters} \label{epiparameters}
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/model-schematic.png}
-\caption{Model schematic of the branching process with isolation, quarantine and contact tracing implemented in the \texttt{\{ringbp\}} R package.
-(A) The transmission chain from the index case (Person A), infection events are shown by black vertical dashed lines.
-The time individuals are infected is shown by the red dot, the time of symptom onset is shown by the orange dot, and the time time individuals are isolated is shown by the blue dot.
-Quarantine is optional in the model, assuming quarantine is active it is shown by the small green dot with a Q.
-The time at whcih a infectors isolation triggers the isolation of a symtomatic case is shown by the blue vertical dashed line (Person A to Person B).
-The isolation times that would have been triggered after a delay from symptom onset, but are after isolation was triggered by contact tracing shown by a blue dot with a cross and connected by a dotted line (Person B).
-Person C is an asymptomatic individual (asymptomatic cases can infect others).
-Person D is missed by contact tracing (contacts are traced with a probability $\rho$, and thus missed at $1 - \rho$).
-The time Person D would have isolated if ascertained is shown by the vertical blue mark.
-Assuming quarantine is active, Person F would be isolated before symptom onset at the time of the infector's (Person D) isolation time.
-If quarantine is not active, then Person F would be isolated after a delay from symptom onset, as they had yet to show symptoms when the infector was isolated.
-(B) A branching process diagram showing who-infected-whom.}
-\label{fig:ringbp-model}
-\end{figure}
+In this study we evaluate the feasibility of controlling outbreaks of A/H5N1, A/H1N1, and A/H7N9-like infections.\cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}A).
+\cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed.
+For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}A).
+Comparatively, the incubation period for H1N1 shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}A).
+We consider four reproduction number ($R$) values when simulating transmission: 1.1, 1.5, 2.5 and 3.5.
+These are based on similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution.
 
-The time between an infector being infected and then infecting an infectee (i.e.
-generation time) is computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
-This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, onset of symptoms, and onward transmission \citep{hellewellFeasibilityControllingCOVID192020}.
-This allows the proportion presymptomatic transmission to be specified, rather than defining incubation period and generation time distributions, which only indirectly allows assumptions of presymptomatic infectiousness to be derived.
-This simple formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is estimated but the generation time is estimated with more uncertainty or unknown, and the serial interval cannot be substituted in without introducing bias \citep{brittonEstimationEmergingEpidemics2019a, lehtinenRelationshipSerialInterval2021}.
-\\
-
-Given the incubation period for each case, the proportion of the infections that occur prior to, or after, the symptom onset of an infector is controlled by the parameterisation of the skew normal distribution (Figure \ref{fig:delay-distributions}C).
-The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed.
-The scale parameter ($\omega$) is fixed at 2, this parameter (with the shape parameter) defines the variance of the distribution.
-We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2).
-The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution).
-In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector.
-\\
-
-The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing (Figure \ref{fig:ringbp-model}).
-When a case is symptomatic they are isolated after a delay.
-The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation).
-Any infection times that are sampled to occur after the isolated of an infector are remove, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times.
-In other words, once a case is isolated at time $t_{isolated}$, the probability of that case infecting anyone at $t > t_{isolated}$ is zero.
-Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).
-Symptomatic individuals that are traced are isolated either when 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated.
-The isolation of the infectee is whatever occurs first of these two events.
-\\
-
-There are several mechanisms where isolation and contact tracing are imperfect at reducing onward disease transmission.
-The proportion of contacts missed by contact tracing is defined by the proportion of ascertainment and can range between zero, that is nobody is traced, to one, every case is traced.
-Symptomatic cases that have been missed when contact tracing their infectors are isolated at the onset-to-isolation delay after their symptom onset time.
-Additionally, contact tracing misses infections by an asymptomatic, or paucisymptomatic (subclinical), infector.
-In this event the infectees are missed by contact tracing owing to the infector not being known to the response system.
-\\
+We three presymptomatic transmission scenarios: <1\%, 15\%, 30\% of transmission occurring prior to symptom onset. % AK: Probably need a 50%?
+We assumed that asymptomatic infections can occur and infect others, but are less common that symptomatic infectious indivdiuals and that the majority of cases will develop symptoms during their infectious period.
+We ran three scenarios, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.
 
@@
 
-We use the same definition for outbreak control as \cite{hellewellFeasibilityControllingCOVID192020}: no new cases between 12 and 16 weeks after the start of an outbreak.
+We define outbreak control as no new cases between 12 and 16 weeks after the start of an outbreak\cite{hellewellFeasibilityControllingCOVID192020}.
 The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct.
+There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable within 12-16 weeks.
+We set this limit to 500 cases.
+To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial independent cases.
+This could represent importation of multiple cases in a single event, such as a flight with multiple positive cases.
 \\
 
-There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable within 12-16 weeks.
-We set this limit to 500 cases.
-This number needs to be larger than the expected cluster size from an subcritical outbreak produce by chance (i.e.
-larger than $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number).
-
-\subsection*{Pathogen epidemiological parameters} \label{epiparameters}
-
-In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previously caused sporadic human-to-human outbreaks with pandemic potential.
-We include A/H5N1, A/H1N1, and A/H7N9.
-These have caused outbreaks of various size and severity over the past 100 years.
-H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}.
-H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}.
-The first human case with the H7N9 subtype was in 2013, subsequently causing five outbreak waves, but no human cases since 2019 \citep{liH7N9InfluenzaVirus2021}.
-Most cases are of zoonotic origin with recorded animal exposure but there have been some small clusters (2 --- 3 cases) of possible human-to-human transmission \citep{liEpidemiologyHumanInfections2014}.
-For the purposes of this study we assume that there is no preexisting population immunity for these influenza subtypes.
-This is generally true for H5N1 and H7N9 at present, and for H1N1pdm09 prior to 2009.
-\\
-
-We extracted incubation periods from the epidemiological literature that have been previously estimated for the chosen Influenza subtypes.
-All incubation periods were found to be well described by a Weibull distribution, \cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}A).
-\cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed.
-For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}A).
-Comparatively, the incubation period for H1N1 is noticably shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}A).
-\\
-
-The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5.
-We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution (e.g.
-genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens.
-Current estimates for H5N1 and H7N9 subtypes are below unity (i.e.
-$R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention; with most outbreaks being small except in rare instances of superspreading events.
-Therefore evaluating isolation and contact tracing requires simulating outbreaks that have pandemic potential with exponential spread in the absence of control measures.
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/patho_param_space.png}
-\caption{The pathogen characteristics, in terms of tranmissibility ($R$) and proportion of disease transmission that occurs before the onset of symptoms (including cases that are entirely asymptomatic).
-The figure includes: Influenza, SARS-CoV-1, SARS-CoV-2, Smallpox, Ebola, Marburg, and HIV shown by green shaded areas.
-The size of the shaded area reflects the differences in pathogen characteristics, both between variants and outbreaks, the blurred edges reflects uncertainty in parameter estimates.
-The parameter space of transmissibility and pre- and asymptomatic transmission explored in this paper is shown in the orange box.}
-\label{fig:patho-param-space}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth,height=0.85\textheight]{../plots/delay_distributions.png}
-\caption{(A) Incubation period for the Influenza subtypes: H1N1, H5N1 and H7N9.
-(B) Three assumed onset-to-isolation delays used in the simuluation experiment.
-The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China \citep{liEarlyTransmissionDynamics2020}, the median onset-to-isolatiojn is 8.1 days (blue dashed line).
-The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a}, the median onset-to-isolation time is 3.4 days (red dashed line).
-The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days (orange dashed line).
-(C) Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skew-normal distribution.}
-\label{fig:delay-distributions}
-\end{figure}
-
-\clearpage
-
-\subsection*{Simulation scenarios}
-
-To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of isolation and contact tracing we set up a simulation experiment.
-We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmission, and number of initial cases.
-\\
-
-The ability to contain an outbreak by isolating infected individuals is highly contingent on the response time, or delay, between individuals becoming symptomatic, getting tested, the test results being returned and going into isolation.
-The probability that this series of steps will take an given amount of time is encapsulated in the onset-to-isolation delay distribution, i.e.
-onset-to-isolation distribution determines the time between a case developing symptoms and then becoming isolated.
-Here we define the onset-to-isolation as a response time as it is determined by the time delay between an individual responding to symptoms by getting tested and the turn around time for the tests to be analysed and the result returned.
-We view the onset-to-isolation as a feature of the public health outbreak response rather than an epidemiological characteristic of a disease.
-As such it is possible for pandemic response to adapt their response time to meet the needs of an outbreak.
-\\
-
-We assume that an isolated case cannot transmit, so isolation is assumed to be 100\% effective once enacted ($R_{isolated} = 0$).
 The onset-to-isolation delay is parameterised with three scenarios: a fast response, based on the SARS in Hong Kong (Donnelly et al.
@@
 Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:delay-distributions}B).
-These two empirically informed delay distributions provide a guide to possible response times in the case of an influenza outbreak and will enable us to evaluate if such delays are too long to contain a disease that can transmit rapidly such as influenza.
 We also include a third onset-to-isolation scenario based on the case where lateral flow rapid tests (LFTs) are provided to all individuals in the susceptible population and these people are freely able to self-test if they suspect they are symptomatic.
@@
 We assume that LFTs are 100\% accurate so all symptomatic individuals will receive a positive test result.
-\\
-
-We define the proportion of presymptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$).
-The shape parameter controls the proportion of presymptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:delay-distributions}C).
-We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:delay-distributions}C).
-Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective isolation and contact tracing is hypothesied to be.
-\\
-
-The proportion of pandemic-potential influenza infections that are asymptomatic varies for past outbreaks.
-There is evidence from past human outbreaks of H1N1pdm and H5N1 (2.3.4.4b) that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009, gargHighlyPathogenicAvian2025}; but whether human-to-human transmission can occur while subclinical is unknown.
-There is no evidence of asymptomatic transmission for the H7N9 subtype \citep{xuSerologicalInvestigationSubclinical2013}.
-Here we assume that asymptomatic cases occur and can infect others, but are uncommon and that the majority of cases will develop symptoms during their infectious period.
-We ran three scenarios, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.
-\\
-
-The feasibility of epidemic containment by contact tracing is expected to depend on the proportion of close contacts of cases that are traced and isolated if symptomatic.
-We varied the proportion of contacts ascertained in the model between 0 and 1, at intervals of 0.2.
-In cases when the proportion of contacts traced is zero, isolation after the delay from symptom onset is the only NPI reducing disease transmission.
-To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial indepedent cases.
-This could represent importation of multiple cases in a single event, such as a flight with multiple positive cases.
-\\
-
-We fixed some parameters across all simulations.
-The dispersion parameter for the negative binomial offspring distribution was fixed at 1 for isolated cases, and 0.16 for subclinical and community cases.
-For computational efficiency we capped the maximum number of days the simulation could run for to 365.
-This is substantially longer than the 12-16 week window for evaluating outbreak control so does not influence the results.
-Additionally, we limit simulations to 500 cases.
-This is an arbitrary threshold we judge to be large epidemic that is not controllable.
 
@@
 The most rapid scenario we simulated to emulate individuals self-testing with LFTs, has the same trend of increased chance of outbreak containment by reducing the time between symptom onset and isolation, as the slow and fast scenarios (Figure \ref{fig:prop-outbreak-control-R}).
-The contol of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C).
+The control of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C).
 Assuming ubiquitous access to home testing and that 63\% of symptomatic cases test and isolate within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$).
@@
 
+\section*{Supplementary Materials}
+
+\subsection*{Branching process model with isolation}
+
+To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transsmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}).
+The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (infinite population size, without depletion of susceptibility).
+Influenza subtypes have shown moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}, but evidence is limited and the negative binomial allows flexibility between highly heterogeneous and homogeneous individual-level transmission.
+The mean of the negative binomial offspring distribution is the effective reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}.
+\\
+
+There are three categories for individuals in the simulation, with each having their own parameterisation for the negative binomial offspring distribution: 1) \textit{community} (i.e.
+general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}.
+Each group have a distinct $R$ and dispersion ($k$) parameter due to the assumed differences in transmission dynamics between groups.
+\\
+
+% AK EDIT: Align with definitions in main methods
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/model-schematic.png}
+\caption{Model schematic of the branching process with isolation, quarantine and contact tracing implemented in the \texttt{\{ringbp\}} R package.
+(A) The transmission chain from the index case (Person A), infection events are shown by black vertical dashed lines.
+The time individuals are infected is shown by the red dot, the time of symptom onset is shown by the orange dot, and the time time individuals are isolated is shown by the blue dot.
+Quarantine is optional in the model, assuming quarantine is active it is shown by the small green dot with a Q.
+The time at whcih a infectors isolation triggers the isolation of a symtomatic case is shown by the blue vertical dashed line (Person A to Person B).
+The isolation times that would have been triggered after a delay from symptom onset, but are after isolation was triggered by contact tracing shown by a blue dot with a cross and connected by a dotted line (Person B).
+Person C is an asymptomatic individual (asymptomatic cases can infect others).
+Person D is missed by contact tracing (contacts are traced with a probability $\rho$, and thus missed at $1 - \rho$).
+The time Person D would have isolated if ascertained is shown by the vertical blue mark.
+Assuming quarantine is active, Person F would be isolated before symptom onset at the time of the infector's (Person D) isolation time.
+If quarantine is not active, then Person F would be isolated after a delay from symptom onset, as they had yet to show symptoms when the infector was isolated.
+(B) A branching process diagram showing who-infected-whom.}
+\label{fig:ringbp-model}
+\end{figure}
+
+The time between an infector being infected and then infecting an infectee (i.e.
+generation time) is computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
+This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, onset of symptoms, and onward transmission \citep{hellewellFeasibilityControllingCOVID192020}.
+This allows the proportion presymptomatic transmission to be specified, rather than defining incubation period and generation time distributions, which only indirectly allows assumptions of presymptomatic infectiousness to be derived.
+This simple formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is estimated but the generation time is estimated with more uncertainty or unknown, and the serial interval cannot be substituted in without introducing bias \citep{brittonEstimationEmergingEpidemics2019a, lehtinenRelationshipSerialInterval2021}.
+\\
+
+Given the incubation period for each case, the proportion of the infections that occur prior to, or after, the symptom onset of an infector is controlled by the parameterisation of the skew normal distribution (Figure \ref{fig:delay-distributions}C).
+The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed.
+The scale parameter ($\omega$) is fixed at 2, this parameter (with the shape parameter) defines the variance of the distribution.
+We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2).
+The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution).
+In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector.
+\\
+
+The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing (Figure \ref{fig:ringbp-model}).
+When a case is symptomatic they are isolated after a delay.
+The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation).
+Any infection times that are sampled to occur after the isolated of an infector are removed, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times.
+In other words, once a case is isolated at time $t_{isolated}$, the probability of that case infecting anyone at $t > t_{isolated}$ is zero.
+Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).
+Symptomatic individuals that are traced are isolated either when 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated.
+The isolation of the infectee is whatever occurs first of these two events.
+\\
+
+There are several mechanisms where isolation and contact tracing are imperfect at reducing onward disease transmission.
+The proportion of contacts missed by contact tracing is defined by the proportion of ascertainment and can range between zero, that is nobody is traced, to one, every case is traced.
+Symptomatic cases that have been missed when contact tracing their infectors are isolated at the onset-to-isolation delay after their symptom onset time.
+Additionally, contact tracing misses infections by an asymptomatic, or paucisymptomatic (subclinical), infector.
+In this event the infectees are missed by contact tracing owing to the infector not being known to the response system.
+\\
+
+\subsection*{Defining outbreak control}
+
+We use the same definition for outbreak control as \cite{hellewellFeasibilityControllingCOVID192020}: no new cases between 12 and 16 weeks after the start of an outbreak.
+The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct.
+\\
+
+There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable within 12-16 weeks.
+We set this limit to 500 cases.
+This number needs to be larger than the expected cluster size from an subcritical outbreak produce by chance (i.e.
+larger than $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number).
+
+\subsection*{Pathogen epidemiological parameters} \label{epiparameters}
+
+In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previously caused sporadic human-to-human outbreaks with pandemic potential.
+We include A/H5N1, A/H1N1, and A/H7N9.
+These have caused outbreaks of various size and severity over the past 100 years.
+H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}.
+H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}.
+The first human case with the H7N9 subtype was in 2013, subsequently causing five outbreak waves, but no human cases since 2019 \citep{liH7N9InfluenzaVirus2021}.
+Most cases are of zoonotic origin with recorded animal exposure but there have been some small clusters (2 --- 3 cases) of possible human-to-human transmission \citep{liEpidemiologyHumanInfections2014}.
+For the purposes of this study we assume that there is no preexisting population immunity for these influenza subtypes.
+This is generally true for H5N1 and H7N9 at present, and for H1N1pdm09 prior to 2009.
+\\
+
+We extracted incubation periods from the epidemiological literature that have been previously estimated for the chosen Influenza subtypes.
+All incubation periods were found to be well described by a Weibull distribution, \cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}A).
+\cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed.
+For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}A).
+Comparatively, the incubation period for H1N1 is noticably shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}A).
+\\
+
+The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5.
+We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution (e.g.
+genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens.
+Current estimates for H5N1 and H7N9 subtypes are below unity (i.e.
+$R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention; with most outbreaks being small except in rare instances of superspreading events.
+Therefore evaluating isolation and contact tracing requires simulating outbreaks that have pandemic potential with exponential spread in the absence of control measures.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/patho_param_space.png}
+\caption{The pathogen characteristics, in terms of tranmissibility ($R$) and proportion of disease transmission that occurs before the onset of symptoms (including cases that are entirely asymptomatic).
+The figure includes: Influenza, SARS-CoV-1, SARS-CoV-2, Smallpox, Ebola, Marburg, and HIV shown by green shaded areas.
+The size of the shaded area reflects the differences in pathogen characteristics, both between variants and outbreaks, the blurred edges reflects uncertainty in parameter estimates.
+The parameter space of transmissibility and pre- and asymptomatic transmission explored in this paper is shown in the orange box.}
+\label{fig:patho-param-space}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth,height=0.85\textheight]{../plots/delay_distributions.png}
+\caption{(A) Incubation period for the Influenza subtypes: H1N1, H5N1 and H7N9.
+(B) Three assumed onset-to-isolation delays used in the simuluation experiment.
+The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China \citep{liEarlyTransmissionDynamics2020}, the median onset-to-isolatiojn is 8.1 days (blue dashed line).
+The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a}, the median onset-to-isolation time is 3.4 days (red dashed line).
+The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days (orange dashed line).
+(C) Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skew-normal distribution.}
+\label{fig:delay-distributions}
+\end{figure}
+
+\clearpage
+
+\subsection*{Simulation scenarios}
+
+To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of isolation and contact tracing we set up a simulation experiment.
+We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmission, and number of initial cases.
+\\
+
+The ability to contain an outbreak by isolating infected individuals is highly contingent on the response time, or delay, between individuals becoming symptomatic, getting tested, the test results being returned and going into isolation.
+The probability that this series of steps will take an given amount of time is encapsulated in the onset-to-isolation delay distribution, i.e.
+onset-to-isolation distribution determines the time between a case developing symptoms and then becoming isolated.
+Here we define the onset-to-isolation as a response time as it is determined by the time delay between an individual responding to symptoms by getting tested and the turn around time for the tests to be analysed and the result returned.
+We view the onset-to-isolation as a feature of the public health outbreak response rather than an epidemiological characteristic of a disease.
+As such it is possible for pandemic response to adapt their response time to meet the needs of an outbreak.
+\\
+
+We assume that an isolated case cannot transmit, so isolation is assumed to be 100\% effective once enacted ($R_{isolated} = 0$).
+The onset-to-isolation delay is parameterised with three scenarios: a fast response, based on the SARS in Hong Kong (Donnelly et al.
+2003) and a slower response based on COVID-19 in Wuhan, China (Li et al.
+2020) (Figure \ref{fig:delay-distributions}B).
+Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:delay-distributions}B).
+These two empirically informed delay distributions provide a guide to possible response times in the case of an influenza outbreak and will enable us to evaluate if such delays are too long to contain a disease that can transmit rapidly such as influenza.
+We also include a third onset-to-isolation scenario based on the case where lateral flow rapid tests (LFTs) are provided to all individuals in the susceptible population and these people are freely able to self-test if they suspect they are symptomatic.
+We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test).
+We assume that LFTs are 100\% accurate so all symptomatic individuals will receive a positive test result.
+\\
+
+We define the proportion of presymptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$).
+The shape parameter controls the proportion of presymptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:delay-distributions}C).
+We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:delay-distributions}C).
+Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective isolation and contact tracing is hypothesied to be.
+\\
+
+The proportion of pandemic-potential influenza infections that are asymptomatic varies for past outbreaks.
+There is evidence from past human outbreaks of H1N1pdm and H5N1 (2.3.4.4b) that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009, gargHighlyPathogenicAvian2025}; but whether human-to-human transmission can occur while subclinical is unknown.
+There is no evidence of asymptomatic transmission for the H7N9 subtype \citep{xuSerologicalInvestigationSubclinical2013}.
+Here we assume that asymptomatic cases occur and can infect others, but are uncommon and that the majority of cases will develop symptoms during their infectious period.
+We ran three scenarios, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.
+\\
+
+The feasibility of epidemic containment by contact tracing is expected to depend on the proportion of close contacts of cases that are traced and isolated if symptomatic.
+We varied the proportion of contacts ascertained in the model between 0 and 1, at intervals of 0.2.
+In cases when the proportion of contacts traced is zero, isolation after the delay from symptom onset is the only NPI reducing disease transmission.
+To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial indepedent cases.
+This could represent importation of multiple cases in a single event, such as a flight with multiple positive cases.
+\\
+
+We fixed some parameters across all simulations.
+The dispersion parameter for the negative binomial offspring distribution was fixed at 1 for isolated cases, and 0.16 for subclinical and community cases.
+For computational efficiency we capped the maximum number of days the simulation could run for to 365.
+This is substantially longer than the 12-16 week window for evaluating outbreak control so does not influence the results.
+Additionally, we limit simulations to 500 cases.
+This is an arbitrary threshold we judge to be large epidemic that is not controllable.
+
 \end{document}

### 2025-06-19 inst/paper/main.tex
Add discussion points

@@
 We also did not consider zoonotic spillover infections or formite transmission.
-\item Digital tracing has capacity for large-scale contact tracing (pingdemic), also rapid tests are a complementary approach for limiting spread.
+\item Digital tracing has capacity for large-scale contact tracing, also rapid tests are a complementary approach for limiting spread.
 \item In this study we model isolation and contact tracing as the sole intervention in containing an influenza outbreak.
@@
 oseltamivir), rapid tests, contact tracing, and vaccination.
-\item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic.
-\item In this study we have parameterised the transmission model with parameter from avian influenza subtypes (H5N1 and H7N9) and a influenza subtype that caused an epidemic in the recent past (H1N1), and response delays from SARS and COVID-19 outbreaks.
+\item Mention strengths: interpretability, focus on average impact, speed
+\item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic -- possibly beyond scope if focused on burden.
+\item In this study we have parameterised the transmission model with parameter from avian influenza subtypes (H5N1 and H7N9) and an influenza subtype that caused an epidemic in the recent past (H1N1), and response delays from SARS and COVID-19 outbreaks.
 However, these results can be used to inform other influenza subtypes that exhibit similar epidemiological characteristics in transmissibility and severity from past outbreaks (H2N2 and H3N2) and still exist in animal reservoirs for potential spillovers back into human populations (ref).
@@
 \item The reproduction number for H5N1 (including clade 2.3.4.4b) and H7N9 influenza subtypes have been estimated to be far below 1 (Ward et al., Garg et al., 2025).
-Therefore, the transmissibility scenarios explored in this study are speculative and assume that pathogen evolution increases human-to-human transmission potential.
+Therefore, the transmissibility scenarios explored in this study are mostly focused on planning purposes.
 If transmissibility of these avian influenza subtypes remains constant over time then outbreaks are expected to be subcritical and will not produce epidemics and thus not require interventions.
@@
 Infection severity and fatality risks are of course critical when formulating a pandemic response plan.
-By understanding the efficacy of isolation and contact trace for a given epidemiological characterstic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.
+By understanding the efficacy of isolation and contact trace for a given epidemiological characteristic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.
 \item Our branching process model used a simple forward contact-tracing mechanism that reduces the number of secondary contacts based on the timing of isolation of symptomatic cases.

### 2025-06-19 inst/paper/main.tex
Update main.tex
Add conclusion

@@
 oseltamivir), rapid tests, contact tracing, and vaccination.
-\item Mention strengths: interpretability, focus on average impact, speed
+\item Mention strengths: interpretability, focus on average impact, speed. R captures distribution of secondary transmission in population in concise way.
 \item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic -- possibly beyond scope if focused on burden.
@@
 \item Isolation and contact tracing complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals \citep{haydenPerspectivesAntiviralUse2001}.
-\item This paper has focused on forward contact tracing.
+\item Focus more on punchline, i.e.
+should not assume contact tracing useless for flu given current tools, and intermediate emergence characteristics (i.e.
+slower spread) may offer larger opportunity for window of control.
+%\item This paper has focused on forward contact tracing.
 Backward contact tracing, whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source, is applicable to infectious with long infectious periods (e.g.
 HIV), but is not effective for more ephemeral infections with faster transmission, as is the case for influenza.
-\item Contact tracing and isolation is a targeted approach analogous to ring vaccination.
+%\item Contact tracing and isolation is a targeted approach analogous to ring vaccination.
 It looks to prevent onward transmission by, either isolating or quarantining contacts in the case of contact tracing, or vaccinating contacts in the case of ring vaccination \citep{kucharskiEffectivenessRingVaccination2016, whittakerQuantifyingImpactBroadly2024}.

### 2025-06-19 inst/scripts/run_analysis.R
update run_analysis.R script to use the newest version of {ringbp}

@@ -1,8 +1,8 @@
 library(ringbp)
-library(purrr)
-library(future)
 library(data.table)
 library(epiparameter)
+library(future)
+library(future.apply)
 
 h5n1_weibull_params <- epiparameter::convert_summary_stats_to_params(
   "weibull", mean = 3.3, sd = 1.5
@@ -17,36 +17,38 @@ scenarios <- data.table(
     delay_group = list(data.table(
       delay = c("fast", "slow", "lft"),
       onset_to_isolation = c(
-        \(x) stats::rweibull(n = x, shape = 1.651524, scale = 4.287786),
-        \(x) stats::rweibull(n = x, shape = 2.305172, scale = 9.483875),
-        \(x) stats::rexp(n = x, rate = 0.5)
+        \(n) stats::rweibull(n = n, shape = 1.651524, scale = 4.287786),
+        \(n) stats::rweibull(n = n, shape = 2.305172, scale = 9.483875),
+        \(n) stats::rexp(n = n, rate = 0.5)
       )
-    )),
-    k_group = list(data.table(
-      theta = c("<1%", "15%", "30%"),
-      k = c(30, 1.95, 0.7)
     )),
     incubation_period_group = list(data.table(
       subtype = c("H1N1", "H5N1", "H7N9"),
       incubation_period = c(
-        \(x) stats::rweibull(n = x, shape = 1.77, scale = 1.86),
-        \(x) stats::rweibull(
-          n = x,
+        \(n) stats::rweibull(n = n, shape = 1.77, scale = 1.86),
+        \(n) stats::rweibull(
+          n = n,
           shape = h5n1_weibull_params$shape,
           scale = h5n1_weibull_params$scale
         ),
-        \(x) stats::rweibull(
-          n = x,
+        \(n) stats::rweibull(
+          n = n,
           shape = h7n9_weibull_params$shape,
           scale = h7n9_weibull_params$scale
         )
       )
     )),
-    index_R0 = c(1.1, 1.5, 2.5, 3.5),
-    prop.asym = c(0, 0.1, 0.3),
-    control_effectiveness = seq(0, 1, 0.2),
-    num.initial.cases = c(5, 20, 40),
-    quarantine = FALSE
+    r0_community = c(1.1, 1.5, 2.5, 3.5),
+    r0_isolated = 0,
+    disp_community = 0.16,
+    disp_isolated = 1,
[…]

### 2025-06-19 inst/bash/run_analysis.sh
add run_analysis.sh bash script

@@ -1,0 +1,15 @@
+#!/bin/bash
+#SBATCH --job-name=FluTracer
+#SBATCH --ntasks=1
+#SBATCH --mem=5GB
+#SBATCH --time=01:00:00
+#SBATCH --output=FluTracer.log
+pwd; hostname; date
+
+echo "Running FluTracer analysis script"
+
+module load R
+
+Rscript inst/scripts/run_analysis.R
+
+date

### 2025-06-19 inst/bash/install_FluTracer.sh
add install_FluTracer.sh bash script

@@ -1,0 +1,10 @@
+#!/bin/bash
+#SBATCH --job-name=install_FluTracer
+#SBATCH --ntasks=1
+#SBATCH --mem=5GB
+#SBATCH --time=01:00:00
+#SBATCH --output=install_FluTracer.log
+pwd; hostname; date
+
+ml R
+Rscript -e "remotes::install_github('joshwlambert/FluTracer')"

### 2025-06-19 inst/bash/install_FluTracer.sh
module load R4.4.0 on HPC

@@ -6,5 +6,5 @@
 #SBATCH --output=install_FluTracer.log
 pwd; hostname; date
 
-ml R
+ml R/4.4.0
 Rscript -e "remotes::install_github('joshwlambert/FluTracer')"

### 2025-06-19 inst/bash/run_analysis.sh

@@ -8,7 +8,7 @@ pwd; hostname; date
 
 echo "Running FluTracer analysis script"
 
-module load R
+module load R/4.4.0
 
 Rscript inst/scripts/run_analysis.R
 

### 2025-06-19 inst/bash/run_analysis.sh
try running scenarios in parallel on HPC

@@ -1,6 +1,7 @@
 #!/bin/bash
 #SBATCH --job-name=FluTracer
-#SBATCH --ntasks=1
+#SBATCH --ntasks =8
+#SBATCH --nodes=1
 #SBATCH --mem=5GB
 #SBATCH --time=01:00:00
 #SBATCH --output=FluTracer.log
@@ -9,6 +10,10 @@ pwd; hostname; date
 echo "Running FluTracer analysis script"
 
 module load R/4.4.0
+
+echo "Number of cores available: "
+
+Rscript -e "future::availableCores()"
 
 Rscript inst/scripts/run_analysis.R
 

### 2025-06-19 inst/scripts/run_analysis.R

@@ -69,10 +69,10 @@ n <- 1
 
 # Set up multicore if using see ?future::plan for details
 # Use the workers argument to control the number of cores used.
-# future::plan("multisession", workers = 7)
+future::plan("multicore", workers = 8)
 
 # Run parameter sweep
-scenario_sims[, sims := lapply(data, \(x, n) {
+scenario_sims[, sims := future_lapply(data, \(x, n) {
   scenario_sim(
     n = n,
     initial_cases = x$initial_cases,

### 2025-06-19 inst/bash/run_analysis.sh
rm whitespace from run_analysis.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 #SBATCH --job-name=FluTracer
-#SBATCH --ntasks =8
+#SBATCH --ntasks=8
 #SBATCH --nodes=1
 #SBATCH --mem=5GB
 #SBATCH --time=01:00:00

### 2025-06-19 inst/bash/run_analysis.sh
increase number of tasks in run_analysis.sh and use future.seed = TRUE in run_analysis.R

@@ -1,6 +1,6 @@
 #!/bin/bash
 #SBATCH --job-name=FluTracer
-#SBATCH --ntasks=8
+#SBATCH --ntasks=16
 #SBATCH --nodes=1
 #SBATCH --mem=5GB
 #SBATCH --time=01:00:00

### 2025-06-19 inst/scripts/run_analysis.R

@@ -69,7 +69,7 @@ n <- 1
 
 # Set up multicore if using see ?future::plan for details
 # Use the workers argument to control the number of cores used.
-future::plan("multicore", workers = 8)
+future::plan("multicore", workers = 16)
 
 # Run parameter sweep
 scenario_sims[, sims := future_lapply(data, \(x, n) {
@@ -93,7 +93,8 @@ scenario_sims[, sims := future_lapply(data, \(x, n) {
     sim = sim_opts(cap_max_days = x$cap_max_days, cap_cases = x$cap_cases)
   )
 },
-n = n
+n = n,
+future.seed = TRUE
 )]
 
 saveRDS(scenario_sims, file = file.path("inst", "extdata", "pilot_simulations.rds"))

### 2025-06-19 inst/scripts/run_analysis.R
increase scenario_sim replicates to 10 in run_analysis.R

@@ -65,7 +65,7 @@ scenarios <- merge(
 scenarios[, scenario :=  1:.N]
 scenario_sims <- scenarios[, list(data = list(.SD)), by = scenario]
 
-n <- 1
+n <- 10
 
 # Set up multicore if using see ?future::plan for details
 # Use the workers argument to control the number of cores used.

### 2025-07-02 inst/scripts/plot_delay_distributions.R
update plotting order in plot_delay_distributions.R

@@ -4,9 +4,9 @@ source(file.path("inst", "scripts", "plot_incubation_period.R"))
 source(file.path("inst", "scripts", "plot_onset_to_isolation.R"))
 source(file.path("inst", "scripts", "plot_prop_presymptomatic_transmission.R"))
 
-delay_distributions_plot <- incubation_period_plot /
-  onset_to_isolation_plot /
-  prop_presymptomatic_transmission_plot +
+delay_distributions_plot <- prop_presymptomatic_transmission_plot /
+  incubation_period_plot /
+  onset_to_isolation_plot +
   plot_annotation(tag_levels = 'A')
 
 ggplot2::ggsave(

### 2025-07-02 inst/scripts/plot_prop_outbreak_control_reproduction_number.R
update plot_prop_outbreak_control_reproduction_number script and plot

@@ -2,13 +2,13 @@ library(data.table)
 library(ggplot2)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
-h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
-h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
-h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+h5n1_results <- readRDS(file.path("inst", "extdata", "H5N1_simulations_no_Q.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "H1N1_simulations_no_Q.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "H7N9_simulations_no_Q.rds"))
 
-h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
-h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
-h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
 
 h5n1_data <- rbindlist(h5n1_results$data)
 h5n1_data[, `:=`(scenario = h5n1_results$scenario, pext = h5n1_results$pext)]
@@ -26,12 +26,12 @@ rm(h7n9_results)
 flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
 
 prop_outbreak_control <- flu_data[
-  num.initial.cases == 20 & theta == "15%" & prop.asym == 0,
-  .(control_effectiveness, index_R0, pext, subtype, delay)
+  initial_cases == 20 & prop_presymptomatic == 0.15 & prop_asymptomatic == 0,
+  .(prop_ascertain, r0_community, pext, subtype, delay)
 ]
 
 # convert to percentages for plotting
-prop_outbreak_control[, control_effectiveness := control_effectiveness * 100]
+prop_outbreak_control[, prop_ascertain := prop_ascertain * 100]
 prop_outbreak_control[, pext := pext * 100]
 
[…]

### 2025-07-02 inst/scripts/plot_prop_outbreak_control_var_reproduction_number.R
update prop_outbreak_control_var_reproduction_number script and plot

@@ -1,27 +1,38 @@
 library(data.table)
 library(ggplot2)
-library(dplyr)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
-sweep_results <- readRDS(file.path("inst", "extdata", "simulations.rds"))
+h5n1_results <- readRDS(file.path("inst", "extdata", "H5N1_simulations_no_Q.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "H1N1_simulations_no_Q.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "H7N9_simulations_no_Q.rds"))
 
-res <- sweep_results %>%
-  group_by(scenario) %>%
-  mutate(pext = ringbp::extinct_prob(sims[[1]], cap_cases = 500)) %>%
-  ungroup(scenario)
+h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
 
-dt <- as.data.table(res)
-dt_data <- rbindlist(dt$data)
-dt_data <- cbind(dt_data, scenario = dt$scenario, pext = dt$pext)
+h5n1_data <- rbindlist(h5n1_results$data)
+h5n1_data[, `:=`(scenario = h5n1_results$scenario, pext = h5n1_results$pext)]
+
+h1n1_data <- rbindlist(h1n1_results$data)
+h1n1_data[, `:=`(scenario = h1n1_results$scenario, pext = h1n1_results$pext)]
+
+h7n9_data <- rbindlist(h7n9_results$data)
+h7n9_data[, `:=`(scenario = h7n9_results$scenario, pext = h7n9_results$pext)]
+
+rm(h5n1_results)
+rm(h1n1_results)
+rm(h7n9_results)
+
+flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
 
 # convert to percentages for plotting
-dt_data[, control_effectiveness := control_effectiveness * 100]
-dt_data[, pext := pext * 100]
+flu_data[, prop_ascertain := prop_ascertain * 100]
+flu_data[, pext := pext * 100]
 
 # calculate sd by subtype, R and control
-subtype_sd <- dt_data[
+subtype_sd <- flu_data[
[…]

### 2025-07-02 inst/scripts/plot_prop_outbreak_control_num_init_cases.R
update prop_outbreak_control_num_init_cases script and plot

@@ -2,13 +2,13 @@ library(ggplot2)
 library(data.table)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
-h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
-h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
-h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+h5n1_results <- readRDS(file.path("inst", "extdata", "H5N1_simulations_no_Q.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "H1N1_simulations_no_Q.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "H7N9_simulations_no_Q.rds"))
 
-h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
-h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
-h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
 
 h5n1_data <- rbindlist(h5n1_results$data)
 h5n1_data[, `:=`(scenario = h5n1_results$scenario, pext = h5n1_results$pext)]
@@ -26,12 +26,12 @@ rm(h7n9_results)
 flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
 
 prop_outbreak_control <- flu_data[
-  theta == "15%" & prop.asym == 0 & delay == "fast",
-  .(control_effectiveness, index_R0, num.initial.cases, pext, subtype)
+  prop_presymptomatic == 0.15 & prop_asymptomatic == 0 & delay == "fast",
+  .(prop_ascertain, r0_community, initial_cases, pext, subtype)
 ]
 
 # convert to percentages for plotting
-prop_outbreak_control[, control_effectiveness := control_effectiveness * 100]
[…]

### 2025-07-02 inst/scripts/plot_prop_presymptomatic_transmission.R
update prop_presymptomatic_transmission script and plot

@@ -1,3 +1,4 @@
+library(ringbp)
 library(data.table)
 library(ggplot2)
 library(sn)
@@ -7,20 +8,29 @@ density_eval <- seq(0, 15, 0.1)
 prop_presymptomatic_transmission <- data.table(
   x = density_eval,
   prop_presym_trans_30 = sn::dsn(
-    x = density_eval, xi = 5, omega = 2, alpha = 0.7
+    x = density_eval,
+    xi = 5,
+    omega = 2,
+    alpha = ringbp:::presymptomatic_transmission_to_alpha(presymptomatic_transmission = 0.3)
   ),
   prop_presym_trans_15 = sn::dsn(
-    x = density_eval, xi = 5, omega = 2, alpha = 1.95
+    x = density_eval,
+    xi = 5,
+    omega = 2,
+    alpha = ringbp:::presymptomatic_transmission_to_alpha(presymptomatic_transmission = 0.15)
   ),
-  prop_presym_trans_lt1 = sn::dsn(
-    x = density_eval, xi = 5, omega = 2, alpha = 30
+  prop_presym_trans_1 = sn::dsn(
+    x = density_eval,
+    xi = 5,
+    omega = 2,
+    alpha = ringbp:::presymptomatic_transmission_to_alpha(presymptomatic_transmission = 0.01)
   )
 )
 
 prop_presymptomatic_transmission <- melt(
   data = prop_presymptomatic_transmission,
   id.vars = "x",
-  measure.vars = c("prop_presym_trans_30", "prop_presym_trans_15", "prop_presym_trans_lt1"),
+  measure.vars = c("prop_presym_trans_30", "prop_presym_trans_15", "prop_presym_trans_1"),
   variable.name = "prop",
   value.name = "prop_presym_trans"
 )
@@ -53,9 +63,18 @@ prop_presymptomatic_transmission_plot <- ggplot2::ggplot(
   ) +
   ggplot2::scale_x_continuous(name = "Days since infection") +
   ggplot2::scale_y_continuous(name = "Density") +
-  ggplot2::labs(colour = "Proportion presymptomatic transmission", fill = "Proportion presymptomatic transmission") +
-  ggplot2::scale_colour_manual(values = c("#009e74", "#56b3e9", "#cc79a7"), labels = c("30%", "15%", "<1%")) +
[…]

### 2025-07-02 inst/scripts/plot_prop_outbreak_control_prop_asym_iso.R
update prop_outbreak_control_prop_asym_iso and prop_outbreak_control_prop_presym_iso scripts and plots

@@ -1,15 +1,14 @@
 library(data.table)
 library(ggplot2)
-library(dplyr)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
-h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
-h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
-h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations_no_Q.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations_no_Q.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations_no_Q.rds"))
 
-h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
-h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
-h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
 
 h5n1_data <- rbindlist(h5n1_results$data)
 h5n1_data[, `:=`(scenario = h5n1_results$scenario, pext = h5n1_results$pext)]
@@ -27,12 +26,12 @@ rm(h7n9_results)
 flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
 
 prop_outbreak_control <- flu_data[
-  num.initial.cases == 20 & theta == "15%",
-  .(control_effectiveness, index_R0, pext, subtype, delay, prop.asym)
+  initial_cases == 20 & prop_presymptomatic == 0.15,
+  .(prop_ascertain, r0_community, pext, subtype, delay, prop_asymptomatic)
 ]
 
 # convert to percentages for plotting
-prop_outbreak_control[, control_effectiveness := control_effectiveness * 100]
+prop_outbreak_control[, prop_ascertain := prop_ascertain * 100]
[…]

### 2025-07-02 inst/scripts/plot_prop_outbreak_control_prop_presym_iso.R

@@ -1,15 +1,14 @@
 library(data.table)
 library(ggplot2)
-library(dplyr)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
-h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
-h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
-h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations_no_Q.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations_no_Q.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations_no_Q.rds"))
 
-h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
-h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
-h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 500), by = scenario]
+h5n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h1n1_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
+h7n9_results[, pext := ringbp::extinct_prob(sims[[1]], cap_cases = 5000), by = scenario]
 
 h5n1_data <- rbindlist(h5n1_results$data)
 h5n1_data[, `:=`(scenario = h5n1_results$scenario, pext = h5n1_results$pext)]
@@ -27,50 +26,52 @@ rm(h7n9_results)
 flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
 
 prop_outbreak_control <- flu_data[
-  num.initial.cases == 20 & prop.asym == 0.1,
-  .(control_effectiveness, index_R0, pext, subtype, delay, theta)
+  initial_cases == 20 & prop_asymptomatic == 0.1,
+  .(prop_ascertain, r0_community, pext, subtype, delay, prop_presymptomatic)
 ]
 
 # convert to percentages for plotting
-prop_outbreak_control[, control_effectiveness := control_effectiveness * 100]
+prop_outbreak_control[, prop_ascertain := prop_ascertain * 100]
[…]

### 2025-07-02 inst/scripts/plot_controlled_outbreak.R
update max_weekly_cases and median_controlled_outbreak_size scripts and plots

@@ -2,75 +2,75 @@ library(data.table)
 library(ggplot2)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
-h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
-h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
-h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+h5n1_results <- readRDS(file.path("inst", "extdata", "H5N1_simulations_no_Q.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "H1N1_simulations_no_Q.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "H7N9_simulations_no_Q.rds"))
 
 h5n1_total_cases <- lapply(h5n1_results$sims, function(x) {
   total_cases <- x[, .SD[which.max(week)], by = sim][, cumulative]
-  total_cases[total_cases > 500] <- 500
+  total_cases[total_cases > 5000] <- 5000
   return(total_cases)
 })
 
 h1n1_total_cases <- lapply(h1n1_results$sims, function(x) {
   total_cases <- x[, .SD[which.max(week)], by = sim][, cumulative]
-  total_cases[total_cases > 500] <- 500
+  total_cases[total_cases > 5000] <- 5000
   return(total_cases)
 })
 
 h7n9_total_cases <- lapply(h7n9_results$sims, function(x) {
   total_cases <- x[, .SD[which.max(week)], by = sim][, cumulative]
-  total_cases[total_cases > 500] <- 500
+  total_cases[total_cases > 5000] <- 5000
   return(total_cases)
 })
 
 median_h5n1_controlled_outbreak_size <- vapply(
   h5n1_total_cases,
-  function(x) median(x[x != 500]),
+  function(x) median(x[x != 5000]),
   FUN.VALUE = numeric(1)
 )
 
 median_h1n1_controlled_outbreak_size <- vapply(
   h1n1_total_cases,
-  function(x) median(x[x != 500]),
+  function(x) median(x[x != 5000]),
   FUN.VALUE = numeric(1)
 )
 
 median_h7n9_controlled_outbreak_size <- vapply(
   h7n9_total_cases,
-  function(x) median(x[x != 500]),
+  function(x) median(x[x != 5000]),
   FUN.VALUE = numeric(1)
 )
 
[…]

### 2025-07-02 inst/scripts/plot_max_weekly_cases.R

@@ -2,9 +2,9 @@ library(data.table)
 library(ggplot2)
 
 # read simulation results saved by running inst/scripts/run_analysis.R
-h5n1_results <- readRDS(file.path("inst", "extdata", "h5n1_simulations.rds"))
-h1n1_results <- readRDS(file.path("inst", "extdata", "h1n1_simulations.rds"))
-h7n9_results <- readRDS(file.path("inst", "extdata", "h7n9_simulations.rds"))
+h5n1_results <- readRDS(file.path("inst", "extdata", "H5N1_simulations_no_Q.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "H1N1_simulations_no_Q.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "H7N9_simulations_no_Q.rds"))
 
 h5n1_median_max_weekly_cases <- vapply(
   h5n1_results$sims,
@@ -30,15 +30,15 @@ h7n9_median_max_weekly_cases <- vapply(
   FUN.VALUE = numeric(1)
 )
 
-h5n1_max_weekly_cases <- rbindlist(h5n1_results$data)[, c("index_R0", "control_effectiveness", "delay")]
+h5n1_max_weekly_cases <- rbindlist(h5n1_results$data)[, c("r0_community", "prop_ascertain", "delay")]
 h5n1_max_weekly_cases[, max_weekly_cases := h5n1_median_max_weekly_cases]
 h5n1_max_weekly_cases[, subtype := as.factor("H5N1")]
 
-h1n1_max_weekly_cases <- rbindlist(h1n1_results$data)[, c("index_R0", "control_effectiveness", "delay")]
+h1n1_max_weekly_cases <- rbindlist(h1n1_results$data)[, c("r0_community", "prop_ascertain", "delay")]
 h1n1_max_weekly_cases[, max_weekly_cases := h1n1_median_max_weekly_cases]
 h1n1_max_weekly_cases[, subtype := as.factor("H1N1")]
 
-h7n9_max_weekly_cases <- rbindlist(h7n9_results$data)[, c("index_R0", "control_effectiveness", "delay")]
+h7n9_max_weekly_cases <- rbindlist(h7n9_results$data)[, c("r0_community", "prop_ascertain", "delay")]
 h7n9_max_weekly_cases[, max_weekly_cases := h7n9_median_max_weekly_cases]
 h7n9_max_weekly_cases[, subtype := as.factor("H7N9")]
 
[…]

### 2025-07-02 inst/scripts/plot_r0.R
add plot_r0 script and r0_high_presym_asym and r0_low_presym_asym plots

@@ -1,0 +1,163 @@
+library(data.table)
+library(ggplot2)
+
+# read simulation results saved by running inst/scripts/run_r0_analysis.R
+h5n1_results <- readRDS(file.path("inst", "extdata", "H5N1_r0_analysis_no_Q.rds"))
+h1n1_results <- readRDS(file.path("inst", "extdata", "H1N1_r0_analysis_no_Q.rds"))
+h7n9_results <- readRDS(file.path("inst", "extdata", "H7N9_r0_analysis_no_Q.rds"))
+
+h5n1_Q_results <- readRDS(file.path("inst", "extdata", "H5N1_r0_analysis_Q.rds"))
+h1n1_Q_results <- readRDS(file.path("inst", "extdata", "H1N1_r0_analysis_Q.rds"))
+h7n9_Q_results <- readRDS(file.path("inst", "extdata", "H7N9_r0_analysis_Q.rds"))
+
+h5n1_results <- rbindlist(list(h5n1_results, h5n1_Q_results))
+h1n1_results <- rbindlist(list(h1n1_results, h1n1_Q_results))
+h7n9_results <- rbindlist(list(h7n9_results, h7n9_Q_results))
+
+rm(h5n1_Q_results)
+rm(h1n1_Q_results)
+rm(h7n9_Q_results)
+
+# the effective R0 is the same for all weeks within each replicate so getting
+# from week 20 is the same as getting from any other week
+h5n1_median_r0 <- sapply(h5n1_results$sims, function(x) median(x[week == 20, effective_r0]))
+h1n1_median_r0 <- sapply(h1n1_results$sims, function(x) median(x[week == 20, effective_r0]))
+h7n9_median_r0 <- sapply(h7n9_results$sims, function(x) median(x[week == 20, effective_r0]))
+
+h5n1_data <- rbindlist(h5n1_results$data)
+h1n1_data <- rbindlist(h1n1_results$data)
+h7n9_data <- rbindlist(h7n9_results$data)
+
+rm(h5n1_results)
+rm(h1n1_results)
+rm(h7n9_results)
+
+h5n1_data[, median_r0 := h5n1_median_r0]
+h1n1_data[, median_r0 := h1n1_median_r0]
+h7n9_data[, median_r0 := h7n9_median_r0]
+
+flu_data <- rbindlist(list(h5n1_data, h1n1_data, h7n9_data))
+
+# convert to percentages for plotting
+flu_data[, prop_ascertain := prop_ascertain * 100]
+
[…]

### 2025-07-02 inst/paper/main.tex
update paper .tex and .pdf and FluTracer.bib

--- 2025-06-19 inst/paper/main.tex
@@
 \title{Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study}
-\author{Joshua W. Lambert, Authors TBC, Adam J. Kucharski}
+\author{Joshua W. Lambert, Adam J. Kucharski}
 \date{}
@@
 Influenza subtypes to which populations have low levels of pre-existing immunity are a particular concern as potential pandemic candidates.
-This includes influenza A/H5N1 \citep{Ward2024.12.11.24318702}, A/H7N9 \citep{tannerPandemicPotentialAvian2015}, and A/H1N1p when it caused caused a pandemic in 2009 \citep{fraserPandemicPotentialStrain2009}.
-Recent human cases of avian influenza A/H5N1, alongside transmission, further illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption [CITE: Antia et al, Nature, 2003 + key papers on H5N1 - see my PLOS Bio piece].
-It is therefore crucial to understand the effectiveness and feasibility of different intervention strategies in the face of such infections.
-
-%Other pathogens, including those less transmissible, with longer delays between infections, and with less presymptomatic transmission, can also cause sustained outbreaks leading to the excess death of thousands \citep{whoebolaresponseteamEbolaVirusDisease2014}.
-Timely and accurate response is therefore critical early in an outbreak when the pandemic or epidemic potential is still uncertain \citep{kucharskiControllingMinorOutbreaks2024}.
-An effective response requires understanding the efficacy and feasibility of intervention strategies across pathogens with varying epidemiological characteristics \citep{fraserFactorsThatMake2004}.
-\\ AK: Removed to improve flow, as this is mentioned in more detail later in intro.
-
-%The likelihood of a novel outbreak causing sustained transmission depends both the characteristics of the pathogen and its host population, as well control measures  implemented in response to early cases.
-A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases by a typical infectious host.
-If the reproduction number is greater than 1, we would expect the incidence of infections to increase over time.
-If $R$ is below 1, we would expect incidence to decline on average, but several cases could still occur before transmission ceases \citep{farringtonDistributionTimeExtinction1999}.
-Two additional factors can influence the probability a pathogen with $R$ above 1 will successfully establish within a population~\citep{lloyd-smithSuperspreadingEffectIndividual2005}.
-If there is a large degree of individual-level variation in transmission, the majority of infectious individuals will infect nobody, even though a minority will infect many; as a result, most newly introduced outbreaks will fail to take off.
-In addition, the likelihood of establishment increases if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}.
-\\ % AK: Note 'primary case' means first in outbreak
+This includes influenza A/H5N1 \citep{Ward2024.12.11.24318702} and A/H7N9 \citep{tannerPandemicPotentialAvian2015}, as well as A/H1N1pdm when it caused a pandemic in 2009 \citep{fraserPandemicPotentialStrain2009}.
+Recent human cases of avian influenza A/H5N1, alongside transmission, further illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption \citep{antiaRoleEvolutionEmergence2003, gargHighlyPathogenicAvian2025, peacockGlobalH5N1Influenza2025}.
+It is therefore crucial to understand the effectiveness and feasibility of different intervention strategies in the face of such infections \citep{fraserFactorsThatMake2004, kucharskiControllingMinorOutbreaks2024}.
+\\
 
 Early in a novel outbreak, pharmaceutical options such as treatments and vaccines may not yet be available, leading to reliance on non-pharmaceutical interventions (NPIs) to reduce infection and disease.
-These interventions include individual-level targeted measures such as isolation and contact tracing, to population-level policies such as venue closures or restrictions on social mixing (Huag et al., 2020) [AK: probably a better reference out there, e.g.
-Sharma et al, Nature Comms].
+These interventions include individual-level targeted measures such as isolation and contact tracing, to population-level policies such as venue closures or restrictions on social mixing \citep{haugRankingEffectivenessWorldwide2020, sharmaUnderstandingEffectivenessGovernment2021}.
 The success of such interventions in curtailing transmission before it becomes sustained depends on several factors, including: the type of intervention, timing of introduction, and the intervention's effectiveness \citep{longiniContainingPandemicInfluenza2005}.
@@
 pandemic influenza), incidence of new infections can grow quickly, outpacing a response.
-Moreover, pathogens such as influenza, which have a large proportion of onward transmission prior to symptom onset and have asymptomatic (subclinical) cases, are less easily controlled via isolation of symptomatic cases and tracing and quarantining of their contacts~\citep{fraserFactorsThatMake2004}.
+Moreover, pathogens such as influenza, which have a large proportion of onward transmission prior to symptom onset and have asymptomatic (subclinical) cases, are less easily controlled via isolation of symptomatic cases and tracing and quarantining of their contacts \citep{fraserFactorsThatMake2004}.
+\\
 
 Several pre-COVID-19 studies concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and tracing their contacts, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}.
-However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential.
+However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical A/H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential.
 \\
 
-%Isolation of symptomatic infections aims to break the chain of transmission by preventing further infection of susceptible individuals.
-The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to becoming infectious optimal.
-This can make it challenging to control outbreaks with symptom-based isolation, as for most pathogens infectiousness preceeds prodromal symptoms [AK: reference?
-This is a strong claim – best to focus on flu I think].
-Quarantining contacts exposed to an infectious individuals, independent of whether they later become a case, is a control strategy to reduce transmission before or shortly after symptom onset.
-Contact tracing is the process of finding the contacts of infected individuals.
-Both isolation and contact tracing are often used NPIs to try and curtail the growth of an epidemic.
-Forward contact tracing asks infected individuals who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed.
-Isolation, quarantine, and contact tracing is a targeted approach that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020}.
-Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}.
-However, in epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}.
-\\ AK: This is probably too much fundamental background for an applied journal – readers will know what isolation and contact tracing is.
-
-%The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic.
-Thus, Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}.
-\\ [AK: These headline fatality risks are likely to be biased: https://kucharski.substack.com/p/how-fatal-is-h5n1-influenza]
-
 In this study we evaluate the effectiveness of controlling outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model of isolation and contact tracing.
-In particular, we investigate the assumption that influenza will inevitably outpace ability to trace and isolate sufficiently effectively for control, and whether this assumption holds for all pathogen subtypes given emerging estimates of incubation period and serial interval \citep{Ward2024.12.11.24318702}.
-Finally, we quantify the speed and effectiveness of isolation for an effective response under different response delays across epidemic scenarios, and assess the impact of reducing the delay to isolation on outbreak containment alongside use of rapid testing.
+In particular, we investigate the assumption that influenza will inevitably outpace our ability to trace and isolate sufficiently effectively for control, and whether this assumption holds for all pathogen subtypes given emerging estimates of incubation period and serial interval \citep{Ward2024.12.11.24318702}.
+Finally, we quantify the speed and effectiveness of isolation for containment under different response delays across epidemic scenarios, and assess the impact of reducing the delay to isolation on outbreak control alongside the use of rapid testing.
 
@@
 \subsection*{Transmission model}
-We used a branching process transmission model to assess the potential effectiveness of isolation and contact tracing to control an outbreak of pandemic-potential influenza \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}).
-In the model, each infectious individual generates secondary cases from a negative binomial distribution with a given reproduction number $R$ and dispersion parameter ($k$).
-Onwards transmission from infectious individuals could be reduced in the following ways: 1) \textit{symptomatic individuals isolate effectively}, 2) \textit{contacts of symptomatic individuals are traced and isolate if they become symptomatic}, 3) \textit{contacts of symptomatic individuals are traced and effectively quarantined immediately}.
+
+We used a branching process transmission model to assess the potential effectiveness of isolation and contact tracing to control an outbreak of pandemic-potential influenza, implemented in the R package \texttt{\{ringbp\}} (Figure \ref{fig:ringbp-model}) \citep{hellewellRingbpSimulateEvaluate2025}.
+In the model, each infectious individual generates secondary cases from a negative binomial distribution with a given reproduction number ($R$) and dispersion parameter ($k$).
+Onwards transmission from infectious individuals could be reduced in the following ways: 1) symptomatic individuals isolate effectively, 2) contacts of symptomatic individuals are traced and isolate if they become symptomatic, 3) contacts of symptomatic individuals are traced and effectively quarantined immediately.
 Symptomatic individuals that have been traced (i.e.
-infectees) are isolated either 1) after the onset-to-isolation delay after symptom onset, or 2) when their infector is isolated.
-The isolation time of the infectee is whichever occurs first of these two events.
+infectees) are isolated at whichever of two events occurs first, either 1) after the onset-to-isolation delay after symptom onset, or 2) when their infector is isolated (if quarantined or symptomatic) or when they become symptomatic if the infector is already isolating.
+\\
 
 We include several mechanisms by which isolation and contact tracing can be imperfect at reducing onward disease transmission.
-The proportion of contacts identified by contact tracing can range between 0\% (i.e.
+The proportion of contacts ascertained by contact tracing can range between 0\% (i.e.
 no contacts are traced) to 100\% (i.e.
@@
 Subsequent symptomatic cases that have not been contact traced are only isolated after the onset-to-isolation delay post symptom onset.
-Additionally, we assume infections have an independent probability of being asymptomatic or paucisymptomatic; these infections were not identified and therefore their contacts were not traced.
+Additionally, we assume infections have an independent probability of being asymptomatic (or paucisymptomatic); these infections were not identified and therefore their contacts were not traced.
 The advantage of this approach is that it enables estimation of the overall population transmission effect of multiple interventions without requiring each step to be individually specified as a parameter.
 
-\subsection*{Pathogen epidemiological parameters} \label{epiparameters}
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/model-schematic.png}
+\caption{Model schematic of the branching process with isolation, quarantine and contact tracing implemented in the \texttt{\{ringbp\}} R package.
+(A) The transmission chain from the index case (Person A), infection events are shown by black vertical dashed lines.
+Individuals can be infected (red circle), become symptomatic (orange circle), and isolated (blue circle).
+Quarantine is optional in the model, assuming quarantine is active contacts are quarantined when the infector is isolated independent of the contacts symptom status (green circle with Q).
+The time at which a infectors isolation triggers the isolation of a symtomatic case is shown by the blue vertical dashed line (Person A to Person B).
+The isolation times that would have been triggered after a delay from symptom onset, but are after isolation was triggered by contact tracing is shown by a blue circle with a cross and connected by a dotted line (Person B).
+Person C is an asymptomatic individual (asymptomatic cases can infect others).
+Person D is missed by contact tracing (contacts are traced with a probability $\rho$, and thus missed at $1 - \rho$).
+The time Person D would have isolated if ascertained is shown by the vertical blue mark.
+Assuming quarantine is active, Person F would be isolated before symptom onset at the time of the infector's (Person D) isolation time.
+If quarantine is not active, then Person F would be isolated after a delay from symptom onset, as they had yet to show symptoms when the infector was isolated.
+(B) A branching process diagram showing who-infected-whom.}
+\label{fig:ringbp-model}
+\end{figure}
 
-In this study we evaluate the feasibility of controlling outbreaks of A/H5N1, A/H1N1, and A/H7N9-like infections.\cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}A).
+\subsection*{Pathogen epidemiological parameters}
+
+In this study we evaluate the feasibility of controlling outbreaks of A/H5N1-, A/H1N1-, and A/H7N9-like infections.
+The generation time is modelled using the incubation period and the proportion of presymptomatic transmission to simulate a correlated structure between the two delay distributions to generate sensible delays between exposure, symptom onset and onward transmission (Figure \ref{fig:delay-distributions}A) \citep{hellewellFeasibilityControllingCOVID192020, lintonCorrelationTimesSARSCoV22022}.
+We extracted previously estimated incubation periods from the literature for each subtype.
+\cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}B).
 \cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed.
-For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}A).
-Comparatively, the incubation period for H1N1 shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}A).
-We consider four reproduction number ($R$) values when simulating transmission: 1.1, 1.5, 2.5 and 3.5.
-These are based on similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution.
+For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}B).
+Comparatively, the incubation period for H1N1 shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}B).
+\\
 
-We three presymptomatic transmission scenarios: <1\%, 15\%, 30\% of transmission occurring prior to symptom onset. % AK: Probably need a 50%?
+We simulated three presymptomatic transmission scenarios: 1\%, 15\%, and 30\% of transmission occurring prior to symptom onset.
 We assumed that asymptomatic infections can occur and infect others, but are less common that symptomatic infectious indivdiuals and that the majority of cases will develop symptoms during their infectious period.
-We ran three scenarios, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.
+We ran scenarios with three different assumed proportion of asymptomatic cases, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.
+These proportions of presymptomatic transmission and asymptomatic fraction cover a broad range of possible pandemic influenza possibilities (Figure \ref{fig:patho-param-space}) \citep{lesslerOutbreak2009Pandemic2009, xuSerologicalInvestigationSubclinical2013, montgomeryRoleAsymptomaticInfections2024, gargHighlyPathogenicAvian2025}.
+\\
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth,height=0.85\textheight]{../plots/delay_distributions.png}
+\caption{(A) Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skew-normal distribution.
+Three skew-normal distribution probability density functions are plotted with different skew parameters, 0.73, 1.96, 31.82, which correspond to 30\%, 15\% and 1\% presymptomatic transmission.
+(B) Incubation period for the influenza subtypes: A/H1N1, A/H5N1 and A/H7N9.
+The vertical dashed lines are the median incubation period: A/H1N1 is 1.5 days (yellow dashed line), A/H5N1 is 3.2 days (purple dashed line), and A/H7N9 is 3.0 days (blue dashed line).
+(C) Three assumed onset-to-isolation delays used in the simuluation experiment.
+The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China \citep{liEarlyTransmissionDynamics2020}, the median onset-to-isolation is 8.1 days (blue dashed line).
+The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a}, the median onset-to-isolation time is 3.4 days (red dashed line).
+The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days (orange dashed line).}
+\label{fig:delay-distributions}
+\end{figure}
+
+We considered four reproduction number ($R$) values when simulating transmission in the community: 1.1, 1.5, 2.5 and 3.5 (Figure \ref{fig:patho-param-space}).
+These are based on similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including A/H1N1pdm \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent inflated values of current estimates of A/H5N1 and A/H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution enhances the human-to-human transmissibility (e.g.
+genetic reassortment \citep{peacockGlobalH5N1Influenza2025} or mutations in receptor binding preference \citep{linSingleMutationBovine2024}).
+Once isolated, cases are assumed to not infect anyone ($R_{iso} = 0$).
+The reproduction number of asymptomatic cases is assumed equal to the community.
+The dispersion paramter ($k$) for the negative binomial offspring distributions was fixed at 0.8 for community and asymptomatic cases following evidence that influenza subtypes have moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/patho_param_space.png}
+\caption{The pathogen characteristics, in terms of tranmissibility ($R$) and proportion of disease transmission that occurs before the onset of symptoms (including cases that are entirely asymptomatic).
+The figure includes: Influenza, SARS-CoV-1, SARS-CoV-2, Smallpox, Ebola, Marburg, and HIV shown by green shaded areas.
+The size of the shaded area reflects the differences in pathogen characteristics, both between variants and outbreaks, the blurred edges reflects uncertainty in parameter estimates.
+The parameter space of transmissibility and pre- and asymptomatic transmission explored in this paper is shown in the orange box.}
+\label{fig:patho-param-space}
+\end{figure}
+
+\clearpage
+
+\subsection*{Scenarios}
+
+The onset-to-isolation delay is parameterised under three response scenarios.
+A fast response, based on the SARS in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a} and a slower response based on COVID-19 in Wuhan, China \citep{liEarlyTransmissionDynamics2020} (Figure \ref{fig:delay-distributions}C).
+Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:delay-distributions}C).
+The third onset-to-isolation scenario is based on a situation where lateral flow rapid tests (LFTs) are provided to all susceptible individuals in the population and these people are freely able to self-test if they are symptomatic.
+We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test) (Figure \ref{fig:delay-distributions}C).
+We assume that LFT sensitivity is 100\% accurate so all symptomatic individuals will receive a positive test result.
+\\
+
+To assess the feasibility of outbreak containment under different proportions of contacts traced and isolated we varied the proportion of contacts ascertained between 0\% (i.e.
+no contacts traced) and 100\% (i.e.
+all contacts traced) at intervals of 20\%.
+When contact ascertainment is zero, isolation is the only NPI reducing transmission.
+To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial independent cases.
+This could represent importation of multiple cases in a single event, such as a flight.
+Lastly, we determine if addtionally quarantining contacts, that is isolating contacts once the infector is isoloted independent of whether the contact is symptomatic, improves the ability to contain influenza outbreaks.
 
@@
 
-We define outbreak control as no new cases between 12 and 16 weeks after the start of an outbreak\cite{hellewellFeasibilityControllingCOVID192020}.
-The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct.
-There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable within 12-16 weeks.
-We set this limit to 500 cases.
-To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial independent cases.
-This could represent importation of multiple cases in a single event, such as a flight with multiple positive cases.
-\\
-
-The onset-to-isolation delay is parameterised with three scenarios: a fast response, based on the SARS in Hong Kong (Donnelly et al.
-2003) and a slower response based on COVID-19 in Wuhan, China (Li et al.
-2020) (Figure \ref{fig:delay-distributions}B).
-Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:delay-distributions}B).
-We also include a third onset-to-isolation scenario based on the case where lateral flow rapid tests (LFTs) are provided to all individuals in the susceptible population and these people are freely able to self-test if they suspect they are symptomatic.
-We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test).
-We assume that LFTs are 100\% accurate so all symptomatic individuals will receive a positive test result.
+We define outbreak control as no new cases after 12 weeks from the start of an outbreak.
+The rationale for this criteria is that some outbreaks that do not take off and become sustained may still have cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is contained \citep{hellewellFeasibilityControllingCOVID192020}.
+Due to the exponential dynamics of outbreaks and no depletion of susceptibles in our model we implement a maximum number of cases in each simulation.
+When this cumulative case limit is reached, it is judged to be a sustained epidemic, not controllable within 12 weeks.
+We set this limit to 5,000 cases.
 
@@
 
-The ability to control an influenza epidemic is highly dependent on the reproduction number and the speed of isolation after symptom onset (Figure \ref{fig:prop-outbreak-control-R}).
-A higher outbreak reproduction number results in a smaller percentage of outbreaks controlled at a given percentage of contacts traced, except in cases where they are equal because zero or all outbreaks are contained, across all onset-to-isolation scenarios (Figure \ref{fig:prop-outbreak-control-R}).
-When the reponse to isolate cases is slow then outbreaks with $R = 1.5$ require at least 50\% of contacts traced to contain at least half of simulated outbreaks.
-When $R$ is 2.5 or 3.5 then contact tracing needs to be extremely comprehensive ($>90\%$) to have a chance at containment, and even then $R = 3.5$ influenza outbreaks on average are uncontrolled (Figure \ref{fig:prop-outbreak-control-R}A).
-Especially for H1N1, given its faster tempo of transmission, has a much lower probability of containment at 100\% of contacts traced with a slow, COVID-19-like, isolation response time (Figure \ref{fig:prop-outbreak-control-R}A).
-In a scenario with a fast, Hong Kong SARS-like, onset-to-isolation response, outbreaks with $R = 1.1$ are almost always contained at any proportion of contacts traced (Figure \ref{fig:prop-outbreak-control-R}B).
-For slow and fast isolation response delays there is a substantial difference between outbreaks with a $R$ slightly above unity ($R = 1.1$) which were controlled at a low levels ($\geq 20\%$) of isolation and contact tracing for all three influenza subtypes (Figure \ref{fig:prop-outbreak-control-R}A,B).
-As $R$ increases to 1.5 there is a dramatic decline in the ability to control the epidemic without isolation of traced contacts (i.e.
-0\% contact ascertainment), however, outbreaks with $R = 1.5$ are mostly contained once the percentage of contacts ascertained in tracing exceeds 50\% (Figure \ref{fig:prop-outbreak-control-R}B).
-For more transmissible influenza scenarios, even a fast isolation response, with an average of 3.4 days between symptom onset and isolation, requires $>70\%$ of contacts traced for most outbreaks to be controlled (Figure \ref{fig:prop-outbreak-control-R}B).
-\\
-
-The most rapid scenario we simulated to emulate individuals self-testing with LFTs, has the same trend of increased chance of outbreak containment by reducing the time between symptom onset and isolation, as the slow and fast scenarios (Figure \ref{fig:prop-outbreak-control-R}).
-The control of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C).
-Assuming ubiquitous access to home testing and that 63\% of symptomatic cases test and isolate within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$).
-The LFT isolation strategy is also the only onset-to-isolation delay which can achieve 100\% of outbreaks controlled when all contacts are traced (Figure \ref{fig:prop-outbreak-control-R}C).
-\\
-
-The difference in influenza subtype incubation period influences the efficacy of isolating and tracing.
-This effect is most pronounced when the percentage of outbreaks controlled is greater than 0\% and less than 100\% (i.e.
-not at the boundary) (Figure \ref{fig:prop-outbreak-control-R}).
-The shortest incubation period, H1N1, is predominantly the least controllable for a given reproduction number, showing that infections that are symptomatic shortly after infection and in some instances transmission while subclinical limit our ability to effectively trace and isolate their potential infectees in time to prevent disease transmission.
-The variance between the influenza subtypes at each percentage of contacts traced and isolated varies (i.e.
-difference between circle, triangle and square points at each contacts traced \% in Figure \ref{fig:prop-outbreak-control-R}).
-This variance is greatest at intermediate (20\% - 80\%) contact tracing effectiveness and intermediate values of $R$ (1.5 and 2.5) (Figure \ref{fig:prop-outbreak-control-R}).
-This could indicate a substantive role for incubation period durartion in the control of disease transmission.
-However, this variance in proportion of outbreaks controlled between pathogen subtypes does not show a strong pattern across all scenarios simulated (Figure \ref{fig:prop-outbreak-control-var-R}).
-There is a weak pattern of increased variance in controllability between subtypes at intermediate values of $R$, overall it seems that incubation period causes differences in control-by-isolation effectivness when not at the boundary of epidemic controllability.
-In other words, differences between influenza subtypes arise when not in a situation of high transmissibility and low contact tracing effectivness that cannot contain any outbreaks, nor low transmissibility and high contact tracing effectivenss easily contain all outbreaks (Figure \ref{fig:prop-outbreak-control-var-R}).
+The reproduction number, contact tracing ascertainment and speed of isolation all affect the ability to control a pandemic potential influenza outbreak (Figure \ref{fig:prop-outbreak-control-R}).
+Isolation alone can control over 50\% of epidemics with an $R$ of 1.1 with a slow onset-to-isolation response (median 8.1 days); while a fast (median 3.4 days) or LFT (median 1.4 days) response almost always controls $R = 1.1$ outbreaks (Figure \ref{fig:prop-outbreak-control-R}).
+When $R = 1.5$, $>$75\% of contacts need to be traced to control at least 75\% of outbreaks for a slow onset-to-isolation, whereas only 40\% of contacts require tracing for at least 75\% controlled under a fast isolation delay (Figure \ref{fig:prop-outbreak-control-R}A,B).
+The isolation upon rapid LFT self-testing controls epidemics with $R \leq 1.5$ without tracing contacts (Figure \ref{fig:prop-outbreak-control-R}C).
+High transmissibility scenarios ($R \geq 2.5$) require comprehensive contact tracing to contain at least 50\% of simulated outbreaks ($\sim$ 100\% for a slow response and $\sim$ 80-100\% for a fast response).
+Isolation and tracing 80\% of contacts using LFTs to trigger isolation controls over 80\% of simulated outbreaks for all subtype and transmissibility scenarios (Figure \ref{fig:prop-outbreak-control-R}C).
 \\
@@
 
-\clearpage
-
-The proportion of transmission that occurs before on the onset of symptoms has a moderate affect on epidemic containment for across isolation response rates (Figure \ref{fig:prop-outbreak-control-prop-presym-iso}).
-In all scenarios decreasing the proportion of presympatomatic transmission result in an greater or equal probability of outbreak control.
-When presymptomatic transmission is minimal, most $R = 1.1$ outbreaks go extinct without any contact tracing, at any rate of isolation after symptom onset, with extinction ensured if isolated after LFT testing (Figure \ref{fig:prop-outbreak-control-prop-presym-iso} leftmost column).
-Furthermore, when  $<1\%$ of transmission is presymptomatic, rapidly isolating after symptoms (e.g.
-after self-testing) can contain more than half of outbreaks for all transmissibility scenarios ($R \leq 3.5$) (Figure \ref{fig:prop-outbreak-control-prop-presym-iso}).
-Conversely, when 30\% of transmission is presymptomatic, less than half of highly transmissible epidemics are controlled, even with the most optimistic isolation response time and contact tracing coverage (Figure \ref{fig:prop-outbreak-control-prop-presym-iso} rightmost column).
+There is some variability in the effectiveness of isolation and contact tracing for outbreak control between influenza subtypes.
+A/H1N1, because of its shorter incubation period and thus faster tempo of transmission, has a lower proportion of outbreaks controlled, most noticably for $R \geq 1.5$ (Figure \ref{fig:prop-outbreak-control-R}).
+There is not a strong pattern for the variance in outbreak control between influenza subtypes across scenarios; the few scenarios with higher variance between subtypes are at high percentages of contacts traced ($\geq$ 60\%) and for high transmissibility scenarios (Figure \ref{fig:prop-outbreak-control-var-R}).
 \\
 
-The pattern of outbreak control is broadly similar between 10\% asymptomatic cases and the 0\% and 30\% asymptomatic scenarios (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
-Qualitative difference are in the 30\% asymptomatic scenario for LFT isolation, where the ability to control highly transmissible influenzas reduces to less than a quarter of outbreaks controlled at maximum contact tracing coverage (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
-When every case is symptomatic the likelihood of containing highly transmissible outbreaks drastically increases for all onset-to-isolation response delays (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
+Increased presymptomatic transmission and asymptomatic cases decreases epidemic containment (Figure \ref{fig:prop-outbreak-control-prop-presym-iso} and \ref{fig:prop-outbreak-control-prop-asym-iso}).
+When 30\% of transmission occurs before symptom onset, in high transmissibility scenarios ($R \geq 2.5$) more than half of outbreaks become sustained epidemics for all onset-to-isolation delays (Figure \ref{fig:prop-outbreak-control-prop-presym-iso}).
+When only 1\% of transmission is presymptomatic and contact tracing ascertains all contacts, outbreaks with $R = 3.5$ are contained in $>$75\% of outbreaks with LFT testing, and in $\sim$25\% of outbreaks with a fast isolation response (Figure \ref{fig:prop-outbreak-control-prop-presym-iso}).
+When all cases develop symptoms, isolation and contact tracing all contacts can control all outbreak scenarios, with the sole exception being A/H1N1 with $R = 3.5$ and a slow onset-to-isolation (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
+When 10\% of cases are asymptomatic the ability to contain highly transmissible influenza ($R \geq 2.5$) dramatically decreases, and decreases further when 30\% of cases are asymptomatic or paucisymptomatic, even with comprehensive contact tracing and isolation upon LFT self-testing, less than a quarter of outbreaks are contained (Figure \ref{fig:prop-outbreak-control-prop-asym-iso}).
 \\
 
-The ability to control an epidemic that can readily transmit human-to-human is moderated by the initial number of infectious individuals that seed the outbreak.
-The higher the number of cases the less chance that stochastic extinction of a transmission chain will end the whole epidemic.
-For an influenza pandemic, the number of initial infected individuals influences efficacy of control-by-isolation at all transmissibility scenarios evaluated ($R \in \{1.1, 1.5, 2.5, 3.5\}$, Figure \ref{fig:prop-outbreak-control-num-init-cases}).
-When $R = 1.1$ most outbreaks can be controlled without isolation and contact tracing, even with 40 initial infectors Figure \ref{fig:prop-outbreak-control-num-init-cases}).
-When $R = 1.5$ the increase in the number of initial infections from 5 to 20 has a large decrease in outbreaks controlled at low levels of contact tracing, the same pattern when the number of initial infectors is twofold larger at 40, but these differences in outbreak control diminish as contact tracing increases towards 100\% (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
-When the number of cases seeding the outbreak is $\geq 20$ at $R = 2.5$ a large proportion of outbreaks become uncontrollable until contact tracing reaches $\geq 80\%$.
-In the worst case transmissibility scenario ($R = 3.5$) if contact tracing is not activated until at least 20 individuals are infected then close to 100\% of contacts with need to be traced in order to control the majority of outbreaks; except for H1N1 where only 25\% of outbreaks are controlled if seeded with 40 cases (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+The more cases that initially seed an outbreak with $R \geq 1$, the less chance that it will stochastically go extinct and we find that the higher the number of initial influenza cases reduces the efficacy of control-by-isolation (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+Isolation alone can control outbreaks with $R = 1.1$ even when seeded with 40 initial cases (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
+When $R \geq 1.5$ increasing the number of initial cases makes epidemics less controllable for all influenza subtypes.
+For high transmissibility ($R \geq 2.5$) scenarios that are seeded with at least 20 cases, close to 100\% of contacts need to be traced and isolated upon symptom onset to control the majority of outbreaks, assuming a fast onset-to-isolation response (3.4 days on average) (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
 \\
@@
 
-\clearpage
-
-The median size of contained influenza outbreaks is predominantly below 100 cases for all influenza subtypes and isolation response delays (Figure \ref{fig:median-controlled-outbreak-size}).
-There are a few parameter sets that have a median cumulative outbreak size between 200 --- 400, but very few have 400 --- 500.
-This elucidates that when an outbreak is eliminated by isolation alone or in combination with contact tracing across ascertainment levels, it is done so with a relatively small number of cases (Figure \ref{fig:median-controlled-outbreak-size}).
-The total size of controlled outbreaks is also relatively constant across influenza subtypes and isolation response (Figure \ref{fig:median-controlled-outbreak-size}), noting that that the percentage of controlled outbreaks between parameter sets varies drastically with some only producing uncontrolled outbreaks (Figure \ref{fig:prop-controlled-outbreak}, \ref{fig:prop-outbreak-control-R}).
+In the acute phase of outbreaks there can be thousands of weekly cases, especially for high transmissibility scenarios ($R \geq 2.5$) (Figure \ref{fig:max-weekly-cases}).
+Specifically for controlled outbreaks, the cumulative number of cases is consistently below 500 cases across influenza subtypes (Figure \ref{fig:median-controlled-outbreak-size}).
+Thus when an outbreak is controlled with isolation alone or in combination with contact tracing it is done so without incurring a high initial prevalence.
 \\
 
-Contact tracing systems have a capacity limit which if exceeded will lower ascertainment of case contacts.
-The infinite susceptible population assumption in our model means uncontrolled outbreaks and contacts requiring tracing can become infinitely large.
-However, we find that in the early acute phase of both contained and uncontained epidemics the maximum weekly cases exceeds 600 in some scenarios (Figure \ref{fig:max-weekly-cases}).
-These cases and their contacts require testing and potentially isolating, showing the rapid speed at which outbreaks can overburden surveillance systems.
-The fact that some outbreaks have more maximum weekly case than the outbreak limit we define shows that outbreaks can become almost impossible to control on the scale of weeks from the initial cases (Figure \ref{fig:max-weekly-cases}).
-When isolation and 100\% of contacts are traced in the most optimal isolation response scenario (LFT) the maximum number of cases weekly is around 450 ($R = 3.5$).
+The control of influenza outbreaks with a $R > 1$ require interventions to bring $R$ below 1 for transmission to eventually cease.
+Isolation of symptomatic cases alone reduces $R$ (Figure \ref{fig:r0_low_presym_asym}).
+When the delay between symptom onset and isolation is around 8 days (i.e.
+slow response), the reduction in $R$ can be minimal, whereas if isolation is rapid after symptom onset, around 1.5 days (i.e.
+LFT self-testing response), isolation without contact tracing or quarantine brings $R_t$ below 1 when $R_0 < 3$ (Figure \ref{fig:r0_low_presym_asym}).
+Contact tracing can further suppress transmission, with 100\% contact ascertainment bringing an outbreak under control ($R < 1$) for all influenza subtypes and isolation responses for $R_0 < 2.5$ (Figure \ref{fig:r0_low_presym_asym}).
+Quarantine has a negligible impact on the reducing disease transmission, with a slightly greater reduction in transmission when there is a higher proportion of presymptomatic transmission (Figure \ref{fig:r0_low_presym_asym} and \ref{fig:r0_high_presym_asym}).
+The effectiveness of tracing and isolating contacts is reduced when more cases transmit before symptomatic or are asymptomatic (Figure \ref{fig:r0_low_presym_asym} and \ref{fig:r0_high_presym_asym}).
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=\textwidth]{../plots/r0_low_presym_asym.png}
+\caption{The median effective reproduction number ($R_t$) across simulation replicates calculated from simulated outbreak data under various influenza subtype and intervention scenarios against the basic reproduction number ($R_0$) for the general community used to simulate the outbreaks.
+The $R_0$ value for isolated cases is zero.
+The diagonal line shows an equal value for $R_t$ and community $R_0$.
+The dotted lines are where the values of $R_t$ (horizontal) or community $R_0$ (vertical) equal unity.
+Values of $R_t$ below the diagonal indicate a reduction in transmission from NPIs compared to uncontrolled community transmission.
+Isolation of symptomatic cases is active for all scenarios, contact tracing ascertainment is plotted for 0\% (no contact tracing), 50\% and 100\% (all symptomatic contacts of symptomatic cases traced), we run each scenario with quarantine of contacts active (dashed line) and inactive (solid line).
+This is for a baseline scenario where the proportion of presymptomatic transmission is 1\% and the proportion of asymptomatic cases is 0\%.}
+\label{fig:r0_low_presym_asym}
+\end{figure}
+
+\clearpage
 
@@
 
-\begin{itemize}
-\item This study has looked at the feasibility and effectiveness of isolation and contact tracing for influenza subtypes that have cause epidemic outbreaks.
-\item Outbreaks controlled at 50\% of contacts traced and isolated is over ??\%, highlighting the effectiveness of this intervention for infectious diseases are not highly transmissible between humans.
-\item Discuss the existing flu vaccines and antivirals that can be used an interventions in addition to NPIs
-\item In modelling the effectiveness of isolation and contact tracing we have assumed that there is no capacity limits on contact tracing, which does not hold in large outbreaks such as SARS-CoV-2 (ref).
-We also ignore potential socio-political factors that would prevent health worker from visiting or contacting infected individuals or contacts of infected individuals (Dhillon and Kelly, 2018).
-We also did not consider zoonotic spillover infections or formite transmission.
-\item Digital tracing has capacity for large-scale contact tracing (pingdemic), also rapid tests are a complementary approach for limiting spread.
-\item In this study we model isolation and contact tracing as the sole intervention in containing an influenza outbreak.
-This is to provide an assessment of its effectiveness for rapidly transmission and potentially highly transmissible subtypes.
-However, in reality, pandemic preparedness to a pandemic potential influenza would not need to deploy interventions, such as contact tracing individually or sequentially, instead they can be used in combination, such as antivirals (e.g.
-oseltamivir), rapid tests, contact tracing, and vaccination.
-\item Antivirals for half the UK, these can be deployed in rapid response to flu pandemic.
-\item In this study we have parameterised the transmission model with parameter from avian influenza subtypes (H5N1 and H7N9) and a influenza subtype that caused an epidemic in the recent past (H1N1), and response delays from SARS and COVID-19 outbreaks.
-However, these results can be used to inform other influenza subtypes that exhibit similar epidemiological characteristics in transmissibility and severity from past outbreaks (H2N2 and H3N2) and still exist in animal reservoirs for potential spillovers back into human populations (ref).
-\item This study used a branching process model to simulate individual-level transmission and intervention.
-The model assumes an infinite population size with no interdependence of isolation and contact tracing between cases whereby a case may be in contact with multiple infectors and only needs to be traced in time by one contact to prevent transmission, as could happened in a fixed random network transmission model.
-It is unlikely that this model structure difference would have a considerable different on our results for the control of disease outbreaks with contact tracing.
-We suppose that because our model requires isolation sampled for a single individual, our results are more conservative on the ability to control outbreaks, and that using a random network model would indicate an increased effectivness of contact tracing (Juul and Strogatz, 2023).
-The infinite population assumption also means that although isolated individuals are assumed to not transmit, there are always other individuals that non-isolated infectors can transmit to, whereas a fixed random network model that assumes optimal indefinite isolation will block of parts of the network and overestimate contact tracing effectiveness (Juul and Strogatz, 2023).
-\item The reproduction number for H5N1 (including clade 2.3.4.4b) and H7N9 influenza subtypes have been estimated to be far below 1 (Ward et al., Garg et al., 2025).
-Therefore, the transmissibility scenarios explored in this study are speculative and assume that pathogen evolution increases human-to-human transmission potential.
-If transmissibility of these avian influenza subtypes remains constant over time then outbreaks are expected to be subcritical and will not produce epidemics and thus not require interventions.
-If cases of avian influenza are detected with routine surveillance without a known animal source, it may indicate sustained human-to-human transmission and can be a trigger to commence isolation and contact tracing while the number of initial cases is presumed small, which as we show increases epidemic controllability (Figure \ref{fig:prop-outbreak-control-num-init-cases}).
-\item This study has evaluated the ability to control the outbreak of an infectious disease independent on its severity.
-Infection severity and fatality risks are of course critical when formulating a pandemic response plan.
-By understanding the efficacy of isolation and contact trace for a given epidemiological characterstic, outbreak mortality and morbidity can hopefully be minimised by reducing disease incidence over time to elimination.
-\item Our branching process model used a simple forward contact-tracing mechanism that reduces the number of secondary contacts based on the timing of isolation of symptomatic cases.
-This contact tracing stragety is minimal cost and effort.
-In scenarios where outbreaks are uncontrolled a more comprehensive and costly contact tracing system where infectors and contacts of infectors (i.e.
-backward and forward contact tracing) are followed up, which has been shown to be more effective at outbreak containment (Klinkenberg et al., 2006).
-\item Isolation and contact tracing complementing in combination with other quickly available measures for pandemic potential influenza strains, such as rapid testing and antivirals \citep{haydenPerspectivesAntiviralUse2001}.
-\item This paper has focused on forward contact tracing.
-Backward contact tracing, whereby infected individuals are asked who they have been in contact with prior to being infected to identify the source of infection and potentially forward trace from the source, is applicable to infectious with long infectious periods (e.g.
-HIV), but is not effective for more ephemeral infections with faster transmission, as is the case for influenza.
-\item Contact tracing and isolation is a targeted approach analogous to ring vaccination.
-It looks to prevent onward transmission by, either isolating or quarantining contacts in the case of contact tracing, or vaccinating contacts in the case of ring vaccination \citep{kucharskiEffectivenessRingVaccination2016, whittakerQuantifyingImpactBroadly2024}.
-Ring vaccination for an influenza outbreak would not be effective due to the generation time, proportion of presymptomatic and asymptomatic, and likelihood of geographically widespread (i.e.
-non-localised) transmission.
-\end{itemize}
+Influenza remains a high risk pathogen with pandemic potential as shown by the history of influenza pandemics throughout the 20th century \citep{saunders-hastingsReviewingHistoryPandemic2016}.
+The recent spread of A/H5N1 avian influenza in cattle and poultry farms, primarily in the US, creates a greater spillover risk and adaptations to enhance mammal-to-mammal transmission \citep{linSingleMutationBovine2024, gargHighlyPathogenicAvian2025, peacockGlobalH5N1Influenza2025}.
+In this study we use a mathematical model to assess the feasibility and effectiveness of isolation and contact tracing for avian and non-avian influenza subtypes to determine whether these should be considered effective NPIs in the event that sustained human-to-human transmission is detected.
+\\
+
+We find that the percentage of contacts that need to be successfully traced to control an outbreak depends on the reproduction number of the pathogen, and the speed at which a public health system can respond by testing and communicating with those infected and contacts of infectious persons.
+If the delay between symptom onset and isolation is at least one week, then isolation and comprehensive contact tracing are required to be confident a moderate to highly transmissible ($R \geq 1.5$) influenza can be controlled.
+By reducing the delay between symptom onset and isolation to around three days, influenza pathogens that are marginally supercritical ($R = 1.1$) become subcritical ($R < 1$) and contained by isolation alone, and more transmissible scenarios can be controlled with good contact tracing coverage.
+In addition to two empirically-based onset-to-isolation delays, by testing a rapid isolation scenario where LFTs that do not require a test being sent to a laboratory, as would be the case for polymerase chain reaction (PCR) testing, the benefit to containment probability, and the opportunity to control highly transmissible ($R \geq 2.5$) respiratory diseases with isolation and contact tracing.
+Therefore, the dogma that symptom-based isolation and contact tracing are not effective for influenza is shown not to hold for optimal testing and isolation scenarios, even if some cases are missed due to being asymptomatic and some transmission occurs prior to symptom onset.
+If making LFTs widely accessible is economical relative to the cost of mass contact tracing, a low to moderate transmissibility influenza pandemic response should focus on reducing the onset-to-isolation response.
+Whether the interventions proposed here are cost-effective remains a question for further investigation.
+\\
+
+The transmission model highlights the bimodal nature of the final size of outbreaks, where NPI-controlled epidemics typically range from tens to hundreds of total cases, in contrast, uncontrolled outbreaks can produce thousands of new cases each week.
+The severity of avian influenza subtypes is uncertain, historical estimates are $\sim$30\% --- 55\% \citep{tannerPandemicPotentialAvian2015}, but recent human infections have lower morbidity and mortality \citep{gargHighlyPathogenicAvian2025, krammerHighlyPathogenicAvian2025}.
+Consequently, the difference in disease burden between containing an outbreak while still small compared to allowing thousands of cases will be substantial.
+Even if updated severity estimates are significantly lower, widespread infections could still result in a devastating outcome.
+\\
+
+It is possible that a sentinel surveillance system or community testing to detect new cases of pandemic potential influenza (e.g.
+A/H5N1 or A/H7N9), could miss the primary case(s), whether originating from zoonotic spillover or importation.
+We show that more cases before an NPI is activated hinders the ability to effectively contain the outbreak.
+The magnitude of this impact is moderated more by transmissibility than influenza subtype for the three variants we considered.
+If interventions are activated when case prevalence exceeds 50 cases, then even rapidly isolating over 60\% of people with 2 days of becoming symptomatic will require near-perfect contact tracing coverage.
+For comparison, on the date of the first fatality in the UK from COVID-19 there were already over 100 confirmed cases and probable sustained widespread community transmission.
+A subcritical influenza outbreak could cause transient epidemics, during which, adaptation to enhance human-to-human tranmsmission can cause a critical transission to supercritical transition, at which point more importations exponentially decreases containment without interventions \citep{whittleOutcomeStochasticEpidemicA1955, farringtonDistributionTimeExtinction1999, antiaRoleEvolutionEmergence2003, lloyd-smithSuperspreadingEffectIndividual2005}.
+This study does not model this transition, instead we assume supercritical human-to-human transmission potential for A/H5N1 and A/H7N9 to evaluate NPIs because current estimates for pathogen epidemiology ($R < 1$) do not require interventions for containment.
+Cases detected with routine surveillance without a known animal source may indicate sustained human-to-human transmission and could be a trigger to commence isolation and contact tracing while the number of initial cases is presumed small.
+\\
+
+The strength of our approach is based on the simple parameterisation of the individual-based model to capture the infectiousness, disease progression and onward transmission incorporating individual-level heterogeneity.
+Sampling the generation time using the incubation period and proportion of presymptomatic infections (using a skew-normal distribution) enables simulating transmission dynamics for novel pathogens and early in outbreaks when the incubation period is estimated but the generation time is unknown or estimated with more uncertainty, and the serial interval would introduce bias if used instead \citep{brittonEstimationEmergingEpidemics2019a, lehtinenRelationshipSerialInterval2021}.
+\\
+
+The model and simulation experiment has several limitations.
+We assumed that the LFT rapid tests are accurate and do not provide false negatives, in reality these type II errors will result in a smaller reduction in transmission \citep{petoCOVID19RapidAntigen2021}.
+Secondly, the isolation of infectious individuals is assumed to prevent all onward transmission.
+This may be overly conservative if for example individuals isolate at home while living with others, who may become infected and continue to spread the disease until they become symptomatic and isolate.
+We find that quarantine has a negligible reduction -- and counter intuitively in some cases a small increase -- in reproduction number.
+This is likely due to the model implementation of quarantine, whereby ..., and should not be concluded that quarantine is ineffective for respiratory pathogens.
+Lastly, the transmission model does not take into account network or age structure which has been shown to influence the effectiveness of isolation and contact tracing (Juul and Strogatz, 2023).
+\\
+
+We assessed the benefit of contact tracing across a range of possible contact ascertainment levels.
+In doing so we model up to 5,000 cases before terminating a simulated outbreak as uncontained, but before reaching the case limit we assume that there is no capacity limit on contact tracing.
+This is an optimistic assumption; in reality surveillance systems can be patchy and have finite capacity, exacerbated in resource-limited settings \citep{dhillonWhenContactTracing2018}.
+The scenarios where a low percentage of contacts were traced may be more representative for these situtations.
+The finding that weekly case incidence can reach thousands of new cases within the acute phase of an influenza outbreak highlights how the exponential dynamics of outbreaks, even in the presence of NPIs, can overburden surveillance systems with the volume of testing and isolation required.
+Aside from public health surveillance capacity, we also ignore potential socio-political factors that may prevent contacting infected individuals or contacts of infected individuals \citep{dhillonWhenContactTracing2018}.
+In this study we assume that contacts are known, avoiding the need to define what constitutes a contact.
+In reality the threshold for duration and distance of a contact for a directly transmitted respiratory disease will influence the ascertainment and volume of contact tracing required by balancing the error rates of contacts that result in infections \citep{keelingEfficacyContactTracing2020}.
+\\
+
+The experience of contact tracing from the COVID-19 pandemic has shown that isolating and quarantining contacts can be expedited with the use of digital tracing.
+As opposed to symptom-based testing and isolating, a contact can be determine by proximity and duration between mobile devices, and if someone tests positive it can immediately notify all recent contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020}.
+This approach would be complimentary to the LFT scenario in this study where available rapid tests could be used when notified of an infectious contact (assuming the time between exposure and testing is longer than the detectable latent period).
+Decentralising contact tracing to devices most people carry around at all times (e.g.
+smartphones) would also ease capacity limitations of a centralised contact tracing unit.
+The optimistic scenarios of traditional testing, tracing and isolating included could in theory be improved and epidemic prevention enhanced by integrating digital technology.
+\\
+
+Enacting isolation and contact tracing and other NPIs is expected with the emergence of a novel pathogen due to widespread susceptibility and no pharmaceutical interventions.
+However, influenza has multiple circulating seasonal subtypes and past pandemics, resulting in a complex mosaic of population immunity for different influenza subtypes, with possible immunological imprinting \citep{gosticPotentProtectionH5N12016}.
+There is also seasonal vaccination availability in many high- and middle income countries \citep{goldinSeasonalInfluenzaVaccination2024}, stockpiles of vaccines for avian influenza, as well as licensed effective antivirals \citep{krammerHighlyPathogenicAvian2025}.
+We show that asymptomatic influenza infectious will detrimentally influence containment with symptom-based isolation and contact tracing.
+If an epidemic is highly asymptomatic ($>25$\% cases), countries with a stockpile of influenza antivirals, could deploy those prophylactically in response \citep{haydenPerspectivesAntiviralUse2001}.
+For example, the UK which stockpiles antivirals for $\sim$ 50\% of the population \citep{WrittenQuestionsAnswers}.
+Pandemic response could deploy pharmaceutical and non-pharmaceutical interventions in combination to prevent disease burden by suppressing transmission and severity.
+The stockpile of vaccines for A/H5N1 that shows cross-reactivity with current 2.3.4.4b strain are another line of defence \citep{khuranaLicensedH5N1Vaccines2024}.
+The speed of isolating contacts compared to vaccinating them in the context of a fast generation time and proportion of presymptomatic transmission, renders a ring vaccination less effective \citep{kucharskiEffectivenessRingVaccination2016, whittakerQuantifyingImpactBroadly2024}.
+Overall, the response to an influenza pandemic from an avian or swine subtype would be more multifacetted than the baseline containment results from our model.
+\\
+
+Our study shows that isolation and contact tracing are effective NPIs that should be considered when evaluating a robust outbreak reponse to pandemic potenial influenza.
+If responding to a pathogen where human-to-human transmission is sustained ($R \leq 1.5$) isolation of symptomatic cases alone will bring the reproduction number below one, and comprehensive tracing and isolation of contacts with a short delay to isolation can control outbreaks with $R \leq 3.5$.
+Whether the next public health emergency is from avian (A/H5N1 and A/H7N9) or swine (A/H1N1) influenza, the response should be broadly similar.
+If in the acute phase of the outbreak clinical and epidemiological investigation indicate presymptomatic transmission or asymptomatic infections \citep{liuContributionPresymptomaticInfection2020}, or that multiple imported cases have seeded human-to-human transmission chains \citep{russellEffectInternationallyImported2021}, then increasing contact tracing ascertainment and reducing the onset-to-isolation delay will improve chances of containment.
+Our results broadly cover to novel emerging pathogens that lack pharmaceutical interventions; extending results to other pandemic pathogen candidates depends on the similarity of their epidemiological characteristics.
 
@@
 \centering
-\includegraphics[width=\textwidth]{../plots/prop_controlled_outbreak.png}
-\caption{Median...}
-\label{fig:prop-controlled-outbreak}
+\includegraphics[width=\textwidth]{../plots/max_weekly_cases.png}
+\caption{The median maximum number of weekly cases across simulation replicates for different values of community reproduction number ($R_0$).
+This is facetted by influenza subtype (A/H5N1, A/H1N1, A/H7N9) and onset-to-isolation response delays (slow COVID-like response, fast SARS-like response and LFT self-testing) The boxplots contain all scenarios (e.g., proportion of asymptomatic cases or number of initial cases) within that reproduction number scenario, the whiskers correspond to the 25th and 75th quartiles Data outside of 1.5 $\times$ the interquartile range are plotted as outlier points.}
+\label{fig:max-weekly-cases}
 \end{figure}
@@
 \centering
-\includegraphics[width=\textwidth]{../plots/max_weekly_cases.png}
-\caption{Max...}
-\label{fig:max-weekly-cases}
+\includegraphics[width=\textwidth]{../plots/r0_high_presym_asym.png}
+\caption{The median effective reproduction number ($R_t$) across simulation replicates calculated from simulated outbreak data under various influenza subtype and intervention scenarios against the basic reproduction number ($R_0$) for the general community used to simulate the outbreaks.
+The $R_0$ value for isolated cases is zero.
+The diagonal line shows an equal value for $R_t$ and community $R_0$.
+The dotted lines are where the values of $R_t$ (horizontal) or community $R_0$ (vertical) equal unity.
+Values of $R_t$ below the diagonal indicate a reduction in transmission from NPIs compared to uncontrolled community transmission.
+Isolation of symptomatic cases is active for all scenarios, contact tracing ascertainment is plotted for 0\% (no contact tracing), 50\% and 100\% (all symptomatic contacts of symptomatic cases traced), we run each scenario with quarantine of contacts active (dashed line) and inactive (solid line).
+This is for a baseline scenario where the proportion of presymptomatic transmission is 1\% and the proportion of asymptomatic cases is 0\%.}
+\label{fig:r0_high_presym_asym}
 \end{figure}
 
-\section*{Supplementary Materials}
-
-\subsection*{Branching process model with isolation}
-
-To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transsmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}).
-The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (infinite population size, without depletion of susceptibility).
-Influenza subtypes have shown moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}, but evidence is limited and the negative binomial allows flexibility between highly heterogeneous and homogeneous individual-level transmission.
-The mean of the negative binomial offspring distribution is the effective reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}.
-\\
-
-There are three categories for individuals in the simulation, with each having their own parameterisation for the negative binomial offspring distribution: 1) \textit{community} (i.e.
-general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}.
-Each group have a distinct $R$ and dispersion ($k$) parameter due to the assumed differences in transmission dynamics between groups.
-\\
-
-% AK EDIT: Align with definitions in main methods
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/model-schematic.png}
-\caption{Model schematic of the branching process with isolation, quarantine and contact tracing implemented in the \texttt{\{ringbp\}} R package.
-(A) The transmission chain from the index case (Person A), infection events are shown by black vertical dashed lines.
-The time individuals are infected is shown by the red dot, the time of symptom onset is shown by the orange dot, and the time time individuals are isolated is shown by the blue dot.
-Quarantine is optional in the model, assuming quarantine is active it is shown by the small green dot with a Q.
-The time at whcih a infectors isolation triggers the isolation of a symtomatic case is shown by the blue vertical dashed line (Person A to Person B).
-The isolation times that would have been triggered after a delay from symptom onset, but are after isolation was triggered by contact tracing shown by a blue dot with a cross and connected by a dotted line (Person B).
-Person C is an asymptomatic individual (asymptomatic cases can infect others).
-Person D is missed by contact tracing (contacts are traced with a probability $\rho$, and thus missed at $1 - \rho$).
-The time Person D would have isolated if ascertained is shown by the vertical blue mark.
-Assuming quarantine is active, Person F would be isolated before symptom onset at the time of the infector's (Person D) isolation time.
-If quarantine is not active, then Person F would be isolated after a delay from symptom onset, as they had yet to show symptoms when the infector was isolated.
-(B) A branching process diagram showing who-infected-whom.}
-\label{fig:ringbp-model}
-\end{figure}
-
-The time between an infector being infected and then infecting an infectee (i.e.
-generation time) is computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
-This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, onset of symptoms, and onward transmission \citep{hellewellFeasibilityControllingCOVID192020}.
-This allows the proportion presymptomatic transmission to be specified, rather than defining incubation period and generation time distributions, which only indirectly allows assumptions of presymptomatic infectiousness to be derived.
-This simple formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is estimated but the generation time is estimated with more uncertainty or unknown, and the serial interval cannot be substituted in without introducing bias \citep{brittonEstimationEmergingEpidemics2019a, lehtinenRelationshipSerialInterval2021}.
-\\
-
-Given the incubation period for each case, the proportion of the infections that occur prior to, or after, the symptom onset of an infector is controlled by the parameterisation of the skew normal distribution (Figure \ref{fig:delay-distributions}C).
-The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed.
-The scale parameter ($\omega$) is fixed at 2, this parameter (with the shape parameter) defines the variance of the distribution.
-We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2).
-The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution).
-In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector.
-\\
-
-The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing (Figure \ref{fig:ringbp-model}).
-When a case is symptomatic they are isolated after a delay.
-The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation).
-Any infection times that are sampled to occur after the isolated of an infector are removed, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times.
-In other words, once a case is isolated at time $t_{isolated}$, the probability of that case infecting anyone at $t > t_{isolated}$ is zero.
-Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$).
-Symptomatic individuals that are traced are isolated either when 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated.
-The isolation of the infectee is whatever occurs first of these two events.
-\\
-
-There are several mechanisms where isolation and contact tracing are imperfect at reducing onward disease transmission.
-The proportion of contacts missed by contact tracing is defined by the proportion of ascertainment and can range between zero, that is nobody is traced, to one, every case is traced.
-Symptomatic cases that have been missed when contact tracing their infectors are isolated at the onset-to-isolation delay after their symptom onset time.
-Additionally, contact tracing misses infections by an asymptomatic, or paucisymptomatic (subclinical), infector.
-In this event the infectees are missed by contact tracing owing to the infector not being known to the response system.
-\\
-
-\subsection*{Defining outbreak control}
-
-We use the same definition for outbreak control as \cite{hellewellFeasibilityControllingCOVID192020}: no new cases between 12 and 16 weeks after the start of an outbreak.
-The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct.
-\\
-
-There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable within 12-16 weeks.
-We set this limit to 500 cases.
-This number needs to be larger than the expected cluster size from an subcritical outbreak produce by chance (i.e.
-larger than $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number).
-
-\subsection*{Pathogen epidemiological parameters} \label{epiparameters}
-
-In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previously caused sporadic human-to-human outbreaks with pandemic potential.
-We include A/H5N1, A/H1N1, and A/H7N9.
-These have caused outbreaks of various size and severity over the past 100 years.
-H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}.
-H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}.
-The first human case with the H7N9 subtype was in 2013, subsequently causing five outbreak waves, but no human cases since 2019 \citep{liH7N9InfluenzaVirus2021}.
-Most cases are of zoonotic origin with recorded animal exposure but there have been some small clusters (2 --- 3 cases) of possible human-to-human transmission \citep{liEpidemiologyHumanInfections2014}.
-For the purposes of this study we assume that there is no preexisting population immunity for these influenza subtypes.
-This is generally true for H5N1 and H7N9 at present, and for H1N1pdm09 prior to 2009.
-\\
-
-We extracted incubation periods from the epidemiological literature that have been previously estimated for the chosen Influenza subtypes.
-All incubation periods were found to be well described by a Weibull distribution, \cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}A).
-\cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed.
-For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}A).
-Comparatively, the incubation period for H1N1 is noticably shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}A).
-\\
-
-The reproduction number ($R$) use to simulate transmission for each influenza subtype was 1.1, 1.5, 2.5 and 3.5.
-We chose these values and they represent either similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent an inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution (e.g.
-genetic reassortment (Peacock et al., 2025) or mutations in receptor binding preference (Lin et al., 2024)) enhances the human-to-human transmissibility of these pathogens.
-Current estimates for H5N1 and H7N9 subtypes are below unity (i.e.
-$R < 1$) (\citealt{tannerPandemicPotentialAvian2015}; \citealt{Ward2024.12.11.24318702}, but see \citealt{yangDetectingHumanhumanTransmission2007a} for a H5N1 $R$ estimate above 1) so transmission is expected to stop without intervention; with most outbreaks being small except in rare instances of superspreading events.
-Therefore evaluating isolation and contact tracing requires simulating outbreaks that have pandemic potential with exponential spread in the absence of control measures.
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth]{../plots/patho_param_space.png}
-\caption{The pathogen characteristics, in terms of tranmissibility ($R$) and proportion of disease transmission that occurs before the onset of symptoms (including cases that are entirely asymptomatic).
-The figure includes: Influenza, SARS-CoV-1, SARS-CoV-2, Smallpox, Ebola, Marburg, and HIV shown by green shaded areas.
-The size of the shaded area reflects the differences in pathogen characteristics, both between variants and outbreaks, the blurred edges reflects uncertainty in parameter estimates.
-The parameter space of transmissibility and pre- and asymptomatic transmission explored in this paper is shown in the orange box.}
-\label{fig:patho-param-space}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=\textwidth,height=0.85\textheight]{../plots/delay_distributions.png}
-\caption{(A) Incubation period for the Influenza subtypes: H1N1, H5N1 and H7N9.
-(B) Three assumed onset-to-isolation delays used in the simuluation experiment.
-The \textit{Slow} scenario is based on the onset-to-isolation for the SARS-CoV-2 response in Wuhan, China \citep{liEarlyTransmissionDynamics2020}, the median onset-to-isolatiojn is 8.1 days (blue dashed line).
-The \textit{Fast} scenario is based on the onset-to-isolation for SARS-CoV-1 response in Hong Kong \citep{donnellyEpidemiologicalDeterminantsSpread2003a}, the median onset-to-isolation time is 3.4 days (red dashed line).
-The lateral flow test (LFT) scenario is an assumed distributed for rapid self-administered home testing upon isolation, the median delay from symptom onset to isolation in this scenario is 1.4 days (orange dashed line).
-(C) Proportion of presymptomatic transmission determined by the incubation period, here assumed to be 5 days (dashed vertical line), controlled by a skew-normal distribution.}
-\label{fig:delay-distributions}
-\end{figure}
-
-\clearpage
-
-\subsection*{Simulation scenarios}
-
-To assess under what scenarios pandemic-potential influenza outbreaks could be contained with the use of isolation and contact tracing we set up a simulation experiment.
-We defined a parameter space to explore across a variety of epidemiological characteristics (as defined in \nameref{epiparameters} section), contact tracing effectiveness, proportion of aysmptomatic cases and presymptomatic transmission, and number of initial cases.
-\\
-
-The ability to contain an outbreak by isolating infected individuals is highly contingent on the response time, or delay, between individuals becoming symptomatic, getting tested, the test results being returned and going into isolation.
-The probability that this series of steps will take an given amount of time is encapsulated in the onset-to-isolation delay distribution, i.e.
-onset-to-isolation distribution determines the time between a case developing symptoms and then becoming isolated.
-Here we define the onset-to-isolation as a response time as it is determined by the time delay between an individual responding to symptoms by getting tested and the turn around time for the tests to be analysed and the result returned.
-We view the onset-to-isolation as a feature of the public health outbreak response rather than an epidemiological characteristic of a disease.
-As such it is possible for pandemic response to adapt their response time to meet the needs of an outbreak.
-\\
-
-We assume that an isolated case cannot transmit, so isolation is assumed to be 100\% effective once enacted ($R_{isolated} = 0$).
-The onset-to-isolation delay is parameterised with three scenarios: a fast response, based on the SARS in Hong Kong (Donnelly et al.
-2003) and a slower response based on COVID-19 in Wuhan, China (Li et al.
-2020) (Figure \ref{fig:delay-distributions}B).
-Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:delay-distributions}B).
-These two empirically informed delay distributions provide a guide to possible response times in the case of an influenza outbreak and will enable us to evaluate if such delays are too long to contain a disease that can transmit rapidly such as influenza.
-We also include a third onset-to-isolation scenario based on the case where lateral flow rapid tests (LFTs) are provided to all individuals in the susceptible population and these people are freely able to self-test if they suspect they are symptomatic.
-We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test).
-We assume that LFTs are 100\% accurate so all symptomatic individuals will receive a positive test result.
-\\
-
-We define the proportion of presymptomatic transmission ($\theta$) follwowing the approach of \cite{hellewellFeasibilityControllingCOVID192020} by parameterising a skew-normal distribution to sample the infection times using the incubation period and a shape parameter ($\alpha$).
-The shape parameter controls the proportion of presymptomatic transmission by modulating the proportion of tranmission times that are less than the incubation time (Figure \ref{fig:delay-distributions}C).
-We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: <1\%, 15\%, 30\%, these correspond to $\alpha$ values of 30, 1.95 and 0.7, respectively (Figure \ref{fig:delay-distributions}C).
-Therefore, in all scenarios most transmission occurs after symptom onset, but due to isolation only impacting symptomatic individuals the greater proportion of transmission that occurs before symptom onset, the less effective isolation and contact tracing is hypothesied to be.
-\\
-
-The proportion of pandemic-potential influenza infections that are asymptomatic varies for past outbreaks.
-There is evidence from past human outbreaks of H1N1pdm and H5N1 (2.3.4.4b) that some laboratory (polymerase chain reaction) positive cases showed no symptoms or symptoms too mild to be considered influenza-like illness \citep{lesslerOutbreak2009Pandemic2009, gargHighlyPathogenicAvian2025}; but whether human-to-human transmission can occur while subclinical is unknown.
-There is no evidence of asymptomatic transmission for the H7N9 subtype \citep{xuSerologicalInvestigationSubclinical2013}.
-Here we assume that asymptomatic cases occur and can infect others, but are uncommon and that the majority of cases will develop symptoms during their infectious period.
-We ran three scenarios, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.
-\\
-
-The feasibility of epidemic containment by contact tracing is expected to depend on the proportion of close contacts of cases that are traced and isolated if symptomatic.
-We varied the proportion of contacts ascertained in the model between 0 and 1, at intervals of 0.2.
-In cases when the proportion of contacts traced is zero, isolation after the delay from symptom onset is the only NPI reducing disease transmission.
-To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial indepedent cases.
-This could represent importation of multiple cases in a single event, such as a flight with multiple positive cases.
-\\
-
-We fixed some parameters across all simulations.
-The dispersion parameter for the negative binomial offspring distribution was fixed at 1 for isolated cases, and 0.16 for subclinical and community cases.
-For computational efficiency we capped the maximum number of days the simulation could run for to 365.
-This is substantially longer than the 12-16 week window for evaluating outbreak control so does not influence the results.
-Additionally, we limit simulations to 500 cases.
-This is an arbitrary threshold we judge to be large epidemic that is not controllable.
-
 \end{document}

### 2025-07-02 inst/paper/main.tex
add supplementary methods

@@
 
+\subsection*{Supplementary Methods}
+
+\subsubsection*{Pandemic potential influenza subtypes}
+
+In this study we evaluate the feasibility of controlling outbreaks of selected influenza subtypes that have previously caused sporadic human-to-human outbreaks with pandemic potential.
+We include A/H5N1, A/H1N1, and A/H7N9.
+These have caused outbreaks of various size and severity over the past 100 years.
+H5N1 has caused small outbreaks ($<$ 10 cases) of human-to-human transmission \citep{yangDetectingHumanhumanTransmission2007a, aditamaAvianInfluenzaH5N12012a}.
+H1N1 is now an endemic seasonal influenza with substantial population immunity, but a novel strain, H1N1pdm09, emerged in 2009 and caused a pandemic outbreak \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}.
+The first human case with the H7N9 subtype was in 2013, subsequently causing five outbreak waves, but no human cases since 2019 \citep{liH7N9InfluenzaVirus2021}.
+Most cases are of zoonotic origin with recorded animal exposure but there have been some small clusters (2 --- 3 cases) of possible human-to-human transmission \citep{liEpidemiologyHumanInfections2014}.
+For the purposes of this study we assume that there is no preexisting population immunity for these influenza subtypes.
+This is generally true for H5N1 and H7N9 at present, and for H1N1pdm09 prior to 2009.
+
+\subsubsection*{Transmission model generation time from incubation period}
+
+Here we provide a more detailed description of how the generation time is calculated from the incubation period.
+The time between an infector being infected and then infecting an infectee (i.e.
+generation time) is computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}.
+This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, onset of symptoms, and onward transmission \citep{hellewellFeasibilityControllingCOVID192020, lintonCorrelationTimesSARSCoV22022}.
+This allows the proportion presymptomatic transmission to be specified, rather than defining incubation period and generation time distributions, which only indirectly allows assumptions of presymptomatic infectiousness to be derived.
+This simple formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is estimated but the generation time is estimated with more uncertainty or unknown, and the serial interval cannot be substituted in without introducing bias \citep{brittonEstimationEmergingEpidemics2019a, lehtinenRelationshipSerialInterval2021}.
+\\
+
+Given the incubation period for each case, the proportion of the infections that occur prior to, or after, the symptom onset of an infector is controlled by the parameterisation of the skew normal distribution (Figure \ref{fig:delay-distributions}C).
+The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed.
+The scale parameter ($\omega$) is fixed at 2, this parameter (with the shape parameter) defines the variance of the distribution.
+We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2).
+The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution).
+In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector.
+\\
+
+We parameterise the skew-normal with three proportion of presymptomatic transmission scenarios: 1\%, 15\%, 30\%, these correspond to $\alpha$ values of 31.82053, 1.962614, and 0.726558, respectively (Figure \ref{fig:delay-distributions}C).
+
+\subsubsection*{Modelling non-pharmaceutical interventions}
+
+The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (infinite population size, without depletion of susceptibility).
+Influenza subtypes have shown moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}, but evidence is limited and the negative binomial allows flexibility between highly heterogeneous and homogeneous individual-level transmission.
+The mean of the negative binomial offspring distribution is the basic reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}.
+\\
+
+There are three categories for individuals in the simulation, with each having their own parameterisation for the negative binomial offspring distribution: 1) \textit{community} (i.e.
+general population), 2) \textit{isolated}, and 3) \textit{asymptomatic}.
+Each group have a distinct $R$ and dispersion ($k$) parameter due to the assumed differences in transmission dynamics between groups.
+\\
+
+The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing (Figure \ref{fig:ringbp-model}).
+When a case is symptomatic they are isolated after a delay.
+The timing of isolation is drawn from the onset-to-isolation distribution.
+Any infection times that are sampled to occur after the isolated of an infector are remove, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times.
+In other words, once a case is isolated at time $t_{isolated}$, the probability of that case infecting anyone at $t > t_{iso}$ is zero.
+Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{iso}$).
+Symptomatic individuals that are traced are isolated either when 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated.
+The isolation of the infectee is whatever occurs first of these two events.
+
+\subsubsection*{Defining outbreak control}
+
+Outbreak control, or the extinction or elimination of the pathogen, is a binary classification, which is true if there are no new cases from the 12th week of the outbreak onwards, and false otherwise.
+This is the same definition as used in \cite{hellewellFeasibilityControllingCOVID192020}.
+The 12 week threshold is somewhat arbitrary but is judged to be long enough that the NPIs can bring the effective reproduction number below one for all transmission to cease.
+The number of generations in an outbreak lasting 12 weeks will be approximately between 18 generations (H5N1 median incubation period and 0.1\% presymptomatic transmission) and 84 (H1N1 median incubation period and 30\% presymptomatic transmission), although realisations of the stochastic simulation may be greater or less than these estimates from distribution medians.
+\\
+
+The maximum number of cases in each simulation replicate before terminating the simulation and concluding it is uncontrolled was set to 5,000.
+This number should be larger than the expected cluster size from a subcritical outbreak which is given by $N_i / (1 - R)$, where $N_i$ is the number of initial cases and $R$ is the reproduction number.
+For example, using the largest initial outbreak size in this study $N_i = 40$, and assuming $R$ is subcritical but very close to unity, $R = 0.99$, gives a possible outbreak size of 4,000 cases.
+Therefore, our limit that 5,000 cases is an uncontrolled outbreak should not include any outbreaks that will go to extinction eventually but produce transient outbreaks by chance.
+\\
+
+For computational efficiency we capped the maximum number of days the simulation could run for to 140 (20 weeks).
+This is substantially longer than the 12 cutoff evaluating outbreak control so does not influence the results.
+
+\clearpage
+
+\subsection*{Supplementary Results}
+
 \begin{figure}[ht]
